---
title: "Exploring Within-Colony Thermal Tolerance Variability in Reef Building Corals"
author: "Marlen Schlotheuber"
format: html
editor: visual
---

# R code for Manuscript

### Install needed packages for data analysis

```{r}
# install packages
install.packages("ggmap")
install.packages("osmdata")
install.packages("ggspatial")
install.packages("seqinr")

if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("dada2")

BiocManager::install("dada2", force = TRUE)
install.packages(c("cubature", "ape", "reshape2", "edgeR", "plyr"))

install.packages("dada2")
install.packages("Rcpp")
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

# Install Biostrings
BiocManager::install("Biostrings")
# Install phyloseq
BiocManager::install("phyloseq")

if(!require(devtools)){
   install.packages("devtools")
}

devtools::install_github("reefgenomics/CBASSED50@0.1.5.1", force=TRUE)

# load packages
library(seqinr)
library(sf)
library(rnaturalearth)
library(ggmap)
library(dplyr)
library(tidyr)
library(readxl)
library(rstudioapi)
library(RColorBrewer)
library(CBASSED50)
library(tidyverse)
library(car)
library(mgcv)
library(phyloseq)
library(Biostrings)
library(reshape2)
library(ggplot2)
library(scales)
library(gridExtra)
library(dada2)
library(Rcpp)
library(openxlsx)
library(patchwork)
library(stringr)

# set seed for reproducability
set.seed(123)
```

## **Fig.1 Study regions and the sampling locations within massive Porites colonies**

```{r}
# Study 
# Get country boundaries
world <- ne_countries(scale = "medium", returnclass = "sf")

# Subset for Indonesia and Malaysia
countries <- world[world$name %in% c("Indonesia", "Malaysia"), ]

# plot
map_plot <-  ggplot() +
  geom_sf(data = world, fill = "white", color = "black") +
  geom_sf(data = subset(countries, name == "Indonesia"), fill = "white", color = "black") +
  geom_sf(data = subset(countries, name == "Malaysia"), fill = "white", color = "black") +
  ggtitle("Sampling sites: Indonesia and Malaysia") +
  coord_sf(xlim = c(90, 155), ylim = c(-35, 35), expand = FALSE) +
  scale_x_continuous(name = "Longitude (°E)", breaks = seq(90, 155, 10)) +
  scale_y_continuous(name = "Latitude (°N)", breaks = seq(-35, 35, 10)) +
  theme_minimal() +
  theme(
    panel.background = element_rect(fill = "grey90", color = NA),  # ocean
    plot.background  = element_rect(fill = "white", color = NA),   # outer plot bg
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )

map_plot

ggsave("/yourpathname/Fig.1_Indonesia_Malaysia_Map.png",
       plot = map_plot, width = 7, height = 10, units = "in")
```

# Thermal tolerance data from Indonesia and Malaysia

To examine the photosynthetic efficiency data from Malaysia and Indonesia, the R-package from Iakovleva, Colin & Voolstra, 2025 was used ([https://doi.org/10.5281/zenodo.8370644](https://doi.org/10.5281/zenodo.8370644.)).

The photosynthetic efficiency data from the four temperature profiles of each sampling location from Indonesia and Malysia was used to fit log-logistic dose-response curves, using the ‘drc’ R package. These curves were used to determine the effective dose 50 (ED50) for each sampling location (Evensen et al. 2021; Voolstra et al. 2021). The ED50 value defines the temperature at which photosynthetic efficiency has declined by 50% relative to the baseline measurement, and is therefore used as an empirically derived proxy for heat stress tolerance (Evensen et al. 2022). 

### Load Data ED50 data

```{r}
# Indonesia
# Upload Indonesia data
input_data_path <- selectFile(
  caption = "Select XLSX or CSV Input File")

# ID
cbass_dataset_ID <- read_data(input_data_path)

# To specify the prefix for output files
output_prefix <- tools::file_path_sans_ext(input_data_path)

rlog::log_info(paste("Your current directory is", getwd()))
rlog::log_info(paste("Your input filename is", basename(input_data_path)))
rlog::log_info(paste("The output files will be written into", output_prefix))

# Malaysia data
# upload Malaysia data
input_data_path <- selectFile(
  caption = "Select XLSX or CSV Input File")

# MY
cbass_dataset_MY <- read_data(input_data_path)

# To specify the prefix for output files
output_prefix <- tools::file_path_sans_ext(input_data_path)

rlog::log_info(paste("Your current directory is", getwd()))
rlog::log_info(paste("Your input filename is", basename(input_data_path)))
rlog::log_info(paste("The output files will be written into", output_prefix))

```

### Analysis and Preprocessing

```{r}
# ID
cbass_dataset_ID <- preprocess_dataset(cbass_dataset_ID)
validate_cbass_dataset(cbass_dataset_ID)

# MY
cbass_dataset_MY <- preprocess_dataset(cbass_dataset_MY)
validate_cbass_dataset(cbass_dataset_MY)
```

### Explore ED50

```{r}
# ID
# grouping properties Indonesia
grouping_properties <- c("Country", "Condition")

# PAM data Indonesia
drm_formula <- "PAM ~ Temperature"
models <- fit_drms(cbass_dataset_ID, grouping_properties, drm_formula, is_curveid = TRUE)
str(cbass_dataset_ID)

# PAM data 
ed50_ID <- get_ed50_by_grouping_property(models)
cbass_dataset_ID <- define_grouping_property(cbass_dataset_ID, grouping_properties) %>%
  mutate(GroupingProperty = paste(GroupingProperty, Genotype, sep = "_"))

ed50_df_ID <- 
  left_join(ed50_ID, cbass_dataset_ID, by = "GroupingProperty") %>%
  dplyr::select(names(ed50_ID), Depth, Genotype, all_of(grouping_properties)) %>%
  distinct()

# MY
# grouping properties
grouping_properties <- c("Country", "Condition")
# important: keep gouping property: Country, Condition to get the right EDS50 plot !!!!

# PAM data Malaysia
drm_formula <- "PAM ~ Temperature"
models <- fit_drms(cbass_dataset_MY, grouping_properties, drm_formula, is_curveid = TRUE)

# PAM data
ed50_MY <- get_ed50_by_grouping_property(models)
cbass_dataset_MY <- define_grouping_property(cbass_dataset_MY, grouping_properties) %>%
  mutate(GroupingProperty = paste(GroupingProperty, Genotype, sep = "_"))

ed50_df_MY <- 
  left_join(ed50_MY, cbass_dataset_MY, by = "GroupingProperty") %>%
  dplyr::select(names(ed50_MY), Depth, Genotype, all_of(grouping_properties)) %>%
  distinct()

```

### Adjust dataset for plotting

```{r}
# ID
# Create the Location column and name it as the sampling locazions based on the Condition column
cbass_dataset_ID <- cbass_dataset_ID %>%
  mutate(Location = case_when(
    Condition %in% c("Top") ~ "top",
    Condition %in% c("Bottom") ~ "bottom"))

ed50_df_ID <- ed50_df_ID %>%
  mutate(Location = case_when(
    Condition %in% c("Top") ~ "top",
    Condition %in% c("Bottom") ~ "bottom"))

# ID
str(cbass_dataset_ID)
cbass_dataset_ID$Genotype <- as.numeric(cbass_dataset_ID$Genotype)
cbass_dataset_ID$Location <- as.factor(cbass_dataset_ID$Location)
cbass_dataset_ID$Temperature <- as.factor(cbass_dataset_ID$Temperature)
cbass_dataset_ID$PAM <- as.factor(cbass_dataset_ID$PAM)



# MY
# add top, middle, bottom to the cbass MY dataset as sampling locations based on the Condition column 
cbass_dataset_MY <- cbass_dataset_MY %>%
  mutate(Location = case_when(
    Condition %in% c("E1", "W1") ~ "top",    
    Condition %in% c("E2", "W2") ~"upper_middle",
    Condition %in% c("E3", "W3") ~ "lower_middle",
    Condition %in% c("E4", "W4") ~ "bottom", 
    TRUE ~ NA_character_  
  ))

# also add to the ed50_df dataset
ed50_df_MY <- ed50_df_MY %>%
  mutate(Location = case_when(
    Condition %in% c("E1", "W1") ~ "top",    
    Condition %in% c("E2", "W2") ~"upper_middle",
    Condition %in% c("E3", "W3") ~ "lower_middle",
    Condition %in% c("E4", "W4") ~ "bottom", 
    TRUE ~ NA_character_  
  ))

# MY
# change genotype numbers to numeric so its the same as for the MY dataset
str(cbass_dataset_MY)
cbass_dataset_MY$Genotype <- as.numeric(cbass_dataset_MY$Genotype)
cbass_dataset_MY$Location <- as.factor(cbass_dataset_MY$Location)
```

### Temperature response curve

```{r}
# ID
# set order of the Locations
cbass_dataset_ID$Location <- factor(cbass_dataset_ID$Location, 
                                   levels = c("bottom", "lower_middle", "upper_middle", "top"))

str(cbass_dataset_ID)

# visualize temperature response curve 
exploratory_curve <-
  ggplot(data = cbass_dataset_ID,
       aes(
         x = Temperature,
         y = PAM,
         # You can play around with the group value (e.g., Species, Site, Condition)
         group = GroupingProperty,
         color = Location)) +
  geom_smooth(
    method = drc::drm,
    method.args = list(
      fct = drc::LL.3()),
    se = FALSE,
    size = 0.7
  ) +
  geom_point(size = 1.5) +
  facet_wrap( ~ Genotype, nrow = 2) +
  scale_color_brewer(palette = "Paired") # Colorblind-friendly palette

#plot
exploratory_curve


# MY
# set order of the Locations
cbass_dataset_MY$Location <- factor(cbass_dataset_MY$Location, 
                                   levels = c("bottom", "lower_middle", "upper_middle", "top"))

# visualize temperature response curve
exploratory_curve <-
  ggplot(data = cbass_dataset_MY,
       aes(
         x = Temperature,
         y = PAM,
         # You can play around with the group value (e.g., Species, Site, Condition)
         group = GroupingProperty,
         color = Location)) +
  geom_smooth(
    method = drc::drm,
    method.args = list(
      fct = drc::LL.3()),
    se = FALSE,
    size = 0.7
  ) +
  geom_point(size = 1.5) +
  facet_wrap( ~ Genotype, nrow = 2) +
  scale_color_brewer(palette = "Paired") # Colorblind-friendly palette

# plot
exploratory_curve
```

## **Fig 2. Thermal tolerance across the surface of massive Porites sp. colonies from Indonesia and Malaysia.**

```{r}
# Indonesia
# only select needed columns from the indonesia dataset
ed50_df_ID_filtered <- ed50_df_ID %>%
  dplyr::select(c("ED50", "Depth", "Location", "Genotype", "Country"))

# calculate mean ED50 per colony/genotype and calculate differences
ED50_ID_means <- ed50_df_ID_filtered %>%
  group_by(Genotype) %>%
  summarise(mean_ED50 = mean(ED50))

Final_ED50_ID <- ed50_df_ID_filtered %>%
  left_join(ED50_ID_means, by = c("Genotype")) %>%
  mutate(ED50_diff = ED50 - mean_ED50) %>%
  select(-contains(".x")) %>% # remove duplicates
  distinct()  # Re

# change genotype numbre from 01 to 1
Final_ED50_ID$Genotype <- as.numeric(Final_ED50_ID$Genotype)

# prepare for plotting
# set the order of the locations
Final_ED50_ID$Location <- factor(Final_ED50_ID$Location, 
                                   levels = c("bottom", "top"))

# set the order of the coral colonies
Final_ED50_ID <- Final_ED50_ID %>%
  mutate(
    Genotype = factor(Genotype, levels = c("1", "2", "3", "4", "5", "6", "7"))  # Order genotypes
  )

# we don't have ITS2 and 16S data for colony 5 so remove colony 5 also from the ED50 data
# Remove Genotype 5 from the dataset
Final_ED50_ID_filtered <- Final_ED50_ID[Final_ED50_ID$Genotype != "5", ]

 # set the colors               
genotype_colors <- c("1" = "pink4", 
                     "2" = "orchid4", 
                     "3" = "pink2", 
                     "4" = "mistyrose2", 
                     "6" = "lightcoral",
                     "7" = "hotpink4")

# ED50 Difference Bar Plot
ED50_diff_barplot_ID <- ggplot(Final_ED50_ID_filtered, aes(x = ED50_diff, y = Location, fill = Genotype)) +
  geom_vline(aes(xintercept = 0), color = "black", linetype = "dashed", size = 1) +
  geom_vline(xintercept = -0.5, color = "grey40", linetype = "dashed") +
  geom_vline(xintercept = 0.5, color = "grey40", linetype = "dashed") +
  geom_bar(stat = "identity", position = "dodge") +
  facet_wrap(~Genotype, nrow = 6) +
  xlim(-1, 1) +
  scale_y_discrete(labels = c("Bottom", "Top")) +
  scale_fill_manual(values = genotype_colors) +
  labs(
    x = "Deviation from Mean ED50 [°C]", 
    y = "Location within Colony",
    fill = "Genotype"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    text = element_text(size = 13),
    strip.text = element_blank(),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 13),
    legend.position = "none",
    panel.spacing = unit(0.3, "lines"),
    strip.background = element_blank()
  )


# ED50 Violin Plot
ED50_violin_plot_ID <- ggplot(Final_ED50_ID_filtered, aes(x = 1, y = ED50, fill = Genotype)) +
  geom_violin() +
  facet_wrap(~Genotype, nrow = 7) +
  stat_summary(fun = "mean", geom = "point", color = "black", size = 3, shape = 18, show.legend = FALSE) +
  stat_summary(fun = "mean", geom = "text", color = "black", size = 4,
               aes(label = round(..y.., 2)), vjust = -1) +
  scale_y_continuous(breaks = c(37, 38), limits = c(36.5, 38.5)) +
  scale_fill_manual(values = genotype_colors) +
  labs(
    x = NULL,
    y = "ED50 [°C]",
    fill = "Genotype"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    text = element_text(size = 13),
    strip.text = element_blank(),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 13),
    legend.position = "none",
    panel.spacing = unit(0.3, "lines"),
    strip.background = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  )


# Combine plots
combined_plot_ID <- ED50_violin_plot_ID + ED50_diff_barplot_ID + 
  plot_layout(ncol = 2, widths = c(0.4, 0.8))

# Display combined plot
combined_plot_ID

# save as image
ggsave("/yourpathname/Fig3_ED50_ID_within_colonies.png", plot = combined_plot_ID, width = 12, height = 10, dpi = 300, bg = "white")

# save as pdf
ggsave("/yourpathname/Fig3_ED50_ID_within_colonies.pdf", 
       plot = combined_plot_ID, 
       width = 12, 
       height = 10, 
       dpi = 300, 
       bg = "white")



# Malayisa
# merge datasets from MY to combine ed50 data with general sampling data
Final_ED50_MY  <- merge(cbass_dataset_MY, ed50_df_MY)

# create the column Orientation
Final_ED50_MY <- Final_ED50_MY %>%
  mutate(Orientation = case_when(
    Site %in% c("E") ~ "East",
    Site %in% c("W") ~ "West"))

# set order of the Locations
Final_ED50_MY$Location <- factor(Final_ED50_MY$Location, 
                                   levels = c("bottom", "lower_middle", "upper_middle", "top"))
 
# set order of the colonies with increasing depth
Final_ED50_MY <- Final_ED50_MY %>%
  mutate(
    Genotype = factor(Genotype, levels = c("1", "6", "2", "3", "5", "4")))

# adjust Orientation column
Final_ED50_MY <- Final_ED50_MY %>%
  mutate(
    Orientation = factor(Orientation, levels = c("West", "East")))

# Compute mean ED50 per Genotype
ED50_MY_means <- Final_ED50_MY %>%
  group_by(Genotype) %>%
  summarise(mean_ED50 = mean(ED50))

# Merge back to main dataset (on Genotype *and* Orientation)
ED50_MY_means$Genotype <- as.factor(ED50_MY_means$Genotype)
str(ED50_MY_means)

# merge datasets
Final_ED50_MY <- Final_ED50_MY %>%
  left_join(ED50_MY_means, by = c("Genotype")) 


Final_ED50_MY <- Final_ED50_MY %>%
  mutate(ED50_diff = ED50 - mean_ED50) %>%  # Compute difference
  select(-contains(".x")) %>%  # Remove duplicate columns (like Orientation.x)
  distinct()  # Re
           
# set the colors    
 genotype_colors <- c("1" = "skyblue3", 
                     "2" = "lightblue", 
                     "3" = "dodgerblue4", 
                     "4" = "cornflowerblue", 
                     "5" = "lightskyblue", 
                     "6" = "dodgerblue3")

# Bar plot
ED50_diff_barplot_MY <- ggplot(Final_ED50_MY, aes(x = ED50_diff, y = Location, fill = Genotype)) +
  geom_vline(aes(xintercept = 0), color = "black", linetype = "dashed", size = 1) +
  geom_vline(xintercept = -0.5, color = "grey40", linetype = "dashed") +
  geom_vline(xintercept = 0.5, color = "grey40", linetype = "dashed") +
  geom_bar(stat = "identity", position = "dodge") +
  facet_grid(Genotype ~ Orientation, scales = "free_y") +
  xlim(-1, 1) +
  scale_y_discrete(labels = c("Bottom", "Lower Middle", "Upper Middle", "Top")) +
  scale_fill_manual(values = genotype_colors) +
  labs(
    x = "Deviation from Mean ED50 [°C]", 
    y = "Location within Colony",
    fill = "Genotype"
  ) +
  theme_minimal(base_size = 13) +  # Global text size
  theme(
    text = element_text(size = 13),
    strip.text = element_blank(),
    axis.title = element_text(size = 13, face = "bold"),  
    axis.text = element_text(size = 13),  
    legend.position = "none",
    panel.spacing = unit(0.4, "lines"),
    strip.background = element_blank()
  )


# Violin plot
ED50_violin_plot_MY <- ggplot(Final_ED50_MY, aes(x = 1, y = ED50, fill = Genotype)) +
  geom_violin() +
  facet_wrap(~Genotype, nrow = 6) +
  stat_summary(fun = "mean", geom = "point", color = "black", size = 3, shape = 18, show.legend = FALSE) + 
  stat_summary(fun = "mean", geom = "text", color = "black", size = 4,  # Keep this a little larger for visibility
               aes(label = round(..y.., 2)), vjust = -1) +
  scale_fill_manual(values = genotype_colors) +
  guides(fill = guide_legend(title = "Coral Colony")) +
  labs(
    x = NULL,
    y = "ED50 [°C]",
    fill = "Genotype"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    text = element_text(size = 13),
    strip.text = element_blank(),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 13),
    legend.position = "none",
    panel.spacing = unit(0.4, "lines"),
    strip.background = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  )


# Combine
combined_plot_MY <- ED50_violin_plot_MY + ED50_diff_barplot_MY +
  plot_layout(ncol = 2, widths = c(0.4, 0.8))

combined_plot_MY



# plot together
# Combine plots into one row
final_combined_plot <-  ED50_diff_barplot_ID + ED50_diff_barplot_MY +
  plot_layout(ncol = 4, widths = c(0.4, 0.8, 0.4, 0.8)) +
  plot_annotation(
    theme = theme(
      plot.margin = margin(10, 10, 10, 10)
    )
  )

# Final Figure
final_combined_plot

# figure was safed as PDF and edited in Affinity designer

# Save as PDF with same dimensions
ggsave(
  filename = "ITS2_MY_ID.pdf",
  plot = final_combined_plot,
  width = 11.7,
  height = 8,
  units = "in"
)
getwd()
```

### Testing for significant differences in thermal tolerance

```{r}
# Indonesia
# check for significant differences
# data distribution
hist(Final_ED50_ID$ED50) # looks like its not normally distributed, but also not many data points
shapiro.test(Final_ED50_ID$ED50) # p-value: 0.8591, not signfiicant, so data is normaly distributed
qqnorm(Final_ED50_ID$ED50) # normal distributed

# data distribution
hist(Final_ED50_ID$ED50_diff) # looks like its not normally distributed
shapiro.test(Final_ED50_ID$ED50_diff) # p-value: 0.5761, normaly distributed
qqnorm(Final_ED50_ID$ED50_diff) # normal distributed

# across colonies
# can't test for significant differences across genotypes because for each genotpye I only have one datapoint
ggplot(Final_ED50_ID, aes(x = Genotype, y = ED50, color = Location )) +
  geom_point(size = 3)+
  facet_wrap(~Location)

# within colonies ED50
# check for differences in variance between the top and bottom sampling location
leveneTest(ED50 ~ Location, data = Final_ED50_ID)
# p-value = 0.9143, not significant difference in the variance of groups, can use ANOVA

# two groups only top and bottom
t.test(ED50 ~ Location, data = Final_ED50_ID)
# p-value = 0.797 not significant 


# Malaysia

# filter data
Final_ED50_MY <- Final_ED50_MY %>% 
  filter(Temperature == 30)

# check data distribution: normality and homogeneity
hist(Final_ED50_MY$ED50) # could be normally distirbuted
# normallity test
shapiro.test(Final_ED50_MY$ED50) # p-value: 0.1506, so we do not reject the 0-hypthesis --> data normally distributed 
# plot normallity
qqnorm(Final_ED50_MY$ED50) # looks like its normally distributed

# Ed50 across depth
# check for corrleation of ED50 and Depth, data is normally distributed but check what relationship deoth and ED50 have (linear, non-linear)

# what model fits best: linear or gam?
lm_model <- lm(ED50 ~ Depth, data = Final_ED50_MY)
summary(lm_model) # p-value = <2e-16 ***
gam_model <- gam(ED50 ~ s(Depth), data = Final_ED50_MY)
summary(gam_model) # <2e-16 ***

# check for the best fitting model
AIC(lm_model, gam_model)

# gam fits better has lower AIC, so data is rather not linear -> use spearman to check for correlation
cor.test(Final_ED50_MY$ED50, Final_ED50_MY$Depth, method = "spearman")
# p-value < 7.013e-06
# rho: -0.598 --> -0.60

# within colonies
# check that variance is normally distributed before using a normaly distributed test (e.g. ANOVA)
leveneTest(ED50 ~ Location, data = Final_ED50_MY)
# p-value = 0.2624 --> not significan means the variance in groups is equal so variance in groups is equal we can use ANOVA

# test for significance using ANOVA (4 locations within colonies)
anova_result <- aov(ED50 ~ Location, data = Final_ED50_MY)
summary(anova_result)
# p-value =0.677

# check for East
Final_ED50_East <- subset(Final_ED50_MY, Orientation == "East")

# Check for West
Final_ED50_West <- subset(Final_ED50_MY, Orientation == "West")

# ANOVA for East
# test homogeneity
leveneTest(ED50 ~ Location, data = Final_ED50_East)
# p-value: 0.725, not significant use ANOVA
anova_east <- aov(ED50 ~ Location, data = Final_ED50_East)
summary(anova_east)
# p-value= 0.84, not signf

# ANOVA for West
# test homogeneity
leveneTest(ED50 ~ Location, data = Final_ED50_West)
# p-value = 0.1565, not signficant use ANOVA
anova_west <- aov(ED50 ~ Location, data = Final_ED50_West)
summary(anova_west)
# p-value: 0.884


###########
# check for signficant differences in the ED50_Diff
# data distirbution
hist(Final_ED50_MY$ED50_diff) #  normally distirbuted, or slightly left skewed
# normality
shapiro.test(Final_ED50_MY$ED50_diff) # P-value above 0.8281, data normally distirbuted
# plot normallity
qqnorm(Final_ED50_MY$ED50_diff, main = "Q-Q Plot of ED50") # looks like its normally distribute
# test homogeneity
leveneTest(ED50_diff ~ Location, data = Final_ED50_MY)
# p-value = 0.418, not signficant use ANOVA

# normally distributed data
anova_result <- aov(ED50_diff ~ Location, data = Final_ED50_MY)
summary(anova_result)
# p-value = 0.251 not significant

# ANOVA for East
anova_east <- aov(ED50_diff ~ Location, data = Final_ED50_East)
summary(anova_east)
# p-value = 0.595

# ANOVA for West
anova_west <- aov(ED50_diff ~ Location, data = Final_ED50_West)
summary(anova_west)
# p-value = 0.880
```

# ITS2 type profiles - Symbiodiniaceae community composition

ITS2 sequencing data was analysed using SymPortal (Hume et al., 2019). ITS2 type profiles being representative of putative Symbiodiniaceae genotypes were characterized by specific sets of defining intragenomic ITS2 sequence variants (DIVs) (Hume et al. 2019) and visualized using R studio.

### Indonesia

```{r}
# ITS2 type profile metadata
input_data_path <- selectFile(
  caption = "Select XLSX or CSV Input File")

# ITS2 profile data
its2_metadata_ID <- read_data(input_data_path)


#upload ITS2 profile absolute abundance
its2_profile_ID <- read.table('/yourpathname/582_20250509T135802_DBV_20250510T135051.profiles.relative.abund_and_meta.txt', header = TRUE, sep = "\t", stringsAsFactors = FALSE)

# rename column names
colnames(its2_profile_ID)[1] <- "ITS2_profile"
colnames(its2_profile_ID)[2] <- "sample_name"

# merge dataframes
its2_data_ID <- merge(its2_metadata_ID, its2_profile_ID, by = "sample_name", all.x = TRUE)

# prepare for plotting
# Convert to long format for ggplot
its2_profile_ID_long <- its2_data_ID %>%
  gather(key = "ITS2_profile", value = "value", C15.C15aae:C15.C15ad.C15ai.C116.C116aa)

# rename upper and lower middle 1 to just lower and upper midlee for colony 2
its2_profile_ID_long <- its2_profile_ID_long %>%
  mutate(collection_depthrange = recode(collection_depthrange,
                                       "top" = "Top",
                                      "bottom" = "Bottom"))

# order data for plotting
its2_profile_ID_long <- its2_profile_ID_long %>%
  mutate(
    collection_depthrange= factor(collection_depthrange, levels = c("Bottom", "Top")))


# set order of the colonies with decreasing depth
its2_profile_ID_long <- its2_profile_ID_long %>%
  mutate(
    collection_colony = factor(collection_colony, levels = c("1", "2", "3", "7", "4", "6")))


# plot
ITS2_plot_ID <- ggplot(its2_profile_ID_long, aes(x = value, y = collection_depthrange, fill = ITS2_profile)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_x_continuous(labels = scales::percent_format(), expand = c(0, 0)) +
  facet_grid(collection_colony ~ ., scales = "free_y") +
  theme_minimal(base_size = 13) +
    guides(fill = guide_legend(ncol = 1)) +
  labs(
    x = "Relative Abundance [%]",
    y = "Location within Colony",
    fill = "ITS2 Type Profiles"
  ) +
  scale_fill_manual(
    values = c(
      "C15.C15aae" = '#006633',
      "C15.C15bn.C15by" = "#99CC33",
      "C15.C116aa.C15ad" = "#339900",
      "C57d.C57a" = "#000099",
      "C15.C15kl.C15he" = "#009966",
      "C15.C15ad.C15ai.C116.C116aa" = "#6699CC"
    ),
    labels = c(
      "C15.C15aae" = "C15-C15aae",
      "C15.C15bn.C15by" = "C15-C15bn-C15by",
      "C15.C116aa.C15ad" = "C15/C116aa-C15ad",
      "C57d.C57a" = "C57d/C57a",
      "C15.C15kl.C15he" = "C15-C15kl-C15he",
      "C15.C15ad.C15ai.C116.C116aa" = "C15-C15ad-C15ai-C116-C116aa"
    )
  ) +
  theme(
    legend.position = "bottom",
    legend.box = "horizontal",
    legend.text = element_text(size = 13),
    legend.title = element_text(size = 13),
    axis.text = element_text(size = 13),
    axis.title = element_text(size = 13),
    strip.text = element_text(size = 13),
    panel.grid = element_blank(),
    panel.border = element_blank()
  )

ITS2_plot_ID
```

### Malaysia

```{r}
# upload ITS2 type profile meta data
input_data_path <- selectFile(
  caption = "Select XLSX or CSV Input File")

its2_metadata <- read_data(input_data_path)

#upload ITS2 profile absolute abundance
its2_profile <- read.table('/yourpathname/560_20250219T125105_DBV_20250221T224139.profiles.relative.abund_and_meta.txt', header = TRUE, sep = "\t", stringsAsFactors = FALSE)

# rename column 1
colnames(its2_profile)[1] <- "sample_name"

# Remove additional sampling locations of colony 2 from the its2_profile data frame (in the txt file these additional sampling locations are in)
its2_profile <- its2_profile %>%
  filter(!str_detect(sample_name, "[EW][56]$"))

# merge dataframes
its2_data<- merge(its2_metadata, its2_profile, by = "sample_name", all.x = TRUE)

# prepare for plotting
# Convert to long format for ggplot
its2_data_long <- its2_data %>%
  gather(key = "ITS2_profile", value = "value", C15.C15ev.C15dt.C15kl.C15pm:C15.C15by.C15ai)

# change the naming in collection_location
# rename upper and lower middle 1 to just lower and upper midlee for colony 2
its2_data_long <- its2_data_long %>%
  mutate(collection_depthrange = recode(collection_depthrange,
                                      "lower_middle" = "Lower Middle",
                                      "upper_middle" = "Upper Middle",
                                       "top" = "Top",
                                      "bottom" = "Bottom"))

# organize data
# Ensure correct order for plotting
its2_data_long <- its2_data_long %>%
  mutate(
    collection_location = factor(collection_depthrange, levels = c("Bottom", "Lower Middle","Upper Middle", "Top")),
    collection_orientation = factor(collection_orientation, levels = c("West", "East"))
  )

# set order of the colonies with decreasing depth
its2_data_long <- its2_data_long %>%
  mutate(
    collection_colony = factor(collection_colony, levels = c("1", "6", "2", "3", "5", "4")))

#plot
ITS2_plot_MY <- ggplot(its2_data_long, aes(x = value, y = collection_location, fill = ITS2_profile)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_x_continuous(labels = scales::percent_format(), expand = c(0, 0)) +
  facet_grid(collection_colony ~ collection_orientation, scales = "free_y") +
  theme_minimal(base_size = 13) +
  guides(fill = guide_legend(ncol = 1)) +
  labs(
    x = "Relative Abundance",
    y = "Location within Colony",
    fill = "ITS2 Type Profiles"
  ) +
  scale_fill_manual(
    values = c(
      "C15.C15ev.C15dt.C15kl.C15pm" = "#66CC66",
         "C15.C15ad.C15ai.C116.C116aa" = "#6699CC",
      "C15.C15ev.C15by" = "#66CC99",
      "C15.C15ev.C15dt" = "#009999",
      "C15.C15ev" = "#99CCCC",
        "C15.C15by.C15ai" = "#006666"
    ),
    labels = c(
      "C15.C15ev.C15dt.C15kl.C15pm" = "C15-C15ev-C15dt-C15kl-C15pm",
      "C15.C15ad.C15ai.C116.C116aa" = "C15-C15ad-C15ai-C116-C116aa",
      "C15.C15ev.C15by" = "C15-C15ev-C15by",
      "C15.C15ev.C15dt" = "C15-C15ev-C15dt",
      "C15.C15ev" = "C15-C15ev",
      "C15.C15by.C15ai" = "C15-C15by-C15ai"
    )
  ) +
  theme(
    legend.position = "bottom",
    legend.box = "horizontal",
    legend.text = element_text(size = 13),
    legend.title = element_text(size = 13),
    axis.text = element_text(size = 13),
    axis.title = element_text(size = 13),
    strip.text = element_text(size = 13),
    panel.grid = element_blank(),
    panel.border = element_blank()
  )

ITS2_plot_MY
```

## **Fig. 3 ITS2 type profiles across the surface of massive Porites sp. colonies from Indonesia and Malaysia.**

```{r}
# Combine plots side-by-side with equal width
final_combined_plot <- ITS2_plot_ID + ITS2_plot_MY +
  plot_layout(ncol = 2, widths = c(1, 1)) +  # Equal widths for both plots
  plot_annotation(
    theme = theme(
      plot.margin = margin(10, 10, 10, 10)
    )
  )
final_combined_plot

# plot as saved and edited with Affinity designer.

# Save to A4 size (portrait orientation) as PDF
ggsave(
  filename = "/yourpathname/ITS2_MY_ID_new.pdf",
  plot = final_combined_plot,
  width = 11.7,      # A4 width in inches
  height = 10,     # A4 height in inches
  units = "in"
)
```

### Supplementary Fig. 1 ITS2 sequences - absolute abundance

```{r}
# plot absolute abundance of ITS2 sequences
# Indonesia
# upload data
its2_sequences_ID <- read.table('/yourpathname/582_20250509T135802_DBV_20250510T135051.seqs.absolute.abund_and_meta.txt', header = TRUE, sep = "\t", stringsAsFactors = FALSE)

# remove columns with meta data only keep sample name and ITS2 sequences
its2_sequences_ID_filtered <- its2_sequences_ID[, -c(1, 3:39)]

# remove rows with NA
its2_sequences_ID_filtered <- its2_sequences_ID_filtered %>%
  filter(sample_name != "" & !is.na(sample_name))

# Convert to long format for ggplot
its2_sequences_ID_filtered_long <- its2_sequences_ID_filtered %>%
  pivot_longer(
    cols = -sample_name,          # all columns except SampleID
    names_to = "Sequence",
    values_to = "AbsoluteAbundance"
  )

# make sure the order of sequences is correct (highest abundnace at the bottom of the barplot)
seq_ID_order <- its2_sequences_ID_filtered_long %>%
  group_by(Sequence) %>%
  summarize(total_abundance = sum(AbsoluteAbundance, na.rm = TRUE)) %>%
  arrange(desc(total_abundance)) %>%      # highest abundance first
  pull(Sequence)

# Set factor levels so that highest abundance is at the bottom
its2_sequences_ID_filtered_long$Sequence <- factor(
  its2_sequences_ID_filtered_long$Sequence,
  levels = seq_ID_order
)

# set colors
palette_marlen <- c(
             "#E704C4", "#C2FFED", "#A079BF", "#CC0744", "#C0B9B2", "#C2FF99", "#001E09", "#00489C", "#6F0062",
            "#0CBD66",
            "#EEC3FF", "#456D75", "#B77B68", "#7A87A1", "#788D66", "#885578", "#FAD09F", "#FF8A9A", "#D157A0",
            "#BEC459",
            "#456648", "#0086ED", "#886F4C", "#34362D", "#B4A8BD", "#00A6AA", "#452C2C", "#636375", "#A3C8C9",
            "#FF913F",
            "#938A81", "#575329", "#00FECF", "#B05B6F", "#8CD0FF", "#3B9700", "#04F757", "#C8A1A1", "#1E6E00",
            "#7900D7",
            "#A77500", "#6367A9", "#A05837", "#6B002C", "#772600", "#D790FF", "#9B9700", "#549E79", "#FFF69F",
            "#201625",
            "#72418F", "#BC23FF", "#99ADC0", "#3A2465", "#922329", "#5B4534", "#FDE8DC", "#404E55", "#0089A3",
            "#CB7E98",
            "#A4E804", "#324E72", "#6A3A4C", "#83AB58", "#001C1E", "#D1F7CE", "#004B28", "#C8D0F6", "#A3A489",
            "#806C66",
            "#222800", "#BF5650", "#E83000", "#66796D", "#DA007C", "#FF1A59", "#8ADBB4", "#1E0200", "#5B4E51",
            "#C895C5",
            "#320033", "#FF6832", "#66E1D3", "#CFCDAC", "#D0AC94", "#7ED379", "#012C58", "#7A7BFF", "#D68E01",
            "#353339",
            "#78AFA1", "#FEB2C6", "#75797C", "#837393", "#943A4D", "#B5F4FF", "#D2DCD5", "#9556BD", "#6A714A",
            "#001325",
            "#02525F", "#0AA3F7", "#E98176", "#DBD5DD", "#5EBCD1", "#3D4F44", "#7E6405", "#02684E", "#962B75",
            "#8D8546",
            "#9695C5", "#E773CE", "#D86A78", "#3E89BE", "#CA834E", "#518A87", "#5B113C", "#55813B","#FFB500",
            "#00005F",
            "#A97399", "#4B8160", "#59738A", "#FF5DA7", "#F7C9BF", "#643127", "#513A01", "#6B94AA", "#51A058",
            "#A45B02",
            "#1D1702", "#E20027", "#E7AB63", "#4C6001", "#9C6966", "#64547B", "#97979E", "#006A66", "#391406",
            "#F4D749",
            "#0045D2", "#006C31", "#DDB6D0", "#7C6571", "#9FB2A4", "#00D891", "#15A08A", "#BC65E9", "#FFFFFE",
            "#C6DC99",
            "#203B3C", "#671190", "#6B3A64", "#F5E1FF", "#FFA0F2", "#CCAA35", "#374527", "#8BB400", "#797868",
            "#C6005A",
            "#3B000A", "#C86240", "#29607C", "#402334", "#7D5A44", "#CCB87C", "#B88183", "#AA5199", "#B5D6C3",
            "#A38469",
            "#9F94F0", "#A74571", "#B894A6", "#71BB8C", "#00B433", "#789EC9", "#6D80BA", "#953F00", "#5EFF03",
            "#E4FFFC",
            "#1BE177", "#BCB1E5", "#76912F", "#003109", "#0060CD", "#D20096", "#895563", "#29201D", "#5B3213",
            "#A76F42",
            "#89412E", "#1A3A2A", "#494B5A", "#A88C85", "#F4ABAA", "#A3F3AB", "#00C6C8", "#EA8B66", "#958A9F",
            "#BDC9D2",
            "#9FA064", "#BE4700", "#658188", "#83A485", "#453C23", "#47675D", "#3A3F00", "#061203", "#DFFB71",
            "#868E7E",
            "#98D058", "#6C8F7D", "#D7BFC2", "#3C3E6E", "#D83D66", "#2F5D9B", "#6C5E46", "#D25B88", "#5B656C",
            "#00B57F",
            "#545C46", "#866097", "#365D25", "#252F99", "#00CCFF", "#674E60", "#FC009C", "#92896B")

# plot
ID_ITS2seq <- ggplot(its2_sequences_ID_filtered_long, aes(x = sample_name, y = AbsoluteAbundance, fill = Sequence)) +
  geom_bar(stat = "identity", position = position_stack(reverse = TRUE)) +
  scale_fill_manual(values = palette_marlen ) +
  labs(x = "Sample", y = "Relative Abundance") +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "none"   # hide legend
  )
ID_ITS2seq 



# Malaysia
# upload data
its2_sequences <- read.table('/yourpathname/560_20250219T125105_DBV_20250221T224139.seqs.absolute.abund_and_meta.txt', header = TRUE, sep = "\t", stringsAsFactors = FALSE)

# remove columns with meta data only keep sample name and ITS2 sequences
its2_sequences_filtered <- its2_sequences[, -c(1, 3:39)]

# remove rows with NA
its2_sequences_filtered <- its2_sequences_filtered %>%
  filter(sample_name != "" & !is.na(sample_name))

# Convert to long format for ggplot
its2_sequences_filtered_long <- its2_sequences_filtered %>%
  pivot_longer(
    cols = -sample_name,          # all columns except SampleID
    names_to = "Sequence",
    values_to = "AbsoluteAbundance"
  )

# make sure the order of sequences is correct (highest abundnace at the bottom of the barplot)
seq_order <- its2_sequences_filtered_long %>%
  group_by(Sequence) %>%
  summarize(total_abundance = sum(AbsoluteAbundance, na.rm = TRUE)) %>%
  arrange(desc(total_abundance)) %>%      # highest abundance first
  pull(Sequence)

# Set factor levels so that highest abundance is at the bottom
its2_sequences_filtered_long$Sequence <- factor(
  its2_sequences_filtered_long$Sequence,
  levels = seq_order
)

# plot
MY_ITS2_seq <- ggplot(its2_sequences_filtered_long, aes(x = sample_name, y = AbsoluteAbundance, fill = Sequence)) +
  geom_bar(stat = "identity", position = position_stack(reverse = TRUE)) +
  scale_fill_manual(values = palette_marlen) +
  labs(x = "Sample", y = "Relative Abundance") +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "none"   # hide legend
  )

```

# 16S - bacterial community composition

16S sequencing data was processed in R studio using the DADA2 pipeline. DADA2 (Callahan et al., 2016) filtered and trimmed reads, calculated error rates, inferred sequence variants, merged paired reads, and removed chimeras. Taxonomy was assigned by using the Silva v138 dataset from the Silva reference database (Benjamin, 2024). The top 20 bacterial families across the colony surface were visualised using R studio. <https://benjjneb.github.io/dada2/tutorial.html>

### Malaysia: DADA2

```{r}
# set working directory
# read file
cat("Reading files") # prints a message to indicate the script is starting to read files
path <- "." # current working directory
list.files(path) # lists all files in the current directory 

# Forward and reverse fastq filenames have format: 
fnFs <- sort(list.files(path, pattern="_R1_001.fastq", full.names = TRUE))
fnRs <- sort(list.files(path, pattern="_R2_001.fastq", full.names = TRUE))
# searches for files matching specific patterns _R1_001...and sorts files, get a sorted list of file paths for forward (fnFs) and reverse reads (fnRs)

# Extract sample names
sample.names <- sapply(strsplit(basename(fnFs), "__|_0"), function(x) x[2])

cat("Processing",length(sample.names),"samples:", sample.names)
# outputs a summary of the number of samples detected and their names, legth(): counts the number, cat():prints the message

#Inspect read quality profiles
plotQualityProfile(fnFs[1:2])
# generates a quality profile plot for the first two forward read files in the fnFs list, DADA2 package in R for processing high-throughput sequence data, to visualize the quality score of sequencing reads across all base positions in a FASTQ file
# selected the first two 1:2 forward read files in fnFs 

# gray-scale: heat map of the frequency of each quality score at each base position
# green line: mean quality score at each position
# orange line: quartiles of the quality score distribution
# red line: scaled proportion of reads that extend to at least that position

# forwards reads
# --> Truncate the forward reads at position 280

plotQualityProfile(fnRs[1:2])

# reversed reads
# worse quality, especially at the end which is common in Illumina sequencing
# DADA2 incorporates quality information into its error model which makes the algorithm robust to lower quality sequence

# reversed reads
# --> Truncate the forward reads at position 200

```

### Filtering and Trimming

This pipeline ensures that only high-quality reads are retained for downstream analysis, which is crucial for the accuracy of biological insights derived from sequencing data.

```{r}
# filter and trimm 
cat("Filtering and trimming")

# prints a message to indicate that the filtering and triming steps are about to begin in the sequencing data processing pipeline

# Place filtered files in filtered/ subdirectory
filtFs <- file.path(path, "filtered", paste0(sample.names, "_F_filt.fastq.gz"))
filtRs <- file.path(path, "filtered", paste0(sample.names, "_R_filt.fastq.gz"))
names(filtFs) <- sample.names
names(filtRs) <- sample.names

### set parameters  maxN=0 (DADA2 requires no Ns), truncQ=2, rm.phix=TRUE and maxEE=2. The maxEE parameter sets the maximum number of “expected errors” allowed in a read, which is a better filter than simply averaging quality scores.
out <- filterAndTrim(fnFs, filtFs, fnRs, filtRs, truncLen=c(280,200),maxN=0, maxEE=c(2,2), truncQ=5, rm.phix=TRUE, compress=TRUE, multithread=TRUE) 

head(out)
```

### Learn error rates & sample interference

```{r}
# learn error rates
# The error model is learned using the filtered and trimmed sequences, which helps in determining the probability of errors at each base position across the reads.Forward (errF) and reverse (errR) errors are learned separately because the forward and reverse reads often have different error profiles.

cat("Learning error rates")
errF <- learnErrors(filtFs, multithread=TRUE)
errR <- learnErrors(filtRs, multithread=TRUE)
# This function in DADA2 is used to learn the error rates for sequencing reads.
# multithread=TRUE: Enables multi-threading to speed up the error rate learning process if the system supports it.

# visualize the esstimated error rates as a sanity check
plotErrors(errF, nominalQ = T)
# everything looks resonable (black line follows the points, the model fits) and we proceed with confidence

```

### Sample Inference

```{r}
# This step is essential for transforming the raw sequencing data into biologically meaningful information by correcting errors and identifying the true sequences present in your samples
# apply the core sample inference algorithm to the filtered and trimmed sequence data.
dadaFs <- dada(filtFs, err=errF, multithread=TRUE, pool = T)
dadaRs <- dada(filtRs, err=errR, multithread=TRUE, pool = T)
# It uses the error rates learned from the reverse reads (errR) to apply the error-correction model and infer the true biological sequences.

#Inspecting the returned dada-class object:
dadaFs[[1]]
dadaRs[[1]]
```

### Merging paired reads and checking on read length

```{r}
mergers <- mergePairs(dadaFs, filtFs, dadaRs, filtRs, verbose=TRUE)
# merges forward (F) and reveres (R) sequences and filtered data in one column
# Inspect the merger data.frame from the first sample
head(mergers[[1]])

# Construct sequencing table
seqtab <- makeSequenceTable(mergers)
# creates a sequence table from the merged paired-end reads. A sequence table is a matrix where: rows represent individual samples, columns represent unique DNA sequences, entries contain the counts of each sequence in each sample
dim(seqtab)
# returns the dimensions of the sequence table (1st value number of rows (sample), 2nd value is the number of columns (unique sequences))

# Inspect distribution of sequence lengths
table(nchar(getSequences(seqtab)))
# looks good
# check visually how the districution of read length is
# Get the read length distribution
length_table <- table(nchar(getSequences(seqtab)))

# Convert to data frame for plotting
length_df <- as.data.frame(length_table)
colnames(length_df) <- c("Length", "Count")

# Make sure Length is numeric
length_df$Length <- as.numeric(as.character(length_df$Length))

ggplot(length_df, aes(x = Length, y = Count)) +
  geom_col(fill = "steelblue") +  # vertical bars
  labs(title = "Distribution of Read Lengths",
       x = "Read Length (bp)",
       y = "Number of Reads") +
  xlim(280, 400)+
  theme_minimal()
```

### Remove Chimeras

```{r}
# Remove chimeras  
# Chimeric sequences are artifacts of the PCR amplification process. These sequences do not represent real biological sequences but rather result from errors in the molecular processes used to prepare and sequence DNA.
# Chimeric sequences can inflate diversity estimates. Removing them ensures that the analysis focuses only on true biological sequences.
# This code removes chimeric sequences from the sequence table, calculates the dimensions of the updated sequence table and determines the proportion of non-chimeric sequences retained.

seqtab.nochim_MY <- removeBimeraDenovo(seqtab, method="consensus", multithread=TRUE, verbose=TRUE)
# removeBim..: identifies and removes chimeric sequences from the sequence table
# method: chimera removal method that identifies chimeras based on agreement across samples
# multithread=T: enables multi-threading for faster computation
# verbose=T: provides detailed output about the chimera removal process including the number of chimeric sequences identified.
dim(seqtab.nochim_MY)
# now: 4335
## calculate the  frequency of chimeric sequences 
sum(seqtab.nochim_MY)/sum(seqtab)
# now 0.877

# Track reads through the pipeline 
# This block of code tracks the number of reads at each stage of the DADA2 pipeline for each sample. This tracking helps monitor data loss, identify problematic samples, and ensure the pipeline is functioning as expected.
getN <- function(x) sum(getUniques(x))

# custom function that calculates the total number of unique reads (amplicon sequences)
track <- cbind(out, sapply(dadaFs, getN), sapply(dadaRs, getN), sapply(mergers, getN), rowSums(seqtab.nochim_MY))
# combines data from multiple steps into a single table
# If processing a single sample, remove the sapply calls: e.g. replace sapply(dadaFs, getN) with getN(dadaFs)
colnames(track) <- c("input", "filtered", "denoisedF", "denoisedR", "merged", "nonchim")
rownames(track) <- sample.names
head(track)

# download table
# before downloading remove the tail of the name
track_adjusted <- track

rownames(track_adjusted) <- rownames(track_adjusted) |>
  sub("_S.*", "", x = _)

# dowload as excel table
write.xlsx(track_adjusted, file = "202404_MY_PLT_stats_mothur_overview_table.xlsx", rowNames = TRUE)

# Purpose
# Quality Control: This table allows you to monitor read retention at each step of the pipeline, ensuring that no unexpected data loss occurs.
# Troubleshooting: If a sample shows a significant drop in reads at a specific step, it may i indicate a problem (e.g., poor quality reads or insufficient overlap for merging).
# Summary: Provides an overview of pipeline performance, enabling comparison across samples.
```

### Assign Taxonomy

```{r}
# Assign taxonomy
# This block of code assigns taxonomic classifications to the non-chimeric sequences (ASVs) in your dataset, enabling the identification of the taxa represented by the sequences

taxa_MY <- assignTaxonomy(seqtab.nochim_MY,"/yourpath/silva_nr_v138_train_set.fa.gz" , multithread=TRUE)
# uses a reference database to classify each unique sequence (ASV) in the seqtab.nochim table to a taxonomic level
# rows: correspond to unique sequences (ASV)

##optional: make species level assignments based on exact matching
taxa_MY <- addSpecies(taxa_MY, "/yourpath/silva_species_assignment_v138.fa.gz")
                   
#inspect the taxonomic assignments:
taxa.print_MY <- taxa_MY # Removing sequence rownames for display only
rownames(taxa.print_MY) <- NULL
head(taxa.print_MY)
```

### Prepare Data for plotting

plots the 20 most abundant bacterial families

```{r}
# add needed info to data frame
samples.out_MY <- rownames(seqtab.nochim_MY)
info_split_MY <- strsplit(samples.out_MY, "_")
Coral_Species <- sapply(info_split_MY, function(x) sub("[0-9]+", "", x[1]))
Genotype <- sapply(info_split_MY, function(x) gsub("[^0-9]", "", x[1]))
Orientation <- sapply(info_split_MY, function(x) sub("[0-9]+", "", x[2]))
Location <- sapply(info_split_MY, function(x) gsub("[^0-9]", "", x[2]))
samdf_MY <- data.frame(Coral_Species, Genotype, Orientation, Location, stringsAsFactors = FALSE)
rownames(samdf_MY) <- samples.out_MY

# check the amount of sampling locations
table(samdf_MY$Location)

# add the name of the Location
samdf_MY <- samdf_MY %>%
  mutate(
    Location_names = case_when(
      Location == "1" ~ "Top",
      Location == "2" ~ "Upper Middle",
      Location == "3" ~ "Lower Middle",
      Location == "4" ~ "Bottom",
      TRUE ~ NA_character_   # just in case other values exist
    )
  )


# add column Condition to meta data
samdf_MY$Condition <- paste(samdf_MY$Location_names, samdf_MY$Orientation, sep = "_")

# save metadata
samdf_MY_download <- samdf_MY

# remove tail from sample name
rownames(samdf_MY_download) <- rownames(samdf_MY_download) |>
  sub("_S.*", "", x = _)

# download
write.xlsx(samdf_MY_download, file = "202404_MY_PLT_metadata.xlsx", rowNames = TRUE)


# Prepare data - FILTERING - seqtab.nochim
# turn matrix
seqtab.nochim_MY_t <- t(seqtab.nochim_MY)

# Filter: Keep only rows with sum > 10
row_sums <- rowSums(seqtab.nochim_MY_t)

seqtab.nochim_MY_filtered_t <- 
  seqtab.nochim_MY_t[row_sums > 10, ]

# before filtering
nrow(seqtab.nochim_MY_t)
# 4335

# after filtering
nrow(seqtab.nochim_MY_filtered_t)
# 3623

# turn matrix back so it was as before
seqtab.nochim_MY_filtered <- t(seqtab.nochim_MY_filtered_t)
# filtered seqtab table

# FILTER: taxa table
# filter taxa table based on the rows from the seqtab.nochim table
filtered_asvs <- rownames(seqtab.nochim_MY_filtered_t)

# subset the taxonomy table to keep only these ASVs
taxa_MY_filtered <- taxa_MY[filtered_asvs, ]

# filter contaminations every kingdom that is not eukaryotes
# first check how many rows are not bacteria
taxa_df <- as.data.frame(taxa_MY_filtered)

# now filter
# Keep only rows where Kingdom is "Bacteria" AND Family is not NA
taxa_MY_filtered_onlybac <- taxa_df[ !is.na(taxa_df$Kingdom) &  taxa_df$Kingdom == "Bacteria", ]

# after filtering
dim(taxa_MY_filtered_onlybac)
# 3453

# make the data frame back into a matrix
# Convert to matrix
taxa_MY_filtered_onlybac <- as.matrix(taxa_MY_filtered_onlybac)

# now remove the rows that were removed from the taxa table also from the asv table before constructing physeq dataframe
# what rows should be kept
keep_ASVs <- rownames(taxa_MY_filtered_onlybac)

# remove filtered rows
seqtab.nochim_MY_filtered_onlybac_t <- seqtab.nochim_MY_filtered_t[keep_ASVs, , drop = FALSE]

# after filtering
nrow(seqtab.nochim_MY_filtered_onlybac_t)
# 3453

# turn matrix
seqtab.nochim_MY_filtered_onlybac <- t(seqtab.nochim_MY_filtered_onlybac_t)

# CONSTRUCT PHYLSEQ
# construct a phyloseq object. with filtered seqtab.nochim and taxa data and ASV names
ps_MY_filtered <- phyloseq(otu_table(seqtab.nochim_MY_filtered_onlybac, taxa_are_rows=FALSE), 
               sample_data(samdf_MY), 
               tax_table(taxa_MY_filtered_onlybac))


# store DNA sequences of our ASV in the refseq slot of the phyloseq object
dna_MY <- Biostrings::DNAStringSet(taxa_names(ps_MY_filtered))
names(dna_MY) <- taxa_names(ps_MY_filtered)
ps_MY_filtered <- merge_phyloseq(ps_MY_filtered, dna_MY)
taxa_names(ps_MY_filtered) <- paste0("ASV", seq(ntaxa(ps_MY_filtered)))
ps_MY_filtered


# FOR DOWNLOADING
# asv table
asv_df_MY <- as.data.frame(otu_table(ps_MY_filtered))

# turn matrix
asv_df_MY <- t(asv_df_MY)

# remove not needed sample name tail
asv_df_MY_download <- asv_df_MY

colnames(asv_df_MY_download) <- colnames(asv_df_MY_download) |>
  sub("_S.*", "", x = _)

# download taxa table with DNA sequences and with ASV names
write.xlsx(asv_df_MY_download, file = "202404_MY_PLT_asv_table_filtered_onlybac.xlsx", rowNames = TRUE)

# extract taxa table with ASV
taxa_df_MY <- as.data.frame(tax_table(ps_MY_filtered))

# download taxa table with DNA sequences and with ASV names
write.xlsx(taxa_df_MY, file = "202404_MY_PLT_taxa_table_asv_filtered_onlybac.xlsx", rowNames = TRUE)


# DNA seq table
# use: seqtab.nochim_t_filtered
# rename sample name: only the important info
seqtab.nochim_MY_filtered_onlybac_t_download <- seqtab.nochim_MY_filtered_onlybac_t

colnames(seqtab.nochim_MY_filtered_onlybac_t_download) <- colnames(seqtab.nochim_MY_filtered_onlybac_t_download) |> sub("_S.*", "", x = _)

# download ASV table with DNA sequences (filtered: >10 count removed and the sampling locations in colony 2)
write.xlsx(seqtab.nochim_MY_filtered_onlybac_t_download, file = "202404_MY_PLT_DNA_seq_filtered_onlybac_table.xlsx", rowNames = TRUE)

# create fasta file
# extract DNA sequences from phyloseq object
asv_seqs_MY <- refseq(ps_MY_filtered)

# check number of sequences
length(asv_seqs_MY)   # should be 3453

# write to FASTA file
Biostrings::writeXStringSet(asv_seqs_MY, "202404_MY_PLT_ASV_DNAseq_filtered_onlybac.fasta")
```

### Indonesia: DADA2

Data processing with [DADA2](https://github.com/benjjneb/dada2) to infer ASVs

```{r}
# read file
cat("Reading files") # prints a message to indicate the script is starting to read files
path <- "." # current working directory
list.files(path) # lists all files in the current directory 

# Forward and reverse fastq filenames have format: 
fnFs <- sort(list.files(path, pattern="_R1_001.fastq", full.names = TRUE))
fnRs <- sort(list.files(path, pattern="_R2_001.fastq", full.names = TRUE))
# searches for files matching specific patterns _R1_001...and sorts files, get a sorted list of file paths for forward (fnFs) and reverse reads (fnRs)

# Extract sample names, assuming filenames have format
sample.names <- sapply(strsplit(basename(fnFs), "__|_0"), function(x) x[2])

#basename(fnFs[1])
# basename(fnFs): Removes the directory path and keeps only the file names. Example: "/path/to/SAMPLE1_R1_001.fastq" → "SAMPLE1_R1_001.fastq"
# strsplit(..., "_"): Splits each filename into parts using the underscore (_) as the delimiter. Example: "SAMPLE1_R1_001.fastq" → c("SAMPLE1", "R1", "001.fastq")
# sapply(..., [, 1): Extracts the first element from each split result (the part before the first underscore). Example: c("SAMPLE1", "R1", "001.fastq") → "SAMPLE1"

cat("Processing",length(sample.names),"samples:", sample.names)
# outputs a summary of the number of samples detected and their names, legth(): counts the number, cat():prints the message

#Inspect read quality profiles
plotQualityProfile(fnFs[1:2])
# forwards reads
# --> Truncate the forward reads at position 280

plotQualityProfile(fnRs[1:2])
# reversed reads
# --> Truncate the forward reads at position 200 where the quality crashes

```

### Filtering and Trimming

This pipeline ensures that only high-quality reads are retained for downstream analysis, which is crucial for the accuracy of biological insights derived from sequencing data.

```{r}
# filter and trimm 
cat("Filtering and trimming")

# prints a message to indicate that the filtering and triming steps are about to begin in the sequencing data processing pipeline

# Place filtered files in filtered/ subdirectory
filtFs <- file.path(path, "filtered", paste0(sample.names, "_F_filt.fastq.gz"))
filtRs <- file.path(path, "filtered", paste0(sample.names, "_R_filt.fastq.gz"))
names(filtFs) <- sample.names
names(filtRs) <- sample.names

### set parameters  maxN=0 (DADA2 requires no Ns), truncQ=2, rm.phix=TRUE and maxEE=2. The maxEE parameter sets the maximum number of “expected errors” allowed in a read, which is a better filter than simply averaging quality scores.
out <- filterAndTrim(fnFs, filtFs, fnRs, filtRs, truncLen=c(280,200),maxN=0, maxEE=c(2,2), truncQ=5, rm.phix=TRUE, compress=TRUE, multithread=TRUE) 

head(out)
```

### Learn error rates & sample interference

```{r}
# learn error rates
# The error model is learned using the filtered and trimmed sequences, which helps in determining the probability of errors at each base position across the reads.Forward (errF) and reverse (errR) errors are learned separately because the forward and reverse reads often have different error profiles.

cat("Learning error rates")
errF <- learnErrors(filtFs, multithread=TRUE)
errR <- learnErrors(filtRs, multithread=TRUE)
# This function in DADA2 is used to learn the error rates for sequencing reads.
# multithread=TRUE: Enables multi-threading to speed up the error rate learning process if the system supports it.

# visualize the esstimated error rates as a sanity check
plotErrors(errF, nominalQ = T)
# everything looks resonable (black line follows the points, the model fits) and we proceed with confidence
```

### Sample Inference

```{r}
# This step is essential for transforming the raw sequencing data into biologically meaningful information by correcting errors and identifying the true sequences present in your samples
# apply the core sample inference algorithm to the filtered and trimmed sequence data.
dadaFs <- dada(filtFs, err=errF, multithread=TRUE, pool = T)
dadaRs <- dada(filtRs, err=errR, multithread=TRUE, pool = T)
# It uses the error rates learned from the reverse reads (errR) to apply the error-correction model and infer the true biological sequences.

#Inspecting the returned dada-class object:
dadaFs[[1]]
dadaRs[[1]]
```

### Merging paired reads and checking on read length

```{r}
mergers_ID <- mergePairs(dadaFs, filtFs, dadaRs, filtRs, verbose=TRUE)
# merges forward (F) and reveres (R) sequences and filtered data in one column
# Inspect the merger data.frame from the first sample
head(mergers_ID[[1]])

# We can now construct an amplicon sequence variant table (ASV) table, a higher-resolution version of the OTU table produced by traditional methods.

seqtab_ID <- makeSequenceTable(mergers_ID)
# creates a sequence table from the merged paired-end reads. A sequence table is a matrix where: rows represent individual samples, columns represent unique DNA sequences, entries contain the counts of each sequence in each sample
dim(seqtab_ID) # 3436
# returns the dimensions of the sequence table (1st value number of rows (sample), 2nd value is the number of columns (unique sequences))

# Inspect distribution of sequence lengths
table(nchar(getSequences(seqtab_ID)))
# looks good
# check visually how the districution of read length is
# Get the read length distribution
length_table_ID <- table(nchar(getSequences(seqtab_ID)))

# Convert to data frame for plotting
length_df_ID <- as.data.frame(length_table_ID)
colnames(length_df_ID) <- c("Length", "Count")

# Make sure Length is numeric
length_df_ID$Length <- as.numeric(as.character(length_df_ID$Length))

ggplot(length_df_ID, aes(x = Length, y = Count)) +
  geom_col(fill = "steelblue") +  # vertical bars
  labs(title = "Distribution of Read Lengths",
       x = "Read Length (bp)",
       y = "Number of Reads") +
    xlim(280, 400)+
  theme_minimal()
```

### Remove Chimeras

```{r}
# Remove chimeras  
# Chimeric sequences are artifacts of the PCR amplification process. These sequences do not represent real biological sequences but rather result from errors in the molecular processes used to prepare and sequence DNA.
# Chimeric sequences can inflate diversity estimates. Removing them ensures that the analysis focuses only on true biological sequences.
# This code removes chimeric sequences from the sequence table, calculates the dimensions of the updated sequence table and determines the proportion of non-chimeric sequences retained.

seqtab.nochim_ID <- removeBimeraDenovo(seqtab_ID, method="consensus", multithread=TRUE, verbose=TRUE)
# removeBim..: identifies and removes chimeric sequences from the sequence table
# method: chimera removal method that identifies chimeras based on agreement across samples
# multithread=T: enables multi-threading for faster computation
# verbose=T: provides detailed output about the chimera removal process including the number of chimeric sequences identified.
dim(seqtab.nochim_ID)
# 678

## calculate the  frequency of chimeric sequences 
sum(seqtab.nochim_ID)/sum(seqtab_ID)
# 0.654 = 65.4% of your reads were retained, only the rest% of reads were discarded as chimeras
# adjusted the trimming from 280 to 250 before but chimera removal was almost the same 0.652 -> changing the cut off did not change the chimera removal 

# Track reads through the pipeline 
# This block of code tracks the number of reads at each stage of the DADA2 pipeline for each sample. This tracking helps monitor data loss, identify problematic samples, and ensure the pipeline is functioning as expected.
getN <- function(x) sum(getUniques(x))

# custom function that calculates the total number of unique reads (amplicon sequences)
track_ID <- cbind(out, sapply(dadaFs, getN), sapply(dadaRs, getN), sapply(mergers_ID, getN), rowSums(seqtab.nochim_ID))
# combines data from multiple steps into a single table
# If processing a single sample, remove the sapply calls: e.g. replace sapply(dadaFs, getN) with getN(dadaFs)
colnames(track_ID) <- c("input", "filtered", "denoisedF", "denoisedR", "merged", "nonchim")
rownames(track_ID) <- sample.names
head(track_ID)

# download table
# before downloading remove the tail of the name
track_ID_adjusted <- track_ID

rownames(track_ID_adjusted) <- rownames(track_ID_adjusted) |>
  sub("_S.*", "", x = _) |>
  sub("^GAD_", "", x = _)

write.xlsx(track_ID_adjusted, file = "202404_ID_PLT_stats_mothur_overview_table.xlsx", rowNames = TRUE)



# Purpose
# Quality Control: This table allows you to monitor read retention at each step of the pipeline, ensuring that no unexpected data loss occurs.
# Troubleshooting: If a sample shows a significant drop in reads at a specific step, it may i indicate a problem (e.g., poor quality reads or insufficient overlap for merging).
# Summary: Provides an overview of pipeline performance, enabling comparison across samples.
```

### Assign Taxonomy

```{r}
# Assign taxonomy
# This block of code assigns taxonomic classifications to the non-chimeric sequences (ASVs) in your dataset, enabling the identification of the taxa represented by the sequences

taxa_ID <- assignTaxonomy(seqtab.nochim_ID,"/yourpathname/silva_nr_v138_train_set.fa.gz" , multithread=TRUE)
# uses a reference database to classify each unique sequence (ASV) in the seqtab.nochim table to a taxonomic level
# rows: correspond to unique sequences (ASV)

##optional: make species level assignments based on exact matching
taxa_ID <- addSpecies(taxa_ID, "/yourpathname/silva_species_assignment_v138.fa.gz")
                   
#inspect the taxonomic assignments:
taxa.print_ID <- taxa_ID # Removing sequence rownames for display only
rownames(taxa.print_ID) <- NULL
head(taxa.print_ID)
```

### Prepare Data for plotting

```{r}
# add needed info to data frame
samples.out_ID <- rownames(seqtab.nochim_ID)
info_split_ID <- strsplit(samples.out_ID, "_")
Coral_Species <- sapply(info_split_ID, function(x) sub("[0-9]+", "", x[2]))
Genotype <- sapply(info_split_ID, function(x) gsub("[^0-9]", "", x[2]))
Location <- sapply(info_split_ID, function(x) x[3])
samdf_ID <- data.frame(Coral_Species, Genotype, Location, stringsAsFactors = FALSE)
rownames(samdf_ID) <- samples.out_ID

# adjust Location 
samdf_ID <- samdf_ID %>%
  mutate(Location = recode(Location,
                           "top" = "Top",
                           "bottom" = "Bottom"))


# save metadata
samdf_ID_download <- samdf_ID

# remove tail from sample name
rownames(samdf_ID_download) <- rownames(samdf_ID_download) |>
  sub("_S.*", "", x = _) |>
  sub("^GAD_", "", x = _)


write.xlsx(samdf_ID_download, file = "202404_ID_GAD_metadata.xlsx", rowNames = TRUE)


# Prepare data - FILTERING - seqtab.nochim
# turn matrix
seqtab.nochim_t_ID <- t(seqtab.nochim_ID)

# Filter: Keep only rows with sum > 10
row_sums_ID <- rowSums(seqtab.nochim_t_ID)

seqtab.nochim_t_ID_filtered <- 
  seqtab.nochim_t_ID[row_sums_ID > 10, ]

# before filtering
nrow(seqtab.nochim_t_ID)
# 678

# after filtering
nrow(seqtab.nochim_t_ID_filtered)
#  535

# turn matrix back so it was as before
seqtab.nochim_ID_filtered <- t(seqtab.nochim_t_ID_filtered)
# filtered seqtab table

# FILTER: taxa table
# filter taxa table based on the rows from the seqtab.nochim table
filtered_asvs_ID <- rownames(seqtab.nochim_t_ID_filtered)

# subset the taxonomy table to keep only these ASVs
taxa_ID_filtered <- taxa_ID[filtered_asvs_ID, ]

# filter contaminations every kingdom that is not eukaryotes
# first check how many rows are not bacteria
taxa_df_ID <- as.data.frame(taxa_ID_filtered)

nrow(taxa_df_ID)
# 535

# now filter
# Keep only rows where Kingdom is "Bacteria"
taxa_ID_filtered_onlybac <- taxa_df_ID[!is.na(taxa_df_ID$Kingdom) & taxa_df_ID$Kingdom == "Bacteria", ]

# Check
dim(taxa_ID_filtered_onlybac)
# 532

# make the data frame back into a matrix
# Convert to matrix
taxa_ID_filtered_onlybac <- as.matrix(taxa_ID_filtered_onlybac)

# now remove the rows that were removed from the taxa table also from the asv table before constructing physeq dataframe
# what rows should be kept
keep_ASVs_ID <- rownames(taxa_ID_filtered_onlybac)

# Assuming your ASV table is named asv_table_MY_filtered
seqtab.nochim_t_ID_filtered_onlybac <- seqtab.nochim_t_ID_filtered[keep_ASVs_ID, , drop = FALSE]

# how many rows before filtering
nrow(seqtab.nochim_t_ID_filtered)
# 535

# how many rows after filtering
nrow(seqtab.nochim_t_ID_filtered_onlybac)
# 532, correct

# turn matrix
seqtab.nochim_ID_filtered_onlybac <- t(seqtab.nochim_t_ID_filtered_onlybac)

# CONSTRUCT PHYLSEQ
# construct a phyloseq object. with filtered seqtab.nochim and taxa data and ASV names
ps_ID_filtered <- phyloseq(otu_table(seqtab.nochim_ID_filtered_onlybac, taxa_are_rows=FALSE), 
               sample_data(samdf_ID), 
               tax_table(taxa_ID_filtered_onlybac))

# its easier to work with the short ASV names than with the full DNA sequence
# store DNA sequences of our ASV in the refseq slot of the phyloseq object
dna_ID <- Biostrings::DNAStringSet(taxa_names(ps_ID_filtered))
names(dna_ID) <- taxa_names(ps_ID_filtered)
ps_ID_filtered <- merge_phyloseq(ps_ID_filtered, dna_ID)
taxa_names(ps_ID_filtered) <- paste0("ASV", seq(ntaxa(ps_ID_filtered)))
ps_ID_filtered


# FOR DOWNLOADING
# asv table
asv_df_ID <- as.data.frame(otu_table(ps_ID_filtered))

# turn matrix
asv_df_ID <- t(asv_df_ID)

# remove not needed sample name tail
asv_df_ID_download <- asv_df_ID

# remove tail from sample name
colnames(asv_df_ID_download) <- colnames(asv_df_ID_download) |>
  sub("_S.*", "", x = _) |>
  sub("^GAD_", "", x = _)

nrow(asv_df_ID_download)
# 532, correct

# download taxa table with DNA sequences and with ASV names
write.xlsx(asv_df_ID_download, file = "202404_ID_GAD_asv_table_filtered_onlybac.xlsx", rowNames = TRUE)

# extract taxa table with ASV
taxa_df_ID <- as.data.frame(tax_table(ps_ID_filtered))
nrow(taxa_df_ID)
# 532, correct

# download taxa table with DNA sequences and with ASV names
write.xlsx(taxa_df_ID, file = "202404_ID_GAD_taxa_table_asv_filtered_onlybac.xlsx", rowNames = TRUE)


# DNA seq table
# use: seqtab.nochim_t_filtered
# rename sample name: only the important info
seqtab.nochim_t_ID_filtered_onlybac_download <- seqtab.nochim_t_ID_filtered_onlybac

nrow(seqtab.nochim_t_ID_filtered_onlybac_download)
# 532, correct

# remove tail from sample name
colnames(seqtab.nochim_t_ID_filtered_onlybac_download) <- colnames(seqtab.nochim_t_ID_filtered_onlybac_download) |>
  sub("_S.*", "", x = _) |>
  sub("^GAD_", "", x = _)

# download ASV table with DNA sequences (filtered: >10 count removed and the sampling locations in colony 2)
write.xlsx(seqtab.nochim_t_ID_filtered_onlybac_download, file = "202404_ID_GAD_DNA_seq_filtered_onlybac_table.xlsx", rowNames = TRUE)

# create fasta file
# extract DNA sequences from phyloseq object
asv_seqs_ID <- refseq(ps_ID_filtered)

# check number of sequences
length(asv_seqs_ID)   # should be 532

# write to FASTA file
Biostrings::writeXStringSet(asv_seqs_ID, "202404_ID_GAD_ASV_DNAseq_filtered_onlybac.fasta")
```

### Barplot: Bacterial Family Malaysia and Indonesia

```{r}
##### MALAYSIA

# prepare data for plotting
# Groups ASVs by Family level, Counts are summed up for each Family
ps_MY_family <- tax_glom(ps_MY_filtered, taxrank = "Family")

# what are the most abundant families in %
ps_MY_family_rel <- transform_sample_counts(ps_MY_family, function(x) x / sum(x) * 100)

# calculate mean abundance per family across all samples
family_abundance_stats <- psmelt(ps_MY_family_rel) %>%
  group_by(Family) %>%
  summarise(
    Mean_Abundance = round(mean(Abundance), 3),
    Variance = round(var(Abundance), 3),
    SD = round(sd(Abundance), 3),
    N = n(),
    .groups = "drop"
  ) %>%
  arrange(desc(Mean_Abundance))

# view most abundant families
head(family_abundance_stats, 10)

# download table
write.xlsx(family_abundance_stats, 
           file = "202404_MY_PLT_family_mean_abundance.xlsx", 
           rowNames = FALSE)

# make dataframe
df_ps_MY_family <- psmelt(ps_MY_family)

# clean sample column name
df_ps_MY_family$Sample <- sub("_S.*", "", df_ps_MY_family$Sample)

# rename families that are NA as unknown
df_ps_MY_family <- df_ps_MY_family %>%
  mutate(Family = ifelse(is.na(Family), "Unknown", Family))

# 3. Normalize abundance to relative abundance per sample
df_ps_MY_family <- df_ps_MY_family %>%
  group_by(Sample) %>%
  mutate(RelAbundance = Abundance / sum(Abundance)) %>%
  ungroup()

# download taxa table with DNA sequences and with ASV names
write.xlsx(df_ps_MY_family, file = "202404_MY_PLT_df_ps_MY_family.xlsx", rowNames = TRUE)

# total number of reads
sum(df_ps_MY_family$Abundance)
# total reads:196.693.7

# how many unique Families do we have?
num_unique_families <- length(unique(df_ps_MY_family$Family))
num_unique_families
# 254 unique families

# top 20
top20_families <- df_ps_MY_family %>%
  group_by(Family) %>%
  summarise(TotalRelAbundance = sum(RelAbundance), .groups = "drop") %>%
  arrange(desc(TotalRelAbundance)) %>%
  slice_head(n = 20) %>%
  pull(Family)

# Group all non-top 20 families as "Other"
df_ps_MY_top20_family <- df_ps_MY_family %>%
  mutate(Family = ifelse(Family %in% top20_families, as.character(Family), "Other"))

# check number of top 20 families
df_ps_MY_top20_family %>% distinct(Family) %>% count()
# 21 (including others)

unique(df_ps_MY_top20_family$Family)

# PLOT
# organize Genotype
df_ps_MY_top20_family <- df_ps_MY_top20_family %>%
  mutate(
    Genotype = factor(Genotype, levels = c("1", "6", "2", "3", "5", "4")))

# Orientation
df_ps_MY_top20_family <- df_ps_MY_top20_family %>%
  mutate(
    Orientation = factor(Orientation, levels = c("W", "E")))

# Location
df_ps_MY_top20_family <- df_ps_MY_top20_family %>%
  mutate(
    Location_names = factor(Location_names, levels = c("Bottom", "Lower Middle", "Upper Middle", "Top")))


# combine ID and MY asign colors to overlapping and unique families
families_combined <- c(
  # Overlapping families
  "Halomonadaceae",
  "Marinobacteraceae",
  "Chlorobiaceae",
  "Idiomarinaceae",
  "Alcanivoracaceae1",
  "Alteromonadaceae",
  "Rhodobacteraceae",
  "Vibrionaceae",
  "Flavobacteriaceae",
  
  # Families only in MY
  "Endozoicomonadaceae",
  "Moraxellaceae",
  "Spirochaetaceae",
  "Marinilabiliaceae",
  "Hungateiclostridiaceae",
  "Propionibacteriaceae",
  "Desulfosarcinaceae",
  "Desulfobacteraceae",
  "Cyclobacteriaceae",
  "Pseudomonadaceae",
  "Clade_I",
  
  # Families only in ID
  "Colwelliaceae",
  "Marinomonadaceae",
  "Peptostreptococcales-Tissierellales_fa",
  "Marinifilaceae",
  "Pseudoalteromonadaceae",
  "Stappiaceae",
  "Nitrincolaceae",
  "Cryomorphaceae",
  "Arcobacteraceae",
  "Kordiimonadaceae",
  "Sphingomonadaceae"
)

colors_combined <- c(
  # Overlapping families — use MY colors
  "Halomonadaceae" = "#00538A",
  "Marinobacteraceae" = "#990000",
  "Chlorobiaceae" = "#FF6800",
  "Idiomarinaceae" = "#FFB300",
  "Alcanivoracaceae1" = "#007D35",
  "Alteromonadaceae" = "#803E75",
  "Rhodobacteraceae" = "#A6BDD7",
  "Vibrionaceae" = "#CC6699",
  "Flavobacteriaceae" = "#93AA00",
  
  # Only in MY
  "Endozoicomonadaceae" = "#53377A",
  "Moraxellaceae" = "#FF0000",
  "Spirochaetaceae" = "#FFEA00",
  "Marinilabiliaceae" = "#CC9999",
  "Hungateiclostridiaceae" = "#99FFFF",
  "Propionibacteriaceae" = "#3399CC",
  "Desulfosarcinaceae" = "#003300",
  "Desulfobacteraceae" = "#593318",
  "Cyclobacteriaceae" = "#66CC66",
  "Pseudomonadaceae" = "#CC3333",
  "Clade_I" = "#CC99FF",
  
  # Only in ID — pick from ID colors if available, else reuse MY
  "Colwelliaceae" = "#006666",   # example ID color
  "Marinomonadaceae" = "#00CC99",
  "Peptostreptococcales-Tissierellales_fa" = "#CCCC66",
  "Marinifilaceae" = "#996633",
  "Pseudoalteromonadaceae" = "#FF6699",
  "Stappiaceae" = "#99CC99",
  "Nitrincolaceae" = "#9999FF",
  "Cryomorphaceae" = "#FF99CC",
  "Arcobacteraceae" = "#0099FF",
  "Kordiimonadaceae" = "#CC00FF",
  "Sphingomonadaceae" = "#0033FF")

# map colors to families
family_colors_combined <- setNames(colors_combined, families_combined)

# add color for "Other"
family_colors_combined["Other"] <- "grey70"

# sort families alphabetically but keep Other first
family_colors_sorted_combined <- c("Other", sort(setdiff(names(family_colors_combined), "Other")))

# reorder color vector
family_colors_combined <- family_colors_combined[family_colors_sorted_combined]

# set Family as factor with Other first
df_ps_MY_top20_family$Family <- factor(
  df_ps_MY_top20_family$Family,
  levels = names(family_colors_combined)  # Other first, top families after
)


# plot
bacteria_family_plot_MY <- ggplot(df_ps_MY_top20_family, 
                                  aes(x = RelAbundance, y = Location_names, fill = Family)) +
  geom_bar(stat = "identity", position = "stack") +
  scale_x_continuous(
    limits = c(0, 1),                      # ensures 0–100%
    labels = scales::percent_format(),     # show as %
    expand = c(0, 0)
  ) +
  facet_grid(Genotype ~ Orientation, scales = "free_y") +
  scale_fill_manual(values = family_colors_combined) +
  theme_minimal(base_size = 13) +
  theme(
    strip.text = element_text(face = "bold", size = 13),
    legend.position = "bottom",
    legend.text = element_text(size = 13),
    legend.title = element_text(size = 13),
    axis.text = element_text(size = 13),
    axis.title = element_text(size = 13),
    panel.grid = element_blank(),
    panel.border = element_blank()
  ) +
  guides(fill = guide_legend(ncol = 3)) +
  labs(
    x = "Relative Abundance",
    y = "Sampling Location",
    fill = "Bacterial Family"
  )

bacteria_family_plot_MY


##### INDONESIA

# prepare data for plotting
# Groups ASVs by Family level, Counts are summed up for each Family
ps_ID_family <- tax_glom(ps_ID_filtered, taxrank = "Family")

# what are the most abundant families in %
ps_ID_family_rel <- transform_sample_counts(ps_ID_family, function(x) x / sum(x) * 100)

# calculate mean abundance per family across all samples
family_abundance_stats_ID <- psmelt(ps_ID_family_rel) %>%
  group_by(Family) %>%
  summarise(
    Mean_Abundance = round(mean(Abundance), 3),
    Variance = round(var(Abundance), 3),
    SD = round(sd(Abundance), 3),
    N = n(),
    .groups = "drop"
  ) %>%
  arrange(desc(Mean_Abundance))

# view most abundant families
head(family_abundance_stats_ID, 10)

# download table
write.xlsx(family_abundance_stats_ID, 
           file = "202404_ID_GAD_family_mean_abundance.xlsx", 
           rowNames = FALSE)

# make dataframe
df_ps_ID_family <- psmelt(ps_ID_family)

# clean sample column name
df_ps_ID_family$Sample <- df_ps_ID_family$Sample |>
  sub("_S.*", "", x = _) |>
  sub("^GAD_", "", x = _)


# rename families that are NA as unknown
df_ps_ID_family <- df_ps_ID_family %>%
  mutate(Family = ifelse(is.na(Family), "Unknown", Family))

# 3. Normalize abundance to relative abundance per sample
df_ps_ID_family <- df_ps_ID_family %>%
  group_by(Sample) %>%
  mutate(RelAbundance = Abundance / sum(Abundance)) %>%
  ungroup()

# download taxa table with DNA sequences and with ASV names
write.xlsx(df_ps_ID_family, file = "202404_ID_GAD_df_ps_ID_family.xlsx", rowNames = TRUE)


# total number of reads
sum(df_ps_ID_family$Abundance)
# 501757
# the Abundance columns is the read counts per ASV per sample.

# how many unique Families do we have?
num_unique_families <- length(unique(df_ps_ID_family$Family))
num_unique_families
# 29 unique families

# top 20
top20_families <- df_ps_ID_family %>%
  group_by(Family) %>%
  summarise(TotalRelAbundance = sum(RelAbundance), .groups = "drop") %>%
  arrange(desc(TotalRelAbundance)) %>%
  slice_head(n = 20) %>%
  pull(Family)

# Group all non-top 20 families as "Other"
df_ps_ID_top20_family <- df_ps_ID_family %>%
  mutate(Family = ifelse(Family %in% top20_families, as.character(Family), "Other"))

# check number of top 20 families
df_ps_ID_top20_family %>% distinct(Family) %>% count()
# 21 (including others)

unique(df_ps_ID_top20_family$Family)

# PLOT
# organize Genotype with increasing depth
df_ps_ID_top20_family <- df_ps_ID_top20_family %>%
  mutate(
    Genotype = factor(Genotype, levels = c("01", "02", "03", "07", "04", "06")))

# Location
df_ps_ID_top20_family <- df_ps_ID_top20_family %>%
  mutate(
    Location = factor(Location, levels = c("Bottom", "Top")))


# adjust color settings for ID data
# set Family as factor with Other first
df_ps_ID_top20_family$Family <- factor(
  df_ps_ID_top20_family$Family,
  levels = names(family_colors_combined)  # Other first, top families after
)


# plot
bacteria_family_plot_ID <- ggplot(df_ps_ID_top20_family, aes(x = RelAbundance, y = Location, fill = Family)) +
  geom_bar(stat = "identity", position = "stack") +
  scale_x_continuous(
    limits = c(0, 1),                    # ensures 0–100% scale
    labels = scales::percent_format(),   # show as 0%, 10%, … 100%
    expand = c(0, 0)                     # remove extra padding
  ) +
  facet_wrap(~Genotype, ncol = 1) +
  scale_fill_manual(values = family_colors_combined) +
  theme_minimal(base_size = 13) +
  theme(
    strip.text = element_text(face = "bold", size = 13),
    legend.position = "bottom",
    legend.text = element_text(size = 13),
    legend.title = element_text(size = 13),
    axis.text = element_text(size = 13),
    axis.title = element_text(size = 13),
    panel.grid = element_blank(),
    panel.border = element_blank()
  ) +
  guides(fill = guide_legend(ncol = 3)) +
  labs(
    x = "Relative Abundance [%]",
    y = "Sampling Location",
    fill = "Bacterial Family"
  )

bacteria_family_plot_ID
```

## **Fig. 4 Bacterial community composition across the surface of massive Porites sp. colonies from Indonesia and Malaysia**

```{r}
# Plot MY and ID together
# Combine plots side-by-side with equal width
final_combined_plot <- bacteria_family_plot_ID + bacteria_family_plot_MY +
  plot_layout(ncol = 2, widths = c(1, 1)) +  # Equal widths for both plots
  plot_annotation(
    theme = theme(
      plot.margin = margin(10, 10, 10, 10)
    )
  )

final_combined_plot

# Figure was downloaded and edited in Affinity designer
ggsave(
  filename = "/your_pathname/Fig4_16S_MY_ID_combined.png",
  plot = final_combined_plot,
  width = 11.7,      # Full A4 width (landscape)
  height = 10,      # One third of A4 height
  dpi = 300,
  units = "in",
  bg = "white"
)

# Save to A4 size (portrait orientation) as PDF
ggsave(
  filename =  "/your_pathname/Fig4_16S_MY_ID_combined.pdf",
  plot = final_combined_plot,
  width = 11.7,      # A4 width in inches
  height = 10,     # A4 height in inches
  units = "in"
)

```
