---
title: "Exploring Intra-colonial Variability in Reef Building Corals"
author: "Marlen Schlotheuber"
format: html
editor: visual
---

# ED50 - Thermal tolerance

### Install needed packages: CBASS package

```{r}
if(!require(devtools)){
   install.packages("devtools")
}

devtools::install_github("reefgenomics/CBASSED50@0.1.5.1", force=TRUE)
```

### Other packages

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(readxl)
library(rstudioapi)
library(RColorBrewer)
library(CBASSED50)
library(tidyverse)
library(patchwork)
library (scales)
library(car)
```

### Load Data

```{r}
# Malaysia data
# upload Malaysia data
input_data_path <- selectFile(
  caption = "Select XLSX or CSV Input File")

# MYS
cbass_dataset_MYS <- read_data(input_data_path)

# To specify the prefix for output files
output_prefix <- tools::file_path_sans_ext(input_data_path)

rlog::log_info(paste("Your current directory is", getwd()))
rlog::log_info(paste("Your input filename is", basename(input_data_path)))
rlog::log_info(paste("The output files will be written into", output_prefix))

# Indonesia
# Upload Indonesia data
input_data_path <- selectFile(
  caption = "Select XLSX or CSV Input File")

# ID
cbass_dataset_ID <- read_data(input_data_path)

# To specify the prefix for output files
output_prefix <- tools::file_path_sans_ext(input_data_path)

rlog::log_info(paste("Your current directory is", getwd()))
rlog::log_info(paste("Your input filename is", basename(input_data_path)))
rlog::log_info(paste("The output files will be written into", output_prefix))
```

## Malaysia and Indonesia

### Analysis and Preprocessing

```{r}
# MYS
cbass_dataset_MYS <- preprocess_dataset(cbass_dataset_MYS)
validate_cbass_dataset(cbass_dataset_MYS)

# ID
cbass_dataset_ID <- preprocess_dataset(cbass_dataset_ID)
validate_cbass_dataset(cbass_dataset_ID)
```

### Explore ED50

```{r}
# MYS
# grouping properties
grouping_properties <- c("Country", "Condition")
# important: keep gouping property: Country, Condition to get the right EDS50 plot !!!!

# PAM data Malaysia
drm_formula <- "PAM ~ Temperature"
models <- fit_drms(cbass_dataset_MYS, grouping_properties, drm_formula, is_curveid = TRUE)

# PAM data
ed50_MYS <- get_ed50_by_grouping_property(models)
cbass_dataset_MYS <- define_grouping_property(cbass_dataset_MYS, grouping_properties) %>%
  mutate(GroupingProperty = paste(GroupingProperty, Genotype, sep = "_"))

ed50_df_MYS <- 
  left_join(ed50_MYS, cbass_dataset_MYS, by = "GroupingProperty") %>%
  dplyr::select(names(ed50_MYS), Depth, Genotype, all_of(grouping_properties)) %>%
  distinct()

# ID
# grouping properties Indonesia
grouping_properties <- c("Country", "Condition")

# PAM data Indonesia
drm_formula <- "PAM ~ Temperature"
models <- fit_drms(cbass_dataset_ID, grouping_properties, drm_formula, is_curveid = TRUE)
str(cbass_dataset_ID)

# PAM data 
ed50_ID <- get_ed50_by_grouping_property(models)
cbass_dataset_ID <- define_grouping_property(cbass_dataset_ID, grouping_properties) %>%
  mutate(GroupingProperty = paste(GroupingProperty, Genotype, sep = "_"))

ed50_df_ID <- 
  left_join(ed50_ID, cbass_dataset_ID, by = "GroupingProperty") %>%
  dplyr::select(names(ed50_ID), Depth, Genotype, all_of(grouping_properties)) %>%
  distinct()

```

### Adjust dataset for plotting

```{r}
# MYS
# add top, middle, bottom to the cbass MYS dataset
cbass_dataset_MYS <- cbass_dataset_MYS %>%
  mutate(Location = case_when(
    Condition %in% c("E1", "W1") ~ "top",    
    Condition %in% c("E2", "W2") ~"upper_middle",
    Condition %in% c("E3", "W3") ~ "lower_middle",
    Condition %in% c("E4", "W4") ~ "bottom", 
    TRUE ~ NA_character_  
  ))

# also add to the ed50_df dataset
ed50_df_MYS <- ed50_df_MYS %>%
  mutate(Location = case_when(
    Condition %in% c("E1", "W1") ~ "top",    
    Condition %in% c("E2", "W2") ~"upper_middle",
    Condition %in% c("E3", "W3") ~ "lower_middle",
    Condition %in% c("E4", "W4") ~ "bottom", 
    TRUE ~ NA_character_  
  ))

# ID
# Create the Location column based on the Condition column
cbass_dataset_ID <- cbass_dataset_ID %>%
  mutate(Location = case_when(
    Condition %in% c("Top") ~ "top",
    Condition %in% c("Bottom") ~ "bottom"))

ed50_df_ID <- ed50_df_ID %>%
  mutate(Location = case_when(
    Condition %in% c("Top") ~ "top",
    Condition %in% c("Bottom") ~ "bottom"))

# MYS
# change genotype numbers to numeric so its the same as for the MYS dataset
str(cbass_dataset_MYS)
cbass_dataset_MYS$Genotype <- as.numeric(cbass_dataset_MYS$Genotype)
cbass_dataset_MYS$Location <- as.factor(cbass_dataset_MYS$Location)
# ID
str(cbass_dataset_ID)
cbass_dataset_ID$Genotype <- as.numeric(cbass_dataset_ID$Genotype)
cbass_dataset_ID$Location <- as.factor(cbass_dataset_ID$Location)
cbass_dataset_ID$Temperature <- as.factor(cbass_dataset_ID$Temperature)
cbass_dataset_ID$PAM <- as.factor(cbass_dataset_ID$PAM)
```

### Temperature response curve

```{r}
# exploratory_curve
# MYS
# set order of the Locations
cbass_dataset_MYS$Location <- factor(cbass_dataset_MYS$Location, 
                                   levels = c("bottom", "lower_middle", "upper_middle", "top"))

# visualize temperature response curve
exploratory_curve <-
  ggplot(data = cbass_dataset_MYS,
       aes(
         x = Temperature,
         y = PAM,
         # You can play around with the group value (e.g., Species, Site, Condition)
         group = GroupingProperty,
         color = Location)) +
  geom_smooth(
    method = drc::drm,
    method.args = list(
      fct = drc::LL.3()),
    se = FALSE,
    size = 0.7
  ) +
  geom_point(size = 1.5) +
  facet_wrap( ~ Genotype, nrow = 2) +
  scale_color_brewer(palette = "Paired") # Colorblind-friendly palette


# ID
# set order of the Locations
cbass_dataset_ID$Location <- factor(cbass_dataset_ID$Location, 
                                   levels = c("bottom", "lower_middle", "upper_middle", "top"))
# visualize temperature response curv 
exploratory_curve <-
  ggplot(data = cbass_dataset_ID,
       aes(
         x = Temperature,
         y = PAM,
         # You can play around with the group value (e.g., Species, Site, Condition)
         group = GroupingProperty,
         color = Location)) +
  geom_smooth(
    method = drc::drm,
    method.args = list(
      fct = drc::LL.3()),
    se = FALSE,
    size = 0.7
  ) +
  geom_point(size = 1.5) +
  facet_wrap( ~ Genotype, nrow = 2) +
  scale_color_brewer(palette = "Paired") # Colorblind-friendly palette


exploratory_curve
```

### Fig. 2 **Thermal tolerance across the surface of coral colonies from Malaysia and Indonesia**

```{r}
# merge datasets
Final_ED50_MYS  <- merge(cbass_dataset_MYS, ed50_df_MYS)

# create the column Orientation
Final_ED50_MYS <- Final_ED50_MYS %>%
  mutate(Orientation = case_when(
    Site %in% c("E") ~ "East",
    Site %in% c("W") ~ "West"))


# set order of the Locations
Final_ED50_MYS$Location <- factor(Final_ED50_MYS$Location, 
                                   levels = c("bottom", "lower_middle", "upper_middle", "top"))
 
# set order of the colonies with increasing depth
Final_ED50_MYS <- Final_ED50_MYS %>%
  mutate(
    Genotype = factor(Genotype, levels = c("1", "6", "2", "3", "5", "4")))

# adjust Orientation column
Final_ED50_MYS <- Final_ED50_MYS %>%
  mutate(
    Orientation = factor(Orientation, levels = c("West", "East")))

# Compute mean ED50 per Genotype
ED50_MYS_means <- Final_ED50_MYS %>%
  group_by(Genotype) %>%
  summarise(mean_ED50 = mean(ED50))

# Merge back to main dataset (on Genotype *and* Orientation)
ED50_MYS_means$Genotype <- as.factor(ED50_MYS_means$Genotype)
str(ED50_MYS_means)

# merge datasets
Final_ED50_MYS <- Final_ED50_MYS %>%
  left_join(ED50_MYS_means, by = c("Genotype")) 


Final_ED50_MYS <- Final_ED50_MYS %>%
  mutate(ED50_diff = ED50 - mean_ED50) %>%  # Compute difference
  select(-contains(".x")) %>%  # Remove duplicate columns (like Orientation.x)
  distinct()  # Re
           
# set the colors    
 genotype_colors <- c("1" = "skyblue3", 
                     "2" = "lightblue", 
                     "3" = "dodgerblue4", 
                     "4" = "cornflowerblue", 
                     "5" = "lightskyblue", 
                     "6" = "dodgerblue3")
 
 

# Bar plot
ED50_diff_barplot_MYS <- ggplot(Final_ED50_MYS, aes(x = ED50_diff, y = Location, fill = Genotype)) +
  geom_vline(aes(xintercept = 0), color = "black", linetype = "dashed", size = 1) +
  geom_vline(xintercept = -0.5, color = "grey40", linetype = "dashed") +
  geom_vline(xintercept = 0.5, color = "grey40", linetype = "dashed") +
  geom_bar(stat = "identity", position = "dodge") +
  facet_grid(Genotype ~ Orientation, scales = "free_y") +
  xlim(-1, 1) +
  scale_y_discrete(labels = c("Bottom", "Lower Middle", "Upper Middle", "Top")) +
  scale_fill_manual(values = genotype_colors) +
  labs(
    x = "Deviation from Mean ED50 [°C]", 
    y = "Location within Colony",
    fill = "Genotype"
  ) +
  theme_minimal(base_size = 13) +  # Global text size
  theme(
    text = element_text(size = 13),
    strip.text = element_blank(),
    axis.title = element_text(size = 13, face = "bold"),  
    axis.text = element_text(size = 13),  
    legend.position = "none",
    panel.spacing = unit(0.4, "lines"),
    strip.background = element_blank()
  )


# Violin plot
ED50_violin_plot_MYS <- ggplot(Final_ED50_MYS, aes(x = 1, y = ED50, fill = Genotype)) +
  geom_violin() +
  facet_wrap(~Genotype, nrow = 6) +
  stat_summary(fun = "mean", geom = "point", color = "black", size = 3, shape = 18, show.legend = FALSE) + 
  stat_summary(fun = "mean", geom = "text", color = "black", size = 4,  # Keep this a little larger for visibility
               aes(label = round(..y.., 2)), vjust = -1) +
  scale_fill_manual(values = genotype_colors) +
  guides(fill = guide_legend(title = "Coral Colony")) +
  labs(
    x = NULL,
    y = "ED50 [°C]",
    fill = "Genotype"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    text = element_text(size = 13),
    strip.text = element_blank(),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 13),
    legend.position = "none",
    panel.spacing = unit(0.4, "lines"),
    strip.background = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  )


# Combine
combined_plot_MYS <- ED50_violin_plot_MYS + ED50_diff_barplot_MYS +
  plot_layout(ncol = 2, widths = c(0.4, 0.8))

combined_plot_MYS

# save as image
ggsave("/Users/MarlenSchlotheuber/Nextcloud/10. projects/11. PhD/10. projects/Intraspecific bleaching project/Project planning/MYS_2024/Fieldwork/Data_26042024/Manuscript/Final_Figures/Fig2_ED50_MYS_within_colonies.png", plot = combined_plot_MYS, width = 12, height = 10, dpi = 300, bg = "white")

# save as pdf
ggsave("/Users/MarlenSchlotheuber/Nextcloud/10. projects/11. PhD/10. projects/Intraspecific bleaching project/Project planning/MYS_2024/Fieldwork/Data_26042024/Manuscript/Final_Figures/Fig2_ED50_MYS_within_colonies.pdf", 
       plot = combined_plot_MYS, 
       width = 12, 
       height = 10, 
       dpi = 300, 
       bg = "white")


#### INDONESIA
# only select needed columns from the indonesia dataset
ed50_df_ID_filtered <- ed50_df_ID %>%
  dplyr::select(c("ED50", "Depth", "Location", "Genotype", "Country"))

# calculate mean ED50 per colony/genotype and calculate differences
ED50_ID_means <- ed50_df_ID_filtered %>%
  group_by(Genotype) %>%
  summarise(mean_ED50 = mean(ED50))

Final_ED50_ID <- ed50_df_ID_filtered %>%
  left_join(ED50_ID_means, by = c("Genotype")) %>%
  mutate(ED50_diff = ED50 - mean_ED50) %>%
  select(-contains(".x")) %>% # remove duplicates
  distinct()  # Re

# change genotype numbre from 01 to 1
Final_ED50_ID$Genotype <- as.numeric(Final_ED50_ID$Genotype)

# prepare for plotting
# set the order of the locations
Final_ED50_ID$Location <- factor(Final_ED50_ID$Location, 
                                   levels = c("bottom", "top"))

# set the order of the coral colonies
Final_ED50_ID <- Final_ED50_ID %>%
  mutate(
    Genotype = factor(Genotype, levels = c("1", "2", "3", "4", "5", "6", "7"))  # Order genotypes
  )

 # set the colors               
genotype_colors <- c("1" = "pink4", 
                     "2" = "orchid4", 
                     "3" = "pink2", 
                     "4" = "mistyrose2", 
                     "5" = "palevioletred3", 
                     "6" = "lightcoral",
                     "7" = "hotpink4")

 
# Remove Genotype 5 from the dataset
Final_ED50_ID_filtered <- Final_ED50_ID[Final_ED50_ID$Genotype != "5", ]


# ED50 Difference Bar Plot
ED50_diff_barplot <- ggplot(Final_ED50_ID_filtered, aes(x = ED50_diff, y = Location, fill = Genotype)) +
  geom_vline(aes(xintercept = 0), color = "black", linetype = "dashed", size = 1) +
  geom_vline(xintercept = -0.5, color = "grey40", linetype = "dashed") +
  geom_vline(xintercept = 0.5, color = "grey40", linetype = "dashed") +
  geom_bar(stat = "identity", position = "dodge") +
  facet_wrap(~Genotype, nrow = 6) +
  xlim(-1, 1) +
  scale_y_discrete(labels = c("Bottom", "Top")) +
  scale_fill_manual(values = genotype_colors) +
  labs(
    x = "Deviation from Mean ED50 [°C]", 
    y = "Location within Colony",
    fill = "Genotype"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    text = element_text(size = 13),
    strip.text = element_blank(),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 13),
    legend.position = "none",
    panel.spacing = unit(0.3, "lines"),
    strip.background = element_blank()
  )


# ED50 Violin Plot
ED50_violin_plot <- ggplot(Final_ED50_ID_filtered, aes(x = 1, y = ED50, fill = Genotype)) +
  geom_violin() +
  facet_wrap(~Genotype, nrow = 7) +
  stat_summary(fun = "mean", geom = "point", color = "black", size = 3, shape = 18, show.legend = FALSE) +
  stat_summary(fun = "mean", geom = "text", color = "black", size = 4,
               aes(label = round(..y.., 2)), vjust = -1) +
  scale_y_continuous(breaks = c(37, 38), limits = c(36.5, 38.5)) +
  scale_fill_manual(values = genotype_colors) +
  labs(
    x = NULL,
    y = "ED50 [°C]",
    fill = "Genotype"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    text = element_text(size = 13),
    strip.text = element_blank(),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 13),
    legend.position = "none",
    panel.spacing = unit(0.3, "lines"),
    strip.background = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  )


# Combine plots
combined_plot_ID <- ED50_violin_plot + ED50_diff_barplot + 
  plot_layout(ncol = 2, widths = c(0.4, 0.8))

# Display combined plot
combined_plot_ID

# save as image
ggsave("/Users/MarlenSchlotheuber/Nextcloud/10. projects/11. PhD/10. projects/Intraspecific bleaching project/Project planning/MYS_2024/Fieldwork/Data_26042024/Manuscript/Final_Figures/Fig3_ED50_ID_within_colonies.png", plot = combined_plot_ID, width = 12, height = 10, dpi = 300, bg = "white")

# save as pdf
ggsave("/Users/MarlenSchlotheuber/Nextcloud/10. projects/11. PhD/10. projects/Intraspecific bleaching project/Project planning/MYS_2024/Fieldwork/Data_26042024/Manuscript/Final_Figures/Fig3_ED50_ID_within_colonies.pdf", 
       plot = combined_plot_ID, 
       width = 12, 
       height = 10, 
       dpi = 300, 
       bg = "white")


# Define each row (violin left, barplot right)
row1 <- ED50_violin_plot_MYS + ED50_diff_barplot_MYS + plot_layout(ncol = 2, widths = c(0.4, 0.8))
row2 <- ED50_violin_plot     + ED50_diff_barplot     + plot_layout(ncol = 2, widths = c(0.4, 0.8))

# Combine both rows vertically
final_combined_plot <- row1 / row2 +
  plot_layout(heights = c(1, 1)) +  # Equal row heights
  plot_annotation(
    theme = theme(
      text = element_text(size = 13),
      plot.margin = margin(10, 10, 10, 10)
    )
  )

# Display
final_combined_plot

# Combine plots into one row
final_combined_plot <-  ED50_violin_plot_MYS + ED50_diff_barplot_MYS +
                        ED50_violin_plot + ED50_diff_barplot +
  plot_layout(ncol = 4, widths = c(0.4, 0.8, 0.4, 0.8)) +
  plot_annotation(
    theme = theme(
      plot.margin = margin(10, 10, 10, 10)
    )
  )

# Display
final_combined_plot

# Save as PNG: A4 landscape width, 1/3 height
ggsave(
  filename = "ITS2_MYS_ID.png",
  plot = final_combined_plot,
  width = 10,      # A4 width (landscape)
  height = 8,     # 1/3 of A4 height
  dpi = 300,
  units = "in",
  bg = "white"
)

# Save as PDF with same dimensions
ggsave(
  filename = "ITS2_MYS_ID.pdf",
  plot = final_combined_plot,
  width = 11.7,
  height = 8,
  units = "in"
)
getwd()
```

## Malaysia

### Testing for signficant differences in thermal tolerance

```{r}
# filter data
Final_ED50_MYS <- Final_ED50_MYS %>% 
  filter(Temperature == 30)

# data distribution
hist(Final_ED50_MYS$ED50) # normally distirbuted
# normallity test
shapiro.test(Final_ED50_MYS$ED50) # p-value above 0.05, data normally distributed but we do not have many data points.
# plot normallity
qqnorm(Final_ED50_MYS$ED50) # looks like its normally distributed

#### ACROSS COLONIES
# check for corrleation of ED50 and Depth, data is normally distributed so use Pearson correlation
cor.test(Final_ED50_MYS$ED50, Final_ED50_MYS$Depth, method = "pearson")
# p-value = 1.323e-05
# The test indicates a statistically significant negative correlation of about -0.584 between the two variables (ED50 and Depth).

# test for significance in ED50 across depth
lm_result <- lm(ED50 ~ Depth, data = Final_ED50_MYS)
summary(lm_result)
# p-value = 1.323e-05

# check for signfiicant differences in ED50 across colonies

# check homogeneity in variance using the levene test
# data is normally distributed now check for homogeneity using the Levene's test to make sure there is equal variance across groups
# If Levene’s test is significant --> shouldn't trust the results of a standard ANOVA.
# Use Welch’s ANOVA, which adjusts for unequal variances.
# Or use a non-parametric alternative like the Kruskal-Wallis test.

leveneTest(ED50 ~ Genotype, data = Final_ED50_MYS)
# p-value = 0.013 --> not signficant I can use the ANOVA

# check for differences across colonies by each location
top_data <- Final_ED50_MYS %>% filter(Location == "top")
um_data <- Final_ED50_MYS %>% filter(Location == "upper_middle")
lm_data <- Final_ED50_MYS %>% filter(Location == "lower_middle")
b_data <- Final_ED50_MYS %>% filter(Location == "bottom")

# check for normality
hist(top_data$ED50)
shapiro.test(top_data$ED50)
qqnorm(top_data$ED50)

leveneTest(ED50 ~ Genotype, data = top_data)
# p-value = < 2.2e-16 *** --> significant can't trust ANOVA use non-parametric test e.g. Kruskal Wallis test (for more than 2 groups)

# non-parametric test
kruskal.test(ED50 ~ Genotype, data = top_data)
# p-value = 0.09024,  not significant

# check for normality
hist(um_data$ED50)
shapiro.test(um_data$ED50)
qqnorm(um_data$ED50) 

# check for homogeneity
leveneTest(ED50 ~ Genotype, data = um_data)
# p-value = 2.2e-16 *** --> significant use non-parametric test

# non-parametric test
kruskal.test(ED50 ~ Genotype, data = um_data)
# p-value = 0.07308

# check for normality
hist(lm_data$ED50)
shapiro.test(lm_data$ED50)
qqnorm(lm_data$ED50)

# check for homogeneity
leveneTest(ED50 ~ Genotype, data = lm_data)
# p-value = 2.2e-16 --> significant  use non-parametric test

# non-parametric test
kruskal.test(ED50 ~ Genotype, data = lm_data)
# p-value = 0.06317

# check for normality
hist(b_data$ED50)
shapiro.test(b_data$ED50)
qqnorm(b_data$ED50)

# check for homogeneity
leveneTest(ED50 ~ Genotype, data = b_data)
# p-value = 2.2e-16 --> significant use non

# non-parametric test
kruskal.test(ED50 ~ Genotype, data = b_data)
# p-value = 0.148

#####
# perform the same testing but using the ED50 deviation
# check for normality
hist(top_data$ED50_diff)
shapiro.test(top_data$ED50)
qqnorm(top_data$ED50)

leveneTest(ED50 ~ Genotype, data = top_data)
# p-value = < 2.2e-16 *** --> significant can't trust ANOVA use non-parametric test e.g. Kruskal Wallis test (for more than 2 groups)

# non-parametric test
kruskal.test(ED50_diff ~ Genotype, data = top_data)
# p-value = 0.1363  not significant

# check for normality
hist(um_data$ED50_diff)
shapiro.test(um_data$ED50_diff)
qqnorm(um_data$ED50_diff) 

# check for homogeneity
leveneTest(ED50_diff ~ Genotype, data = um_data)
# p-value = 2.2e-16 *** --> significant use non-parametric test

# non-parametric test
kruskal.test(ED50_diff ~ Genotype, data = um_data)
# p-value = 0.606

# check for normality
hist(lm_data$ED50_diff)
shapiro.test(lm_data$ED50_diff)
qqnorm(lm_data$ED50_diff)

# check for homogeneity
leveneTest(ED50_diff ~ Genotype, data = lm_data)
# p-value = 2.2e-16 --> significant  use non-parametric test

# non-parametric test
kruskal.test(ED50_diff ~ Genotype, data = lm_data)
# p-value = 0.089

# check for normality
hist(b_data$ED50_diff)
shapiro.test(b_data$ED50_diff)
qqnorm(b_data$ED50_diff)

# check for homogeneity
leveneTest(ED50_diff ~ Genotype, data = b_data)
# p-value = 2.2e-16 --> significant use non

# non-parametric test
kruskal.test(ED50_diff ~ Genotype, data = b_data)
# p-value = 0.434


########
# test within colonies
leveneTest(ED50 ~ Location, data = Final_ED50_MYS)
# p-value = 0.2624 --> not significan can use ANOVA

# test for significance using ANOVA (4 locations within colonies)
anova_result <- aov(ED50 ~ Location, data = Final_ED50_MYS)
summary(anova_result)
# p-value = 0.677

# check for East
Final_ED50_East <- subset(Final_ED50_MYS, Orientation == "East")

# Check for West
Final_ED50_West <- subset(Final_ED50_MYS, Orientation == "West")

# ANOVA for East
# test homogeneity
leveneTest(ED50 ~ Location, data = Final_ED50_East)
# not significant use ANOVA

anova_east <- aov(ED50 ~ Location, data = Final_ED50_East)
summary(anova_east)
# p-value= 0.84, not signf

# ANOVA for West
# test homogeneity
leveneTest(ED50 ~ Location, data = Final_ED50_West)
# p-value = 0.1565, not signficant use ANOVA

anova_west <- aov(ED50 ~ Location, data = Final_ED50_West)
summary(anova_west)


###########
# check for signficant differences in the ED50_Diff

# data distirbution
hist(Final_ED50_MYS$ED50_diff) #  normally distirbuted, or slightly left skewed
# normality
shapiro.test(Final_ED50_MYS$ED50_diff) # P-value above 0.8281, data normally distirbuted
# plot normallity
qqnorm(Final_ED50_MYS$ED50_diff, main = "Q-Q Plot of ED50") # looks like its normally distributed

# test homogeneity
leveneTest(ED50_diff ~ Location, data = Final_ED50_MYS)
# p-value = 0.418, not signficant use ANOVA

# normally distributed data
anova_result <- aov(ED50_diff ~ Location, data = Final_ED50_MYS)
summary(anova_result)
# p-value = 0.251 not significant

# ANOVA for East
anova_east <- aov(ED50_diff ~ Location, data = Final_ED50_East)
summary(anova_east)
# p-value = 0.595

# ANOVA for West
anova_west <- aov(ED50_diff ~ Location, data = Final_ED50_West)
summary(anova_west)
# p-value = 0.567

```

## Indonesia

### Testing for significant differences

```{r}
# check for significant differences
# data distribution
hist(Final_ED50_ID$ED50) # looks like its not normally distributed
shapiro.test(Final_ED50_ID$ED50) # p-value: 0.4483, normasly distributed
qqnorm(Final_ED50_ID$ED50) # normal distributed

# data distribution
hist(Final_ED50_ID$ED50_diff) # looks like its not normally distributed
shapiro.test(Final_ED50_ID$ED50_diff) # p-value: 0.5761, normaly distributed
qqnorm(Final_ED50_ID$ED50_diff) # normal distributed

#####
# across colonies
# can't test for significant differences across genotypes because for each genotpye I only have one datapoint
ggplot(Final_ED50_ID, aes(x = Genotype, y = ED50, color = Location )) +
  geom_point(size = 3)+
  facet_wrap(~Location)

######
# within colonies ED50
# check for differences in variance between the top and bottom sampling location
leveneTest(ED50 ~ Location, data = Final_ED50_ID)
# p-value = 0.9143, not significant difference in the variance of groups, can use ANOVA

# two groups only top and bottom
t.test(ED50 ~ Location, data = Final_ED50_ID)
# p-value = 0.797 not significant 

# within colonies ED50_Diff
leveneTest(ED50_diff ~ Location, data = Final_ED50_ID) # p-value = 1, not significant, can use ANOVA or T.test

# two groups only top and bottom
t.test(ED50_diff ~ Location, data = Final_ED50_ID)
# p-value = 0.579, not significant


```

# ITS2 type profiles - Symbiodiniaceae community composition

## Malaysia

### Load Data and Preprocessing

```{r}
# upload ITS2 type profile meta data
input_data_path <- selectFile(
  caption = "Select XLSX or CSV Input File")

its2_metadata <- read_data(input_data_path)

#upload ITS2 profile absolute abundance
its2_profile <- read.table("your_pathname_/560_20250219T125105_DBV_20250221T224139.profiles.relative.abund_and_meta.txt", header = TRUE, sep = "\t", stringsAsFactors = FALSE)

colnames(its2_profile)[1] <- "ITS2_profile"
colnames(its2_profile)[2] <- "sample_name"

# merge dataframes
its2_data<- merge(its2_metadata, its2_profile, by = "sample_name", all.x = TRUE)

# prepare for plotting
# Convert to long format for ggplot
its2_data_long <- its2_data %>%
  gather(key = "ITS2_profile", value = "value", C15.C15ev.C15dt.C15kl.C15pm:C15.C15by.C15ai)

# just select data with, top middle bottom and middle 1
# Colony 2 was so large that we were able to sample at 6 sampling locations within the colonies (top, upper middle 1, upper middle 2, lower middle 1, lower middle 2, bottom) but since no differences within the colony were detected we removed the two additional middle sampling locations upper and lower middle 2).
its2_data_long_filtered <- its2_data_long %>%
  filter(collection_depthrange %in% c("top", "upper_middle", "upper_middle_1", "lower_middle","lower_middle_1", "bottom"))

# rename upper and lower middle 1 to just lower and upper middle
its2_data_long_filtered <- its2_data_long_filtered %>%
  mutate(collection_depthrange = recode(collection_depthrange,
                                      "lower_middle_1" = "lower_middle",
                                      "upper_middle_1" = "upper_middle"))

# First, ensure your collection_location is a factor with appropriate labels.
its2_data_long_filtered$collection_location <- factor(its2_data_long_filtered$collection_location,
                                              levels = c(4, 3, 2, 1),
                                              labels = c("bottom", "lower_middle", "upper_middle", "top"))

# change the naming in collection_location
# rename upper and lower middle 1 to just lower and upper midlee for colony 2
its2_data_long_filtered <- its2_data_long_filtered %>%
  mutate(collection_location = recode(collection_location,
                                      "lower_middle" = "Lower Middle",
                                      "upper_middle" = "Upper Middle",
                                       "top" = "Top",
                                      "bottom" = "Bottom"))

# Ensure 'collection_depthrange' is ordered and 'collection_orientation' is a factor with "West" and "East"
its2_data_long_filtered <- its2_data_long_filtered %>%
  mutate(
    collection_location = factor(collection_location, levels = c("Bottom", "Lower Middle","Upper Middle", "Top")),
    collection_orientation = factor(collection_orientation, levels = c("West", "East"))
  )
```

### Plot ITS2 Type Profiles

```{r}
# organize data
# set order of the colonies with decreasing depth
its2_data_long_filtered <- its2_data_long_filtered %>%
  mutate(
    collection_colony = factor(collection_colony, levels = c("1", "6", "2", "3", "5", "4")))

#plot
ITS2_plot_MYS <- ggplot(its2_data_long_filtered, aes(x = value, y = collection_location, fill = ITS2_profile)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_x_continuous(labels = scales::percent_format(), expand = c(0, 0)) +
  facet_grid(collection_colony ~ collection_orientation, scales = "free_y") +
  theme_minimal(base_size = 13) +
  guides(fill = guide_legend(ncol = 1)) +
  labs(
    x = "Relative Abundance",
    y = "Location within Colony",
    fill = "ITS2 Type Profiles"
  ) +
  scale_fill_manual(
    values = c(
      "C15.C15ev.C15dt.C15kl.C15pm" = "seagreen",
      "C15.C15ad.C15ai.C116.C116aa" = "darkcyan",
      "C15.C15ev.C15by" = "darkolivegreen3",
      "C15.C15ev.C15dt" = "darkseagreen3",
      "C15.C15ev" = "darkgreen",
      "C15.C15by.C15ai" = "aquamarine3"
    ),
    labels = c(
      "C15.C15ev.C15dt.C15kl.C15pm" = "C15-C15ev-C15dt-C15kl-C15pm",
      "C15.C15ad.C15ai.C116.C116aa" = "C15-C15ad-C15ai-C116-C116aa",
      "C15.C15ev.C15by" = "C15-C15ev-C15by",
      "C15.C15ev.C15dt" = "C15-C15ev-C15dt",
      "C15.C15ev" = "C15-C15ev",
      "C15.C15by.C15ai" = "C15-C15by-C15ai"
    )
  ) +
  theme(
    legend.position = "bottom",
    legend.box = "horizontal",
    legend.text = element_text(size = 13),
    legend.title = element_text(size = 13),
    axis.text = element_text(size = 13),
    axis.title = element_text(size = 13),
    strip.text = element_text(size = 13),
    panel.grid = element_blank(),
    panel.border = element_blank()
  )

ITS2_plot_MYS

# save plot as image
ggsave("/Users/MarlenSchlotheuber/Nextcloud/10. projects/11. PhD/10. projects/Intraspecific bleaching project/Project planning/MYS_2024/Fieldwork/Data_26042024/Manuscript/Final_Figures/FIG4_ITS2_type_profiles_MY.png", plot = ITS2_plot, width = 8, height = 8, dpi = 300, bg = "white")

# save as pdf
ggsave("/Users/MarlenSchlotheuber/Nextcloud/10. projects/11. PhD/10. projects/Intraspecific bleaching project/Project planning/MYS_2024/Fieldwork/Data_26042024/Manuscript/Final_Figures/FIG4_ITS2_type_profiles_MY.pdf", 
       plot = ITS2_plot, 
       width = 8, 
       height = 8, 
       dpi = 300, 
       bg = "white")


```

## Indonesia

### Load Data and Preprocessing

```{r}
# ITS2 type profile metadata
input_data_path <- selectFile(
  caption = "Select XLSX or CSV Input File")

# ITS2 profile data
its2_metadata <- read_data(input_data_path)


#upload ITS2 profile absolute abundance
its2_profile_ID <- read.table('your_pathname/582_20250509T135802_DBV_20250510T135051.profiles.relative.abund_and_meta.txt', header = TRUE, sep = "\t", stringsAsFactors = FALSE)

colnames(its2_profile)[1] <- "ITS2_profile"
colnames(its2_profile_ID)[2] <- "sample_name"

# merge dataframes
its2_data_ID <- merge(its2_metadata, its2_profile, by = "sample_name", all.x = TRUE)

# prepare for plotting
# Convert to long format for ggplot
its2_profile_ID_long <- its2_profile_ID %>%
  gather(key = "ITS2_profile", value = "value", C15.C15aae:C15.C15ad.C15ai.C116.C116aa)

# split colunm into seperate columns to extract the inforamtion
# Ensure sample_name is a character vector
its2_profile_ID_long$sample_name <- as.character(its2_profile_ID_long$sample_name)

# Split the sample_naem column and extract the components
split_info <- do.call(rbind, strsplit(its2_profile_ID_long$sample_name, "_"))

# Assign the components to new columns
its2_profile_ID_long$Coral_species <- split_info[, 1]
its2_profile_ID_long$Genotype <- split_info[, 2]
its2_profile_ID_long$Location <- split_info[, 3]

its2_data_ID_final <- its2_profile_ID_long

# Ensure order
its2_profile_ID_long <- its2_profile_ID_long %>%
  mutate(
    collection_location = factor(Location, levels = c("Bottom", "Top")))
```

### Plot ITS2 Type Profiles

```{r}
ITS2_plot_ID <- ggplot(its2_profile_ID_long, aes(x = value, y = Location, fill = ITS2_profile)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_x_continuous(labels = scales::percent_format(), expand = c(0, 0)) +
  facet_grid(Genotype ~ ., scales = "free_y") +
  theme_minimal(base_size = 13) +
    guides(fill = guide_legend(ncol = 1)) +
  labs(
    x = "Relative Abundance",
    y = "Location within Colony",
    fill = "ITS2 Type Profiles"
  ) +
  scale_fill_manual(
    values = c(
      "C15.C15aae" = "darkgreen",
      "C15.C15bn.C15by" = "mediumaquamarine",
      "C15.C116aa.C15ad" = "forestgreen",
      "C57d.C57a" = "olivedrab2",
      "C15.C15kl.C15he" = "mediumseagreen",
      "C15.C15ad.C15ai.C116.C116aa" = "darkcyan"
    ),
    labels = c(
      "C15.C15aae" = "C15-C15aae",
      "C15.C15bn.C15by" = "C15-C15bn-C15by",
      "C15.C116aa.C15ad" = "C15/C116aa-C15ad",
      "C57d.C57a" = "C57d/C57a",
      "C15.C15kl.C15he" = "C15-C15kl-C15he",
      "C15.C15ad.C15ai.C116.C116aa" = "C15-C15ad-C15ai-C116-C116aa"
    )
  ) +
  theme(
    legend.position = "bottom",
    legend.box = "horizontal",
    legend.text = element_text(size = 13),
    legend.title = element_text(size = 13),
    axis.text = element_text(size = 13),
    axis.title = element_text(size = 13),
    strip.text = element_text(size = 13),
    panel.grid = element_blank(),
    panel.border = element_blank()
  )

ITS2_plot_ID

# save plot as image
ggsave("/Users/MarlenSchlotheuber/Nextcloud/10. projects/11. PhD/10. projects/Intraspecific bleaching project/Project planning/MYS_2024/Fieldwork/Data_26042024/Manuscript/Final_Figures/Fig5_ID_ITS2.png", plot = ITS2_plot_ID, width = 8, height = 8, dpi = 300, bg = "white")

# save as pdf
ggsave("/Users/MarlenSchlotheuber/Nextcloud/10. projects/11. PhD/10. projects/Intraspecific bleaching project/Project planning/MYS_2024/Fieldwork/Data_26042024/Manuscript/Final_Figures/Fig5_ID_ITS2.pdf", 
       plot = ITS2_plot_ID, 
       width = 12, 
       height = 10, 
       dpi = 300, 
       bg = "white")

```

### **Fig. 3 ITS2 type profiles across the surface of coral colonies from Malaysia and Indonesia**

```{r}
# Combine plots side-by-side with equal width
final_combined_plot <- ITS2_plot_MYS + ITS2_plot_ID +
  plot_layout(ncol = 2, widths = c(1, 1)) +  # Equal widths for both plots
  plot_annotation(
    theme = theme(
      plot.margin = margin(10, 10, 10, 10)
    )
  )
final_combined_plot


# Save to A4 width, 1/3 height (landscape)
ggsave(
  filename = "ITS2_MYS_ID.png",
  plot = final_combined_plot,
  width = 11.7,      # Full A4 width (landscape)
  height = 10,      # One third of A4 height
  dpi = 300,
  units = "in",
  bg = "white"
)

# Save to A4 size (portrait orientation) as PDF
ggsave(
  filename = "ITS2_MYS_ID.pdf",
  plot = final_combined_plot,
  width = 11.7,      # A4 width in inches
  height = 10,     # A4 height in inches
  units = "in"
)
```

# 16S - bacterial community composition

The script for 16S rRNA amplicon data analysis of different parts within individual Porites sp. colonies.

## Malaysia

The script for 16S rRNA amplicon data analysis of different parts within individual Porites sp. colonies.

### DADA2

Data processing with [DADA2](https://github.com/benjjneb/dada2) to infer ASVs

```{r}
install.packages("seqinr")
library(seqinr)

# Save as FASTA file
write.fasta(sequences = as.list(raw_fasta),
            names = names(raw_fasta),
            file.out = "top_ASVs.fasta")



# load packages
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("dada2")

BiocManager::install("dada2", force = TRUE)
install.packages(c("cubature", "ape", "reshape2", "edgeR", "plyr"))

install.packages("dada2")
install.packages("Rcpp")
library(dada2)
library(Rcpp)

getwd()
# set working directory
# read file
cat("Reading files") # prints a message to indicate the script is starting to read files
path <- "." # current working directory
list.files(path) # lists all files in the current directory 

# Forward and reverse fastq filenames have format: 
fnFs <- sort(list.files(path, pattern="_R1_001.fastq", full.names = TRUE))
fnRs <- sort(list.files(path, pattern="_R2_001.fastq", full.names = TRUE))
# searches for files matching specific patterns _R1_001...and sorts files, get a sorted list of file paths for forward (fnFs) and reverse reads (fnRs)

# Extract sample names, assuming filenames have format
sample.names <- sapply(strsplit(basename(fnFs), "__|_0"), function(x) x[2])

# basename(fnFs): Removes the directory path and keeps only the file names. Example: "/path/to/SAMPLE1_R1_001.fastq" → "SAMPLE1_R1_001.fastq"
# strsplit(..., "_"): Splits each filename into parts using the underscore (_) as the delimiter. Example: "SAMPLE1_R1_001.fastq" → c("SAMPLE1", "R1", "001.fastq")
# sapply(..., [, 1): Extracts the first element from each split result (the part before the first underscore). Example: c("SAMPLE1", "R1", "001.fastq") → "SAMPLE1"

cat("Processing",length(sample.names),"samples:", sample.names)
# outputs a summary of the number of samples detected and their names, legth(): counts the number, cat():prints the message

#Inspect read quality profiles
plotQualityProfile(fnFs[1:2])
# generates a quality profile plot for the first two forward read files in the fnFs list, DADA2 package in R for processing high-throughput sequence data, to visualize the quality score of sequencing reads across all base positions in a FASTQ file
# selected the first two 1:2 forward read files in fnFs 

# gray-scale: heat map of the frequency of each quality score at each base position
# green line: mean quality score at each position
# orange line: quartiles of the quality score distribution
# red line: scaled proportion of reads that extend to at least that position

# forwards reads
# --> Truncate the forward reads at position 280

plotQualityProfile(fnRs[1:2])

# reversed reads
# worse quality, especially at the end which is common in Illumina sequencing
# DADA2 incorporates quality information into its error model which makes the algorithm robust to lower quality sequence

# reversed reads
# --> Truncate the forward reads at position 200


```

### Filtering and Trimming

This pipeline ensures that only high-quality reads are retained for downstream analysis, which is crucial for the accuracy of biological insights derived from sequencing data.

```{r}
# filter and trimm 
cat("Filtering and trimming")

# prints a message to indicate that the filtering and triming steps are about to begin in the sequencing data processing pipeline

# Place filtered files in filtered/ subdirectory
filtFs <- file.path(path, "filtered", paste0(sample.names, "_F_filt.fastq.gz"))
filtRs <- file.path(path, "filtered", paste0(sample.names, "_R_filt.fastq.gz"))
names(filtFs) <- sample.names
names(filtRs) <- sample.names

### set parameters  maxN=0 (DADA2 requires no Ns), truncQ=2, rm.phix=TRUE and maxEE=2. The maxEE parameter sets the maximum number of “expected errors” allowed in a read, which is a better filter than simply averaging quality scores.
out <- filterAndTrim(fnFs, filtFs, fnRs, filtRs, truncLen=c(280,200),maxN=0, maxEE=c(2,2), truncQ=5, rm.phix=TRUE, compress=TRUE, multithread=TRUE) 
head(out)

```

### Learn error rates & sample interference

```{r}
# learn error rates
# The error model is learned using the filtered and trimmed sequences, which helps in determining the probability of errors at each base position across the reads.Forward (errF) and reverse (errR) errors are learned separately because the forward and reverse reads often have different error profiles.

cat("Learning error rates")
errF <- learnErrors(filtFs, multithread=TRUE)
errR <- learnErrors(filtRs, multithread=TRUE)
# This function in DADA2 is used to learn the error rates for sequencing reads.
# multithread=TRUE: Enables multi-threading to speed up the error rate learning process if the system supports it.

# visualize the esstimated error rates as a sanity check
plotErrors(errF, nominalQ = T)
# everything looks resonable (black line follows the points, the model fits) and we proceed with confidence

```

### Sample Inference

```{r}
# This step is essential for transforming the raw sequencing data into biologically meaningful information by correcting errors and identifying the true sequences present in your samples
# apply the core sample inference algorithm to the filtered and trimmed sequence data.
dadaFs <- dada(filtFs, err=errF, multithread=TRUE, pool = T)
dadaRs <- dada(filtRs, err=errR, multithread=TRUE, pool = T)
# It uses the error rates learned from the reverse reads (errR) to apply the error-correction model and infer the true biological sequences.

#Inspecting the returned dada-class object:
dadaFs[[1]]
dadaRs[[1]]
```

### Merging paired reads and checking on read length

```{r}
mergers <- mergePairs(dadaFs, filtFs, dadaRs, filtRs, verbose=TRUE)
# merges forward (F) and reveres (R) sequences and filtered data in one column
# Inspect the merger data.frame from the first sample
head(mergers[[1]])

# Construct sequencing table
seqtab <- makeSequenceTable(mergers)
# creates a sequence table from the merged paired-end reads. A sequence table is a matrix where: rows represent individual samples, columns represent unique DNA sequences, entries contain the counts of each sequence in each sample
dim(seqtab)
# returns the dimensions of the sequence table (1st value number of rows (sample), 2nd value is the number of columns (unique sequences))

# Inspect distribution of sequence lengths
table(nchar(getSequences(seqtab)))
# looks good
# check visually how the districution of read length is
# Get the read length distribution
length_table <- table(nchar(getSequences(seqtab)))

# Convert to data frame for plotting
length_df <- as.data.frame(length_table)
colnames(length_df) <- c("Length", "Count")

# Make sure Length is numeric
length_df$Length <- as.numeric(as.character(length_df$Length))

ggplot(length_df, aes(x = Length, y = Count)) +
  geom_col(fill = "steelblue") +  # vertical bars
  labs(title = "Distribution of Read Lengths",
       x = "Read Length (bp)",
       y = "Number of Reads") +
  xlim(280, 400)+
  theme_minimal()

```

### Remove Chimeras

```{r}
# Remove chimeras  
# Chimeric sequences are artifacts of the PCR amplification process. These sequences do not represent real biological sequences but rather result from errors in the molecular processes used to prepare and sequence DNA.
# Chimeric sequences can inflate diversity estimates. Removing them ensures that the analysis focuses only on true biological sequences.
# This code removes chimeric sequences from the sequence table, calculates the dimensions of the updated sequence table and determines the proportion of non-chimeric sequences retained.

seqtab.nochim <- removeBimeraDenovo(seqtab, method="consensus", multithread=TRUE, verbose=TRUE)
# removeBim..: identifies and removes chimeric sequences from the sequence table
# method: chimera removal method that identifies chimeras based on agreement across samples
# multithread=T: enables multi-threading for faster computation
# verbose=T: provides detailed output about the chimera removal process including the number of chimeric sequences identified.
dim(seqtab.nochim)

## calculate the  frequency of chimeric sequences 
sum(seqtab.nochim)/sum(seqtab)
# 0.885 = 88.5% of your reads were retained, only 11.5% of reads were discarded as chimeras


# Track reads through the pipeline 
# This block of code tracks the number of reads at each stage of the DADA2 pipeline for each sample. This tracking helps monitor data loss, identify problematic samples, and ensure the pipeline is functioning as expected.
getN <- function(x) sum(getUniques(x))

# custom function that calculates the total number of unique reads (amplicon sequences)
track <- cbind(out, sapply(dadaFs, getN), sapply(dadaRs, getN), sapply(mergers, getN), rowSums(seqtab.nochim))
# combines data from multiple steps into a single table
# If processing a single sample, remove the sapply calls: e.g. replace sapply(dadaFs, getN) with getN(dadaFs)
colnames(track) <- c("input", "filtered", "denoisedF", "denoisedR", "merged", "nonchim")
rownames(track) <- sample.names
head(track)

# Purpose
# Quality Control: This table allows you to monitor read retention at each step of the pipeline, ensuring that no unexpected data loss occurs.
# Troubleshooting: If a sample shows a significant drop in reads at a specific step, it may i indicate a problem (e.g., poor quality reads or insufficient overlap for merging).
# Summary: Provides an overview of pipeline performance, enabling comparison across samples.
```

### Assign Taxonomy

```{r}
# Assign taxonomy
# This block of code assigns taxonomic classifications to the non-chimeric sequences (ASVs) in your dataset, enabling the identification of the taxa represented by the sequences

taxa <- assignTaxonomy(seqtab.nochim,"/Users/MarlenSchlotheuber/Nextcloud/10. projects/11. PhD/10. projects/Intraspecific bleaching project/Project planning/MYS_2024/Fieldwork/Data_26042024/16S_Bacteria_data/16S_files/silva_nr_v138_train_set.fa.gz" , multithread=TRUE)
# uses a reference database to classify each unique sequence (ASV) in the seqtab.nochim table to a taxonomic level
# rows: correspond to unique sequences (ASV)

##optional: make species level assignments based on exact matching
taxa <- addSpecies(taxa, "/Users/MarlenSchlotheuber/Nextcloud/10. projects/11. PhD/10. projects/Intraspecific bleaching project/Project planning/MYS_2024/Fieldwork/Data_26042024/16S_Bacteria_data/16S_files/silva_species_assignment_v138.fa.gz")
                   
#inspect the taxonomic assignments:
taxa.print <- taxa # Removing sequence rownames for display only
rownames(taxa.print) <- NULL
head(taxa.print)
```

### Prepare Data for plotting

plots the 20 most abundant bacterial families

```{r}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

# Install Biostrings
BiocManager::install("Biostrings")
# Install phyloseq
BiocManager::install("phyloseq")

library(phyloseq)
library(Biostrings)
library(reshape2)
library(ggplot2)
library(scales)
library(dplyr)
library(gridExtra)

# add needed info to data frame
samples.out_MYS <- rownames(seqtab.nochim)
info_split_MYS <- strsplit(samples.out_MYS, "_")
Coral_Species <- sapply(info_split_MYS, function(x) sub("[0-9]+", "", x[1]))
Genotype <- sapply(info_split_MYS, function(x) gsub("[^0-9]", "", x[1]))
Orientation <- sapply(info_split_MYS, function(x) sub("[0-9]+", "", x[2]))
Location <- sapply(info_split_MYS, function(x) gsub("[^0-9]", "", x[2]))
samdf_MYS <- data.frame(Coral_Species, Genotype, Orientation, Location, stringsAsFactors = FALSE)
rownames(samdf_MYS) <- samples.out_MYS

# also add the column Location
# Convert Location to numeric for mapping
samdf_MYS$Location <- as.numeric(samdf_MYS$Location)

# Map numeric sampling location to descriptive Location
location_map_MYS <- c("top", "upper middle", "lower middle", "bottom")

samdf_MYS$Location <- location_map_MYS[samdf_MYS$Location]

rownames(samdf_MYS) <- samples.out_MYS

# colony 2 has two more sampling regions but for comparison I will remove it for now
# Remove rows with NA in any column
samdf_MYS <- na.omit(samdf_MYS)

# construct a phyloseq object directly from the dada2 outputs
ps_MYS <- phyloseq(otu_table(seqtab.nochim, taxa_are_rows=FALSE), 
               sample_data(samdf_MYS), 
               tax_table(taxa))

# its easier to work with the short ASV names than with the full DNA sequence
# store DNA sequences of our ASC in the refseq slot of the phyloseq object
dna_MYS <- Biostrings::DNAStringSet(taxa_names(ps_MYS))
names(dna_MYS) <- taxa_names(ps_MYS)
ps_MYS <- merge_phyloseq(ps_MYS, dna_MYS)
taxa_names(ps_MYS) <- paste0("ASV", seq(ntaxa(ps_MYS)))
ps_MYS
```

## Indonesia

### DADA2

Data processing with [DADA2](https://github.com/benjjneb/dada2) to infer ASVs

```{r}
install.packages("seqinr")
library(seqinr)

# Save as FASTA file
write.fasta(sequences = as.list(raw_fasta),
            names = names(raw_fasta),
            file.out = "top_ASVs.fasta")



# load packages
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("dada2")

BiocManager::install("dada2", force = TRUE)
install.packages(c("cubature", "ape", "reshape2", "edgeR", "plyr"))

install.packages("dada2")
install.packages("Rcpp")
library(dada2)
library(Rcpp)

getwd()
# set working directory
# read file
cat("Reading files") # prints a message to indicate the script is starting to read files
path <- "." # current working directory
list.files(path) # lists all files in the current directory 

# Forward and reverse fastq filenames have format: 
fnFs <- sort(list.files(path, pattern="_R1_001.fastq", full.names = TRUE))
fnRs <- sort(list.files(path, pattern="_R2_001.fastq", full.names = TRUE))
# searches for files matching specific patterns _R1_001...and sorts files, get a sorted list of file paths for forward (fnFs) and reverse reads (fnRs)

# Extract sample names, assuming filenames have format
sample.names <- sapply(strsplit(basename(fnFs), "__|_0"), function(x) x[2])

# basename(fnFs): Removes the directory path and keeps only the file names. Example: "/path/to/SAMPLE1_R1_001.fastq" → "SAMPLE1_R1_001.fastq"
# strsplit(..., "_"): Splits each filename into parts using the underscore (_) as the delimiter. Example: "SAMPLE1_R1_001.fastq" → c("SAMPLE1", "R1", "001.fastq")
# sapply(..., [, 1): Extracts the first element from each split result (the part before the first underscore). Example: c("SAMPLE1", "R1", "001.fastq") → "SAMPLE1"

cat("Processing",length(sample.names),"samples:", sample.names)
# outputs a summary of the number of samples detected and their names, legth(): counts the number, cat():prints the message

#Inspect read quality profiles
plotQualityProfile(fnFs[1:2])
# generates a quality profile plot for the first two forward read files in the fnFs list, DADA2 package in R for processing high-throughput sequence data, to visualize the quality score of sequencing reads across all base positions in a FASTQ file
# selected the first two 1:2 forward read files in fnFs 

# gray-scale: heat map of the frequency of each quality score at each base position
# green line: mean quality score at each position
# orange line: quartiles of the quality score distribution
# red line: scaled proportion of reads that extend to at least that position

# forwards reads
# --> Truncate the forward reads at position 280

plotQualityProfile(fnRs[1:2])

# reversed reads
# worse quality, especially at the end which is common in Illumina sequencing
# DADA2 incorporates quality information into its error model which makes the algorithm robust to lower quality sequence

# reversed reads
# --> Truncate the forward reads at position 200


```

### Filtering and Trimming

This pipeline ensures that only high-quality reads are retained for downstream analysis, which is crucial for the accuracy of biological insights derived from sequencing data.

```{r}
# filter and trimm 
cat("Filtering and trimming")

# prints a message to indicate that the filtering and triming steps are about to begin in the sequencing data processing pipeline

# Place filtered files in filtered/ subdirectory
filtFs <- file.path(path, "filtered", paste0(sample.names, "_F_filt.fastq.gz"))
filtRs <- file.path(path, "filtered", paste0(sample.names, "_R_filt.fastq.gz"))
names(filtFs) <- sample.names
names(filtRs) <- sample.names

### set parameters  maxN=0 (DADA2 requires no Ns), truncQ=2, rm.phix=TRUE and maxEE=2. The maxEE parameter sets the maximum number of “expected errors” allowed in a read, which is a better filter than simply averaging quality scores.
out <- filterAndTrim(fnFs, filtFs, fnRs, filtRs, truncLen=c(280,200),maxN=0, maxEE=c(2,2), truncQ=5, rm.phix=TRUE, compress=TRUE, multithread=TRUE) 

head(out)
```

### Learn error rates & sample interference

```{r}
# learn error rates
# The error model is learned using the filtered and trimmed sequences, which helps in determining the probability of errors at each base position across the reads.Forward (errF) and reverse (errR) errors are learned separately because the forward and reverse reads often have different error profiles.

cat("Learning error rates")
errF <- learnErrors(filtFs, multithread=TRUE)
errR <- learnErrors(filtRs, multithread=TRUE)
# This function in DADA2 is used to learn the error rates for sequencing reads.
# multithread=TRUE: Enables multi-threading to speed up the error rate learning process if the system supports it.

# visualize the esstimated error rates as a sanity check
plotErrors(errF, nominalQ = T)
# everything looks resonable (black line follows the points, the model fits) and we proceed with confidence
```

### Sample Inference

```{r}
# This step is essential for transforming the raw sequencing data into biologically meaningful information by correcting errors and identifying the true sequences present in your samples
# apply the core sample inference algorithm to the filtered and trimmed sequence data.
dadaFs <- dada(filtFs, err=errF, multithread=TRUE, pool = T)
dadaRs <- dada(filtRs, err=errR, multithread=TRUE, pool = T)
# It uses the error rates learned from the reverse reads (errR) to apply the error-correction model and infer the true biological sequences.

#Inspecting the returned dada-class object:
dadaFs[[1]]
dadaRs[[1]]
```

### Merging paired reads and checking on read length

```{r}
mergers <- mergePairs(dadaFs, filtFs, dadaRs, filtRs, verbose=TRUE)
# merges forward (F) and reveres (R) sequences and filtered data in one column
# Inspect the merger data.frame from the first sample
head(mergers[[1]])

# Construct sequencing table
seqtab <- makeSequenceTable(mergers)
# creates a sequence table from the merged paired-end reads. A sequence table is a matrix where: rows represent individual samples, columns represent unique DNA sequences, entries contain the counts of each sequence in each sample
dim(seqtab)
# returns the dimensions of the sequence table (1st value number of rows (sample), 2nd value is the number of columns (unique sequences))

# Inspect distribution of sequence lengths
table(nchar(getSequences(seqtab)))
# looks good
# check visually how the districution of read length is
# Get the read length distribution
length_table <- table(nchar(getSequences(seqtab)))

# Convert to data frame for plotting
length_df <- as.data.frame(length_table)
colnames(length_df) <- c("Length", "Count")

# Make sure Length is numeric
length_df$Length <- as.numeric(as.character(length_df$Length))

ggplot(length_df, aes(x = Length, y = Count)) +
  geom_col(fill = "steelblue") +  # vertical bars
  labs(title = "Distribution of Read Lengths",
       x = "Read Length (bp)",
       y = "Number of Reads") +
  xlim(280, 400)+
  theme_minimal()
```

### Remove Chimeras

```{r}
# Remove chimeras  
# Chimeric sequences are artifacts of the PCR amplification process. These sequences do not represent real biological sequences but rather result from errors in the molecular processes used to prepare and sequence DNA.
# Chimeric sequences can inflate diversity estimates. Removing them ensures that the analysis focuses only on true biological sequences.
# This code removes chimeric sequences from the sequence table, calculates the dimensions of the updated sequence table and determines the proportion of non-chimeric sequences retained.

seqtab.nochim <- removeBimeraDenovo(seqtab, method="consensus", multithread=TRUE, verbose=TRUE)
# removeBim..: identifies and removes chimeric sequences from the sequence table
# method: chimera removal method that identifies chimeras based on agreement across samples
# multithread=T: enables multi-threading for faster computation
# verbose=T: provides detailed output about the chimera removal process including the number of chimeric sequences identified.
dim(seqtab.nochim)

## calculate the  frequency of chimeric sequences 
sum(seqtab.nochim)/sum(seqtab)
# 0.654 = 65.4% of your reads were retained, only the rest% of reads were discarded as chimeras


# Track reads through the pipeline 
# This block of code tracks the number of reads at each stage of the DADA2 pipeline for each sample. This tracking helps monitor data loss, identify problematic samples, and ensure the pipeline is functioning as expected.
getN <- function(x) sum(getUniques(x))

# custom function that calculates the total number of unique reads (amplicon sequences)
track <- cbind(out, sapply(dadaFs, getN), sapply(dadaRs, getN), sapply(mergers, getN), rowSums(seqtab.nochim))
# combines data from multiple steps into a single table
# If processing a single sample, remove the sapply calls: e.g. replace sapply(dadaFs, getN) with getN(dadaFs)
colnames(track) <- c("input", "filtered", "denoisedF", "denoisedR", "merged", "nonchim")
rownames(track) <- sample.names
head(track)

# Purpose
# Quality Control: This table allows you to monitor read retention at each step of the pipeline, ensuring that no unexpected data loss occurs.
# Troubleshooting: If a sample shows a significant drop in reads at a specific step, it may i indicate a problem (e.g., poor quality reads or insufficient overlap for merging).
# Summary: Provides an overview of pipeline performance, enabling comparison across samples.
```

### Assign Taxonomy

```{r}
{r}
# Assign taxonomy
# This block of code assigns taxonomic classifications to the non-chimeric sequences (ASVs) in your dataset, enabling the identification of the taxa represented by the sequences

taxa <- assignTaxonomy(seqtab.nochim,"/Users/MarlenSchlotheuber/Nextcloud/10. projects/11. PhD/10. projects/Intraspecific bleaching project/Project planning/MYS_2024/Fieldwork/Data_26042024/16S_Bacteria_data/16S_files/silva_nr_v138_train_set.fa.gz" , multithread=TRUE)
# uses a reference database to classify each unique sequence (ASV) in the seqtab.nochim table to a taxonomic level
# rows: correspond to unique sequences (ASV)

##optional: make species level assignments based on exact matching
# taxa <- addSpecies(taxa, "/Users/MarlenSchlotheuber/Nextcloud/10. projects/11. PhD/10. projects/Intraspecific bleaching project/Project planning/MYS_2024/Fieldwork/Data_26042024/16S_Bacteria_data/16S_files/silva_species_assignment_v138.fa.gz")
                   
#inspect the taxonomic assignments:
taxa.print <- taxa # Removing sequence rownames for display only
rownames(taxa.print) <- NULL
head(taxa.print)
```

### Prepare Data for plotting

```{r}
{r}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

# Install Biostrings
BiocManager::install("Biostrings")
# Install phyloseq
BiocManager::install("phyloseq")

library(phyloseq)
library(Biostrings)
library(reshape2)
library(ggplot2)
library(scales)
library(dplyr)
library(gridExtra)

# add needed info to data frame
samples.out <- rownames(seqtab.nochim)
info_split <- strsplit(samples.out, "_")
Species <- sapply(info_split, function(x) sub("[0-9]+", "", x[2]))
Genotype <- sapply(info_split, function(x) gsub("[^0-9]", "", x[2]))
SamplingLocation <- sapply(info_split, function(x) x[3])
samdf <- data.frame(Species, Genotype, SamplingLocation, stringsAsFactors = FALSE)
rownames(samdf) <- samples.out

# construct a phyloseq object directly from the dada2 outputs
ps <- phyloseq(otu_table(seqtab.nochim, taxa_are_rows=FALSE), 
               sample_data(samdf), 
               tax_table(taxa))

# its easier to work with the short ASV names than with the full DNA sequence
# store DNA sequences of our ASC in the refseq slot of the phyloseq object
dna <- Biostrings::DNAStringSet(taxa_names(ps))
names(dna) <- taxa_names(ps)
ps_ID <- merge_phyloseq(ps, dna)
taxa_names(ps_ID) <- paste0("ASV", seq(ntaxa(ps_ID)))
ps_ID
```

### Barplot: Bacterial Family

```{r}
# Malaysia
# plot it again but with all other bacteria families in grey
# Groups ASVs by Family level, Counts are summed up for each Family
ps_MYS_family <- tax_glom(ps_MYS, taxrank = "Family")

# check what happens if I work with the df
df_ps_MYS_family <- psmelt(ps_MYS_family)

# clean sample column name
df_ps_MYS_family$Sample <- sub("_S.*", "", df_ps_MYS_family$Sample)

# 3. Normalize abundance to relative abundance per sample
df_ps_MYS_family <- df_ps_MYS_family %>%
  group_by(Sample) %>%
  mutate(RelAbundance = Abundance / sum(Abundance)) %>%
  ungroup()

# total number of reads
total_reads_MYS <- df_ps_MYS_family %>%
  summarise(TotalReads = sum(Abundance, na.rm = TRUE))

print(total_reads_MYS)
# total reads: 198.725.1	

# how many ASV do we have?
ntaxa(ps_MYS)
# 4417 MYS
length(unique(df_ps_MYS_family$OTU))
# 274

# how many families do we have?
length(unique(tax_table(ps)[, "Family"]))
#  43

# top 20
top20_families <- df_ps_MYS_family %>%
  group_by(Family) %>%
  summarise(TotalRelAbundance = sum(RelAbundance)) %>%
  arrange(desc(TotalRelAbundance)) %>%
  slice_head(n = 20) %>%
  pull(Family)

# Group all non-top 20 families as "Other"
df_ps_MYS_top20_family <- df_ps_MYS_family %>%
  mutate(Family = ifelse(Family %in% top20_families, as.character(Family), "Other"))

# check number of top 20 families
df_ps_MYS_top20_family %>% distinct(Family) %>% count()
# 21 (including others)
unique(df_ps_MYS_top20_family$Family)

# check on % of each family of the top 20
# Step 1: Sum abundance per family
family_sums <- aggregate(RelAbundance ~ Family, data = df_ps_MYS_top20_family, sum)

# Step 2: Compute total abundance
total_abundance <- sum(family_sums$RelAbundance)

# Step 3: Calculate percent abundance
family_sums$Percent_Abundance <- (family_sums$RelAbundance / total_abundance) * 100




# Set factor levels so "Other" appears last in stack
df_ps_MYS_top20_family <- df_ps_MYS_top20_family %>%
  mutate(Family = factor(Family, levels = c("Other", top20_families)))


# PLOT
# organize Genotype
df_ps_MYS_top20_family <- df_ps_MYS_top20_family %>%
  mutate(
    Genotype = factor(Genotype, levels = c("1", "6", "2", "3", "5", "4")))

# Orientation
df_ps_MYS_top20_family <- df_ps_MYS_top20_family %>%
  mutate(
    Orientation = factor(Orientation, levels = c("W", "E")))

library(tidyverse)
library(dplyr)
library(stringr)
# write the y-axis labesl in capital letters
df_ps_MYS_top20_family <- df_ps_MYS_top20_family %>%
  mutate(Location = str_to_title(Location))

# Location
df_ps_MYS_top20_family <- df_ps_MYS_top20_family %>%
  mutate(
    Location = factor(Location, levels = c("Bottom", "Lower Middle", "Upper Middle", "Top")))


kelly_colors <- c(
  "#00538A", # Strong Blue
  "#C10020", # Vivid Red
  "#FF6800", # Vivid Orange
  "#007D34", # Vivid Green
  "#FFB300", # Vivid Yellow
  "#803E75", # Strong Purple
  "#A6BDD7", # Very Light Blue
  "#CEA262", # Grayish Yellow
  "#817066", # Medium Gray
  "#F6768E", # Strong Purplish Pink
  "#FF7A5C", # Strong Yellowish Pink
  "#53377A", # Strong Violet
  "#FF8E00", # Vivid Orange Yellow
  "#B32851", # Strong Purplish Red
  "#F4C800", # Vivid Greenish Yellow
  "#7F180D", # Strong Reddish Brown
  "#93AA00", # Vivid Yellowish Green
  "#593315", # Deep Yellowish Brown
  "#F13A13", # Vivid Reddish Orange
  "#232C16", # Dark Olive Green
  "#000000", # Black
  "#FFFFFF"  # White (usually not used for fill)
)
# Use only the first 20 colors (excluding black and white if desired)
kelly_20 <- kelly_colors[1:20]

# Map to your families
families <- c(
  "Halomonadaceae",
  "Marinobacteraceae",
  "Chlorobiaceae",
  "Endozoicomonadaceae",
  "Idiomarinaceae",
  "Alcanivoracaceae1",
  "Alteromonadaceae",
  "Moraxellaceae",
  "Spirochaetaceae",
  "Rhodobacteraceae",
  "Marinilabiliaceae",
  "Hungateiclostridiaceae",
  "Vibrionaceae",
  "Cyclobacteriaceae",
  "Flavobacteriaceae",
  "Propionibacteriaceae",
  "Desulfosarcinaceae",
  "Desulfobacteraceae",
  "Clade_I",
  "Pseudomonadaceae"
)

family_colors <- setNames(kelly_20, families)

family_colors <- c(family_colors, Other = "grey70")

df_ps_MYS_top20_family <- df_ps_MYS_top20_family %>%
  mutate(Family = factor(Family, levels = c("Other", top20_families)))

  
bacteria_family_plot_MYS <- ggplot(df_ps_MYS_top20_family, aes(x = RelAbundance, y = Location, fill = Family)) +
  geom_bar(stat = "identity", position = "stack") +
  #scale_x_continuous(labels = scales::percent_format(), expand = c(0, 0)) +
  facet_grid(Genotype ~ Orientation, scales = "free_y") +
  scale_x_continuous() +
  scale_fill_manual(values = family_colors) +
  theme_minimal(base_size = 13) +  # base text size for all elements
  theme(
    strip.text = element_text(face = "bold", size = 13),
    legend.position = "bottom",  # Move legend to bottom
    legend.text = element_text(size = 13),
    legend.title = element_text(size = 13),
    axis.text = element_text(size = 13),
    axis.title = element_text(size = 13),
    panel.grid = element_blank(),
    panel.border = element_blank()
   # legend.box = "horizontal"  # Ensure horizontal layout
  ) +
  guides(fill = guide_legend(ncol =3)) +  # You can adjust number of rows to fit your layout
  labs(
    x = "Relative Abundance",
    y = "Sampling Location",
    fill = "Bacterial Family"
  )


bacteria_family_plot_MYS


# Indonesia

# check overall numbers 
# Groups ASVs by Family level, Counts are summed up for each Family
ps_ID_family <- tax_glom(ps_ID, taxrank = "Family")

# check what happens if I work with the df
df_ps_ID_family <- psmelt(ps_ID_family)

#  Normalize abundance to relative abundance per sample
df_ps_ID_family <- df_ps_ID_family %>%
  group_by(Sample) %>%
  mutate(RelAbundance = Abundance / sum(Abundance)) %>%
  ungroup()

# total number of reads
total_reads <- df_ps_ID_family %>%
  summarise(TotalReads = sum(Abundance, na.rm = TRUE))

print(total_reads)
# total reads:502.311	

# how many ASV do we have?
ntaxa(ps)
# 678

# how many families do we have?
length(unique(tax_table(ps)[, "Family"]))
# 44


# top 20
top20_families <- df_ps_ID_family %>%
  group_by(Family) %>%
  summarise(TotalRelAbundance = sum(RelAbundance)) %>%
  arrange(desc(TotalRelAbundance)) %>%
  slice_head(n = 20) %>%
  pull(Family)

# check top 20 families
top20_families

# Group all non-top 20 families as "Other"
df_ID_top20_family <- df_ps_ID_family %>%
  mutate(Family = ifelse(Family %in% top20_families, Family, "Other"))

# check number of bacteria families
df_ID_top20_family %>%
  distinct(Family) %>%
  count()
# 21 which is correct because I have 20 families and the rest is other

# check on % of each family of the top 20
# Step 1: Sum abundance per family
family_sums <- aggregate(RelAbundance ~ Family, data = df_ID_top20_family, sum)

# Step 2: Compute total abundance
total_abundance <- sum(family_sums$RelAbundance)

# Step 3: Calculate percent abundance
family_sums$Percent_Abundance <- (family_sums$RelAbundance / total_abundance) * 100



# PLOT
# organize Genotype
df_ID_top20_family <- df_ID_top20_family %>%
  mutate(
    Genotype = factor(Genotype, levels = sort(unique(Genotype)))  # Preserve order
  )

df_ID_top20_family <- df_ID_top20_family %>%
  mutate(
    Genotype = factor(Genotype, levels = c("01", "02", "03", "04", "05", "06", "07")))


# Location
df_ID_top20_family <- df_ID_top20_family %>%
  mutate(
    SamplingLocation = factor(SamplingLocation, levels = c("bottom", "top")))

df_ID_top20_family <- df_ID_top20_family %>%
  mutate(
    SamplingLocation = factor(SamplingLocation,
                              levels = c("bottom", "top"),
                              labels = c("Bottom", "Top"))
  )

# Set factor levels so "Other" appears last in stack
df_ID_top20_family <- df_ID_top20_family %>%
  mutate(Family = factor(Family, levels = c("Other", top20_families)))



kelly_colors <- c(
  "#00538A", # Strong Blue
  "#C10020", # Vivid Red
  "#FF6800", # Vivid Orange
  "#007D34", # Vivid Green
  "#FFB300", # Vivid Yellow
  "#803E75", # Strong Purple
  "#A6BDD7", # Very Light Blue
  "#CEA262", # Grayish Yellow
  "#817066", # Medium Gray
  "#F6768E", # Strong Purplish Pink
  "#FF7A5C", # Strong Yellowish Pink
  "#53377A", # Strong Violet
  "#FF8E00", # Vivid Orange Yellow
  "#B32851", # Strong Purplish Red
  "#F4C800", # Vivid Greenish Yellow
  "#7F180D", # Strong Reddish Brown
  "#93AA00", # Vivid Yellowish Green
  "#593315", # Deep Yellowish Brown
  "#F13A13", # Vivid Reddish Orange
  "#232C16", # Dark Olive Green
  "#000000", # Black
  "#FFFFFF"  # White (usually not used for fill)
)

# Use only the first 20 colors (excluding black and white if desired)
kelly_20 <- kelly_colors[1:20]

# check number of top 20 families
df_ID_top20_family %>% distinct(Family) %>% count()

# print names for colors
unique(df_ID_top20_family$Family)

# Map to your families
top20_families <- c(
  "Halomonadaceae",
  "Marinobacteraceae",
  "Chlorobiaceae",
  "Saprospiraceae",
  "Idiomarinaceae",
   "Alcanivoracaceae1",
  "Alteromonadaceae",
  "Flavobacteriaceae",
  "Pseudoalteromonadaceae",
    "Rhodobacteraceae",
  "Colwelliaceae",
  "Marinomonadaceae",
  "Peptostreptococcales-Tissierellales_fa",
      "Vibrionaceae",
  "Arcobacteraceae",
  "Nitrincolaceae",
  "Kordiimonadaceae",
  "Stappiaceae",
  "Cryomorphaceae",
  "Marinifilaceae"
)


family_colors <- setNames(kelly_20, top20_families)

family_colors <- c(family_colors, Other = "grey70")

bacteria_family_plot_ID <- ggplot(df_ID_top20_family, aes(x = RelAbundance, y = SamplingLocation, fill = Family)) +
  geom_bar(stat = "identity", position = "stack") +
  facet_wrap(~Genotype, scales = "free_y", ncol = 1) +
  scale_x_continuous() +
  scale_fill_manual(values = family_colors) +
  theme_minimal(base_size = 13) +
  theme(
    strip.text = element_blank(),  # Remove genotype labels
    legend.position = "bottom",    # Move legend underneath the plot
    legend.text = element_text(size = 13),
    legend.title = element_text(size = 13),
    axis.text = element_text(size = 13),
    axis.title = element_text(size = 13),
    panel.grid = element_blank(),
    legend.box = "horizontal"      # Layout legend horizontally
  ) +
  guides(fill = guide_legend(ncol = 3)) +  # Adjust number of rows as needed
  labs(
    x = "Relative Abundance",
    y = "Sampling Location",
    fill = "Bacterial Family"
  )

bacteria_family_plot_ID

# save as image
ggsave("/Users/MarlenSchlotheuber/Nextcloud/10. projects/11. PhD/10. projects/Intraspecific bleaching project/Project planning/MYS_2024/Fieldwork/Data_26042024/Manuscript/Final_Figures/Fig7_ID_bacteria_family_16S_data.png", plot = bacteria_family_plot, width = 8, height = 8, dpi = 300, bg = "white")

# save as pdf
ggsave("/Users/MarlenSchlotheuber/Nextcloud/10. projects/11. PhD/10. projects/Intraspecific bleaching project/Project planning/MYS_2024/Fieldwork/Data_26042024/Manuscript/Final_Figures/Fig7_ID_bacteria_family_16S_data.pdf", 
       plot = bacteria_family_plot, 
       width = 12, 
       height = 10, 
       dpi = 300, 
       bg = "white")
```

### **Fig. 4 Microbial community composition across the surface of coral colonies from Malaysia and Indonesia**

```{r}
# combine ID and MYS

# Combine plots side-by-side with equal width
final_combined_plot <- bacteria_family_plot_MYS + bacteria_family_plot_ID +
  plot_layout(ncol = 2, widths = c(1, 1)) +  # Equal widths for both plots
  plot_annotation(
    theme = theme(
      plot.margin = margin(10, 10, 10, 10)
    )
  )
final_combined_plot

# Save to A4 width, 1/3 height (landscape)
ggsave(
  filename = "16S_MYS_ID.png",
  plot = final_combined_plot,
  width = 11.7,      # Full A4 width (landscape)
  height = 10,      # One third of A4 height
  dpi = 300,
  units = "in",
  bg = "white"
)

# Save to A4 size (portrait orientation) as PDF
ggsave(
  filename = "16S_MYS_ID.pdf",
  plot = final_combined_plot,
  width = 11.7,      # A4 width in inches
  height = 10,     # A4 height in inches
  units = "in"
)


# Combine plots side-by-side with equal width
final_combined_plot <- ITS2_plot_MYS + ITS2_plot_ID +
  plot_layout(ncol = 2, widths = c(1, 1)) +  # Equal widths for both plots
  plot_annotation(
    theme = theme(
      plot.margin = margin(10, 10, 10, 10)
    )
  )
final_combined_plot


# Save to A4 width, 1/3 height (landscape)
ggsave(
  filename = "ITS2_MYS_ID.png",
  plot = final_combined_plot,
  width = 11.7,      # Full A4 width (landscape)
  height = 10,      # One third of A4 height
  dpi = 300,
  units = "in",
  bg = "white"
)

# Save to A4 size (portrait orientation) as PDF
ggsave(
  filename = "ITS2_MYS_ID.pdf",
  plot = final_combined_plot,
  width = 11.7,      # A4 width in inches
  height = 10,     # A4 height in inches
  units = "in"
)
```
