---
title: "Within-Colony Thermal Tolerance Variability in Reef Building Corals"
author: "Marlen Schlotheuber"
format: html
editor: visual
---

# R code for Manuscript: Within-Colony Thermal Tolerance Variability in Reef-Building Corals

### Install needed packages: CBASS package

```{r}
# load packages
library(tidyverse)
library(seqinr)
library(phyloseq)
library(Biostrings)
library(reshape2)
library(scales)
library(gridExtra)
library(dada2)
library(Rcpp)
library(sf)
library(rnaturalearth)
library(ggmap)
library(readxl)
library(rstudioapi)
library(RColorBrewer)
library(CBASSED50)
library(patchwork)
library(car)
library(mgcv)
library(stringr)
library(lme4)
library(lmerTest) 
library(emmeans)
library(microbiome)
library(ANCOMBC)
library(gsl)
library(vegan)
```

## **Fig.1 Study regions and the sampling locations within massive *P. cf. lobata* colonies**

```{r}
# Study 
# Get country boundaries
world <- ne_countries(scale = "medium", returnclass = "sf")

# Subset for Indonesia and Malaysia
countries <- world[world$name %in% c("Indonesia", "Malaysia"), ]

# plot
Fig1A_map <-  ggplot() +
  geom_sf(data = world, fill = "white", color = "black") +
  geom_sf(data = subset(countries, name == "Indonesia"), fill = "white", color = "black") +
  geom_sf(data = subset(countries, name == "Malaysia"), fill = "white", color = "black") +
  ggtitle("Sampling sites: Indonesia and Malaysia") +
  coord_sf(xlim = c(90, 155), ylim = c(-35, 35), expand = FALSE) +
  scale_x_continuous(name = "Longitude (°E)", breaks = seq(90, 155, 10)) +
  scale_y_continuous(name = "Latitude (°N)", breaks = seq(-35, 35, 10)) +
  theme_minimal() +
  theme(
    panel.background = element_rect(fill = "grey90", color = NA),  # ocean
    plot.background  = element_rect(fill = "white", color = NA),   # outer plot bg
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )

Fig1A_map
```

# Thermal tolerance data from Indonesia and Malaysia

To examine the photosynthetic efficiency data from Malaysia and Indonesia, the R-package from Iakovleva, Colin & Voolstra, 2025 was used ([https://doi.org/10.5281/zenodo.8370644](https://doi.org/10.5281/zenodo.8370644.)).

The photosynthetic efficiency data from the four temperature profiles of each sampling location from Indonesia and Malysia was used to fit log-logistic dose-response curves, using the ‘drc’ R package. These curves were used to determine the effective dose 50 (ED50) for each sampling location (Evensen et al. 2021; Voolstra et al. 2021). The ED50 value defines the temperature at which photosynthetic efficiency has declined by 50% relative to the baseline measurement, and is therefore used as an empirically derived proxy for heat stress tolerance (Evensen et al. 2022). 

### Load Data ED50 data

```{r}
# Indonesia
# Upload Indonesia data: 20240827_ID_GA_Por_CBASS_PAM_data_T420
input_data_path <- selectFile(
  caption = "Select XLSX or CSV Input File")

# ID
cbass_dataset_ID <- read_data(input_data_path)

# To specify the prefix for output files
output_prefix <- tools::file_path_sans_ext(input_data_path)

rlog::log_info(paste("Your current directory is", getwd()))
rlog::log_info(paste("Your input filename is", basename(input_data_path)))
rlog::log_info(paste("The output files will be written into", output_prefix))

# Malaysia data
# upload Malaysia data: 20240418_MY_PLT_Por_CBASS_PAM_data_T420
input_data_path <- selectFile(
  caption = "Select XLSX or CSV Input File")

# MYS
cbass_dataset_MY <- read_data(input_data_path)

# To specify the prefix for output files
output_prefix <- tools::file_path_sans_ext(input_data_path)

rlog::log_info(paste("Your current directory is", getwd()))
rlog::log_info(paste("Your input filename is", basename(input_data_path)))
rlog::log_info(paste("The output files will be written into", output_prefix))

```

### Analysis and Preprocessing

```{r}
# ID
cbass_dataset_ID <- preprocess_dataset(cbass_dataset_ID)
validate_cbass_dataset(cbass_dataset_ID)

# MY
cbass_dataset_MY <- preprocess_dataset(cbass_dataset_MY)
validate_cbass_dataset(cbass_dataset_MY)
```

### Explore ED50

```{r}
# ID
# grouping properties Indonesia
grouping_properties <- c("Condition")

# PAM data Indonesia
drm_formula <- "Pam_value ~ Temperature"
models <- fit_drms(cbass_dataset_ID, grouping_properties, drm_formula, is_curveid = TRUE)
str(cbass_dataset_ID)

# PAM data 
ed50_ID <- get_ed50_by_grouping_property(models)
cbass_dataset_ID <- define_grouping_property(cbass_dataset_ID, grouping_properties) %>%
  mutate(GroupingProperty = paste(GroupingProperty, Genotype, sep = "_"))

ed50_df_ID <- 
  left_join(ed50_ID, cbass_dataset_ID, by = "GroupingProperty") %>%
  dplyr::select(Country, Site, Condition, Species, Genotype, Depth,  all_of(grouping_properties), names(ed50_ID)) %>%
  distinct()


# MY
# grouping properties
grouping_properties <- c("Condition")

# PAM data Malaysia
drm_formula <- "Pam_value ~ Temperature"
models <- fit_drms(cbass_dataset_MY, grouping_properties, drm_formula, is_curveid = TRUE)

# PAM data
ed50_MY <- get_ed50_by_grouping_property(models)
cbass_dataset_MY <- define_grouping_property(cbass_dataset_MY, grouping_properties) %>%
  mutate(GroupingProperty = paste(GroupingProperty, Genotype, sep = "_"))

ed50_df_MY <- 
  left_join(ed50_MY, cbass_dataset_MY, by = "GroupingProperty") %>%
  dplyr::select(Country, Site, Condition, Orientation, Location, Species, Genotype, Depth,  all_of(grouping_properties), names(ed50_MY)) %>%
  distinct()
```

### Temperature response curve

```{r}
# ID
# remove not needed columns
cbass_dataset_ID_filtered <- cbass_dataset_ID %>%
  select(-Date, -Project, -Longitude, -Latitude)


# set order of the Locations
cbass_dataset_ID_filtered$Condition <- factor(cbass_dataset_ID_filtered$Condition, 
                                   levels = c("Top", "Bottom"))

# visualize temperature response curve 
exploratory_curve_ID <-
  ggplot(data = cbass_dataset_ID_filtered,
       aes(
         x = Temperature,
         y = Pam_value,
         # You can play around with the group value (e.g., Species, Site, Condition)
         group = GroupingProperty,
         color = Condition)) +
  geom_smooth(
    method = drc::drm,
    method.args = list(
      fct = drc::LL.3()),
    se = FALSE,
    size = 0.7
  ) +
  geom_point(size = 1.5) +
  facet_wrap( ~ Genotype, nrow = 2) +
  scale_color_brewer(palette = "Paired",   name = "Location")  +
  
  labs(
    x = "Temperature [°C]",
    y = "Photosynthetic efficiency (Fv/Fm)"
  )

#plot
exploratory_curve_ID

# MYS
# set order of the Locations
# remove not needed columns
cbass_dataset_MY_filtered <- cbass_dataset_MY %>%
  select(-Date, -Project, -Longitude, -Latitude)


cbass_dataset_MY_filtered$Location <- factor(cbass_dataset_MY_filtered$Location, 
                                   levels = c("Top", "Upper Middle", "Lower Middle", "Bottom"))

# order genotypes
cbass_dataset_MY_filtered$Genotype <- factor(cbass_dataset_MY_filtered$Genotype, 
                                   levels = c("1", "6", "2", "3", "5", "4"))


# visualize temperature response curve
exploratory_curve_MY <-
  ggplot(data = cbass_dataset_MY_filtered,
         aes(
           x = Temperature,
           y = Pam_value,
           group = GroupingProperty,
           color = Location)) +
  
  geom_smooth(
    method = drc::drm,
    method.args = list(
      fct = drc::LL.3()),
    se = FALSE,
    size = 0.7
  ) +
  
  geom_point(size = 1.5) +
  
  facet_wrap(~ Genotype, nrow = 2) +
  
  scale_color_brewer(palette = "Paired") +
  
  labs(
    x = "Temperature [°C]",
    y = "Photosynthetic efficiency (Fv/Fm)"
  )

# plot
exploratory_curve_MY

```

## **Fig 2. Thermal tolerance (ED50) distribution across the surface of massive *P. cf. lobata* colonies from Indonesia and Malaysia.**

```{r}
# Indonesia
# only select needed columns from the indonesia dataset
ed50_df_ID_filtered <- ed50_df_ID %>%
  dplyr::select(c("ED50", "Depth", "Condition", "Genotype"))

# convert Genotype to numeric
ed50_df_ID_filtered$Genotype <- as.numeric(as.character(ed50_df_ID_filtered$Genotype))

# calculate mean ED50 per colony/genotype and calculate differences
ED50_ID_means <- ed50_df_ID_filtered %>%
  group_by(Genotype) %>%
  summarise(mean_ED50 = mean(ED50))

# create final dataframe with mean ED50
Final_ED50_ID <- ed50_df_ID_filtered %>%
  left_join(ED50_ID_means, by = c("Genotype")) %>%
  mutate(ED50_diff = ED50 - mean_ED50) %>%
  select(-contains(".x")) %>% # remove duplicates
  distinct()  # Re

# prepare for plotting
# set the order of the locations
Final_ED50_ID$Condition <- factor(Final_ED50_ID$Condition, 
                                   levels = c("Bottom", "Top"))

# set the order of the coral colonies: organiye by decreasing depth
Final_ED50_ID <- Final_ED50_ID %>%
  mutate(
    Genotype = factor(Genotype, levels = c("1", "2", "3", "7", "4", "6"))  # Order genotypes
  )

# set the colors               
genotype_colors <- c("1" = "skyblue3", 
                     "2" = "lightblue", 
                     "3" = "dodgerblue4", 
                     "7" = "cornflowerblue", 
                     "4" = "dodgerblue3", 
                     "6" = "lightskyblue")


# ED50 Difference Bar Plot
ED50_diff_barplot_ID <- ggplot(Final_ED50_ID, aes(x = ED50_diff, y = Condition, fill = Genotype)) +
  geom_vline(aes(xintercept = 0), color = "black", linetype = "dashed", size = 1) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_wrap(~Genotype, ncol = 1) +
  xlim(-1, 1) +
  scale_y_discrete(labels = c("Bottom", "Top")) +
  scale_fill_manual(values = genotype_colors) +
  labs(
    x = "Deviation from Mean ED50 [°C]", 
    y = "Location within Colony",
    fill = "Genotype"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    text = element_text(size = 13),
    strip.text = element_blank(),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 13),
    legend.position = "none",
    panel.spacing = unit(0.3, "lines"),
    strip.background = element_blank() )+
      # Add mean ED50 above each facet
  geom_text(
    data = ED50_ID_means,
    aes(
      x = 0,              
      y = 2.5,              
      label = paste0("Mean ED50: ", round(mean_ED50, 2))
    ),
    inherit.aes = FALSE,
    size = 3,
    fontface = "bold" )


ED50_diff_barplot_ID


# Malayisa
# only select needed columns from the indonesia dataset
ed50_df_MY_filtered <- ed50_df_MY %>%
  dplyr::select(c("ED50", "Depth", "Orientation", "Condition", "Genotype", "Location"))

# calculate mean ED50 per colony/genotype and calculate differences
ED50_MY_means <- ed50_df_MY_filtered %>%
  group_by(Genotype) %>%
  summarise(mean_ED50 = round(mean(ED50), 2))

# create final dataframe with mean ED50
Final_ED50_MY <- ed50_df_MY_filtered %>%
  left_join(ED50_MY_means, by = c("Genotype")) %>%
  mutate(ED50_diff = ED50 - mean_ED50, 2) %>%
  select(-contains(".x")) %>% # remove duplicates
  distinct()

# set order of the Locations
Final_ED50_MY$Location <- factor(Final_ED50_MY$Location, 
                                   levels = c("Bottom", "Lower Middle", "Upper Middle", "Top"))
 
# set order of the colonies with decreasing depth
Final_ED50_MY <- Final_ED50_MY %>%
  mutate(
    Genotype = factor(Genotype, levels = c("1", "6", "2", "3", "5", "4")))

# adjust Orientation column
Final_ED50_MY <- Final_ED50_MY %>%
  mutate(
    Orientation = factor(Orientation, levels = c("West", "East")))


# set the colors    
genotype_colors <- c("1" = "pink4", 
                     "2" = "orchid4", 
                     "3" = "pink2", 
                     "4" = "mistyrose2", 
                     "5" = "lightcoral",
                     "6" = "hotpink4")


# Bar plot
ED50_diff_barplot_MY <- ggplot(Final_ED50_MY, aes(x = ED50_diff, y = Location, fill = Genotype)) +
  geom_vline(aes(xintercept = 0), color = "black", linetype = "dashed", size = 1) +
  geom_bar(stat = "identity", position = "dodge") +
  facet_grid(Genotype ~ Orientation, scales = "free_y") +
  xlim(-1, 1) +
  scale_y_discrete(labels = c("Bottom", "Lower Middle", "Upper Middle", "Top")) +
  scale_fill_manual(values = genotype_colors) +
  labs(
    x = "Deviation from Mean ED50 [°C]", 
    y = "Location within Colony",
    fill = "Genotype"
  ) +
  theme_minimal(base_size = 13) +  # Global text size
  theme(
    text = element_text(size = 13),
    strip.text = element_blank(),
    axis.title = element_text(size = 13, face = "bold"),  
    axis.text = element_text(size = 13),  
    legend.position = "none",
    panel.spacing = unit(0.4, "lines"),
    strip.background = element_blank() )+
  geom_text(
    data = ED50_MY_means,
    aes(
      x = 0,              
      y = 4.2,              
      label = paste0("Mean ED50: ", round(mean_ED50, 2))
    ),
    inherit.aes = FALSE,
    size = 3,
    fontface = "bold" )


# plot Malaysia
ED50_diff_barplot_MY

# Final Figure. 2: both plots
ED50_plot <- ED50_diff_barplot_ID + ED50_diff_barplot_MY
ED50_plot
```

# ITS2 type profiles - Symbiodiniaceae community composition

ITS2 sequencing data from colonies collected in Indonesia and Malaysia were seperatly analysed using SymPortal (Hume et al., 2019).\
ITS2 type profiles being representative of putative Symbiodiniaceae genotypes were characterized by specific sets of defining intragenomic ITS2 sequence variants (DIVs) (Hume et al. 2019) and visualized using R studio.

### Indonesia: Load Data and Preprocessing

```{r}
# ITS2 type profile metadata
input_data_path <- selectFile(
  caption = "Select XLSX or CSV Input File")

# ITS2 profile data
its2_metadata_ID <- read_data(input_data_path)


#upload ITS2 profile absolute abundance
its2_profile_ID <- read.table('/Users/MarlenSchlotheuber/Nextcloud/10. projects/11. PhD/10. projects/Intraspecific bleaching project/Project planning/MYS_2024/Fieldwork/Data_26042024/SymPortal/Sequencing_data_ITS2_SymPortal/ITS2_SymPortal_data/202408_ID_ITS2_data/its2_type_profiles/582_20250509T135802_DBV_20250510T135051.profiles.relative.abund_and_meta.txt', header = TRUE, sep = "\t", stringsAsFactors = FALSE)

# remove not needed rows and columns
# Set row 6 as column names
colnames(its2_profile_ID) <- its2_profile_ID[6, ]

# Remove rows 1–5, 19, and 20
its2_profile_ID <- its2_profile_ID[-c(1:6, 19, 20), ]

# Remove first column
its2_profile_ID <- its2_profile_ID[,-1]

# Rename new first column
colnames(its2_profile_ID)[1] <- "sample_name"

# merge dataframes
its2_data_ID <- merge(its2_metadata_ID, its2_profile_ID, by = "sample_name", all.x = TRUE)

# prepare for plotting
# Convert to long format for ggplot
its2_profile_ID_long <- its2_data_ID %>%
  gather(key = "ITS2_profile", value = "value", "C15-C15aae":"C15-C15ad-C15ai-C116-C116aa")

# Ensure order
its2_profile_ID_long <- its2_profile_ID_long %>%
  mutate(
    collection_depthrange = factor(collection_depthrange, levels = c("bottom", "top")))

# set order of the colonies with decreasing depth
its2_profile_ID_long <- its2_profile_ID_long %>%
  mutate(
    collection_colony = factor(collection_colony, levels = c("1", "2", "3", "7", "4", "6")))

# check data structure
str(its2_profile_ID_long)
# value is character, convert to numeric
its2_profile_ID_long$value <- as.numeric(its2_profile_ID_long$value)
```

### Indonesia: Plot ITS2 Type profiles

```{r}
#plot
ITS2_plot_ID <- ggplot(its2_profile_ID_long, aes(x = value, y = collection_depthrange, fill = ITS2_profile)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_x_continuous(labels = scales::percent_format(), expand = c(0, 0)) +
  facet_grid(collection_colony ~ ., scales = "free_y") +
  theme_minimal(base_size = 13) +
    guides(fill = guide_legend(ncol = 1)) +
  labs(
    x = "Relative Abundance [%]",
    y = "Location within Colony",
    fill = "ITS2 Type Profiles"
  ) +
  scale_fill_manual(
    values = c(
      "C15-C15aae" = '#006633',
      "C15-C15bn-C15by" = "#99CC33",
      "C15/C116aa-C15ad" = "#339900",
      "C57d/C57a" = "#000099",
      "C15-C15kl-C15he" = "#009966",
      "C15-C15ad-C15ai-C116-C116aa" = "#6699CC"
    )) +
  theme(
    legend.position = "bottom",
    legend.box = "horizontal",
    legend.text = element_text(size = 13),
    legend.title = element_text(size = 13),
    axis.text = element_text(size = 13),
    axis.title = element_text(size = 13),
    strip.text = element_text(size = 13),
    panel.grid = element_blank(),
    panel.border = element_blank()
  )

ITS2_plot_ID
```

### Malaysia: Load Data and Preprocessing

```{r}
# upload ITS2 type profile meta data
input_data_path <- selectFile(
  caption = "Select XLSX or CSV Input File")

its2_metadata_MY <- read_data(input_data_path)

#upload ITS2 profile absolute abundance
its2_profile_MY <- read.table('/Users/MarlenSchlotheuber/Nextcloud/10. projects/11. PhD/10. projects/Intraspecific bleaching project/Project planning/MYS_2024/Fieldwork/Data_26042024/SymPortal/Sequencing_data_ITS2_SymPortal/ITS2_SymPortal_data/202404_MY_ITS2_data/its2_type_profiles/560_20250219T125105_DBV_20250221T224139.profiles.relative.abund_and_meta.txt', header = TRUE, sep = "\t", stringsAsFactors = FALSE)

# remove not needed rows and columns
# Set row 6 as column names
colnames(its2_profile_MY) <- its2_profile_MY[6, ]

# Remove rows 1–5, 19, and 20
its2_profile_MY <- its2_profile_MY[-c(1:6, 59, 60), ]

# Remove first column
its2_profile_MY <- its2_profile_MY[,-1]

# Rename new first column
colnames(its2_profile_MY)[1] <- "sample_name"

# remove additional sampling locations of colony2 that are not needed
its2_profile_MY <- its2_profile_MY[
  !its2_profile_MY$sample_name %in% c("Por2_E5", "Por2_E6", "Por2_W5", "Por2_W6"),
]

# merge dataframes
its2_data_MY <- merge(its2_metadata_MY, its2_profile_MY, by = "sample_name", all.x = TRUE)

# prepare for plotting
# Convert to long format for ggplot
its2_data_MY_long <- its2_data_MY %>%
  gather(key = "ITS2_profile", value = "value", "C15-C15ev-C15dt-C15kl-C15pm":"C15-C15by-C15ai")

# change the naming in collection_depthrange
its2_data_MY_long <- its2_data_MY_long %>%
  mutate(collection_depthrange = dplyr::recode(collection_depthrange,
                                      lower_middle = "Lower Middle",
                                      upper_middle = "Upper Middle",
                                      top = "Top",
                                      bottom = "Bottom"))

# Ensure 'collection_depthrange' is ordered and 'collection_orientation' is a factor with "West" and "East"
its2_data_MY_long <- its2_data_MY_long %>%
  mutate(
    collection_depthrange = factor(collection_depthrange, levels = c("Bottom", "Lower Middle","Upper Middle", "Top")),
    collection_orientation = factor(collection_orientation, levels = c("West", "East"))
  )

# organize data
# set order of the colonies with decreasing depth
its2_data_MY_long <- its2_data_MY_long %>%
  mutate(
    collection_colony = factor(collection_colony, levels = c("1", "6", "2", "3", "5", "4")))

str(its2_data_MY_long)

#change value to numeric
its2_data_MY_long$value <- as.numeric(its2_data_MY_long$value)
```

### Plot ITS2 Type Profile

```{r}
#plot
ITS2_plot_MY <- ggplot(its2_data_MY_long, aes(x = value, y = collection_depthrange, fill = ITS2_profile)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_x_continuous(labels = scales::percent_format(), expand = c(0, 0)) +
  facet_grid(collection_colony ~ collection_orientation, scales = "free_y") +
  theme_minimal(base_size = 13) +
  guides(fill = guide_legend(ncol = 1)) +
  labs(
    x = "Relative Abundance [%]",
    y = "Location within Colony",
    fill = "ITS2 Type Profiles"
  ) +
  scale_fill_manual(
    values = c(
      "C15-C15ev-C15dt-C15kl-C15pm" = "#66CC66",
         "C15-C15ad-C15ai-C116-C116aa" = "#6699CC",
      "C15-C15ev-C15by" = "#66CC99",
      "C15-C15ev-C15dt" = "#009999",
      "C15-C15ev" = "#99CCCC",
        "C15-C15by-C15ai" = "#006666"
    )) +
  theme(
    legend.position = "bottom",
    legend.box = "horizontal",
    legend.text = element_text(size = 13),
    legend.title = element_text(size = 13),
    axis.text = element_text(size = 13),
    axis.title = element_text(size = 13),
    strip.text = element_text(size = 13),
    panel.grid = element_blank(),
    panel.border = element_blank()
  )

ITS2_plot_MY

```

## **Fig. 3 ITS2 type profiles across the surface of massive *P. cf. lobata* colonies from Indonesia and Malaysia.**

```{r}
# Combine plots side-by-side with equal width
ITS2_final_combined_plot <- ITS2_plot_ID + ITS2_plot_MY +
  plot_layout(ncol = 2, widths = c(1, 1)) +  # Equal widths for both plots
  plot_annotation(
    theme = theme(
      plot.margin = margin(10, 10, 10, 10)
    )
  )

ITS2_final_combined_plot
```

# 16S - bacterial community composition

16S sequencing data was processed in R studio using the DADA2 pipeline. DADA2 (Callahan et al., 2016) filtered and trimmed reads, calculated error rates, inferred sequence variants, merged paired reads, and removed chimeras. Taxonomy was assigned by using the Silva v138 dataset from the Silva reference database (Benjamin, 2024). The top 20 bacterial families across the colony surface were visualised using R studio. <https://benjjneb.github.io/dada2/tutorial.html>

## Indonesia: DADA2

```{r}
cat("Reading files") # prints a message to indicate the script is starting to read files
path <- "." # current working directory
list.files(path) # lists all files in the current directory 

# Forward and reverse fastq filenames have format: 
fnFs <- sort(list.files(path, pattern="_R1_001.fastq", full.names = TRUE))
fnRs <- sort(list.files(path, pattern="_R2_001.fastq", full.names = TRUE))
# searches for files matching specific patterns _R1_001...and sorts files, get a sorted list of file paths for forward (fnFs) and reverse reads (fnRs)

# Extract sample names, assuming filenames have format
sample.names <- sapply(strsplit(basename(fnFs), "__|_0"), function(x) x[2])

cat("Processing",length(sample.names),"samples:", sample.names)
# outputs a summary of the number of samples detected and their names, legth(): counts the number, cat():prints the message

#Inspect read quality profiles
plotQualityProfile(fnFs[1:2])
# forwards reads
# --> Truncate the forward reads at position 280

plotQualityProfile(fnRs[1:2])
# reversed reads
# --> Truncate the forward reads at position 200 where the quality crashes
```

### Filtering and Trimming

This pipeline ensures that only high-quality reads are retained for downstream analysis, which is crucial for the accuracy of biological insights derived from sequencing data.

```{r}
# filter and trimm 
cat("Filtering and trimming")

# prints a message to indicate that the filtering and triming steps are about to begin in the sequencing data processing pipeline

# Place filtered files in filtered/ subdirectory
filtFs <- file.path(path, "filtered", paste0(sample.names, "_F_filt.fastq.gz"))
filtRs <- file.path(path, "filtered", paste0(sample.names, "_R_filt.fastq.gz"))
names(filtFs) <- sample.names
names(filtRs) <- sample.names

### set parameters  maxN=0 (DADA2 requires no Ns), truncQ=2, rm.phix=TRUE and maxEE=2. The maxEE parameter sets the maximum number of “expected errors” allowed in a read, which is a better filter than simply averaging quality scores.
out <- filterAndTrim(fnFs, filtFs, fnRs, filtRs, truncLen=c(280,200),maxN=0, maxEE=c(2,2), truncQ=5, rm.phix=TRUE, compress=TRUE, multithread=TRUE) 

head(out)
```

### Learn error rates & sample interference

```{r}
# learn error rates
# The error model is learned using the filtered and trimmed sequences, which helps in determining the probability of errors at each base position across the reads.Forward (errF) and reverse (errR) errors are learned separately because the forward and reverse reads often have different error profiles.

cat("Learning error rates")
errF <- learnErrors(filtFs, multithread=TRUE)
errR <- learnErrors(filtRs, multithread=TRUE)

# visualize the esstimated error rates as a sanity check
plotErrors(errF, nominalQ = T)
```

### Sample Inference

```{r}
# This step is essential for transforming the raw sequencing data into biologically meaningful information by correcting errors and identifying the true sequences present in your samples
# apply the core sample inference algorithm to the filtered and trimmed sequence data.
dadaFs <- dada(filtFs, err=errF, multithread=TRUE, pool = T)
dadaRs <- dada(filtRs, err=errR, multithread=TRUE, pool = T)
# It uses the error rates learned from the reverse reads (errR) to apply the error-correction model and infer the true biological sequences.

#Inspecting the returned dada-class object:
dadaFs[[1]]
dadaRs[[1]]
```

### Merging paired reads and checking on read length

```{r}
mergers_ID <- mergePairs(dadaFs, filtFs, dadaRs, filtRs, verbose=TRUE)
# merges forward (F) and reveres (R) sequences and filtered data in one column
# Inspect the merger data.frame from the first sample
head(mergers_ID[[1]])

# We can now construct an amplicon sequence variant table (ASV) table, a higher-resolution version of the OTU table produced by traditional methods.

seqtab_ID <- makeSequenceTable(mergers_ID)
# creates a sequence table from the merged paired-end reads. A sequence table is a matrix where: rows represent individual samples, columns represent unique DNA sequences, entries contain the counts of each sequence in each sample
dim(seqtab_ID)
# returns the dimensions of the sequence table (1st value number of rows (sample), 2nd value is the number of columns (unique sequences))

# Inspect distribution of sequence lengths
table(nchar(getSequences(seqtab_ID)))

# check visually how the districution of read length is
# Get the read length distribution
length_table_ID <- table(nchar(getSequences(seqtab_ID)))

# Convert to data frame for plotting
length_df_ID <- as.data.frame(length_table_ID)
colnames(length_df_ID) <- c("Length", "Count")

# Make sure Length is numeric
length_df_ID$Length <- as.numeric(as.character(length_df_ID$Length))

ggplot(length_df_ID, aes(x = Length, y = Count)) +
  geom_col(fill = "steelblue") +  # vertical bars
  labs(title = "Distribution of Read Lengths",
       x = "Read Length (bp)",
       y = "Number of Reads") +
    xlim(280, 400)+
  theme_minimal()
```

### Remove Chimeras

```{r}
# Remove chimeras  
# Chimeric sequences are artifacts of the PCR amplification process. These sequences do not represent real biological sequences but rather result from errors in the molecular processes used to prepare and sequence DNA.
# Chimeric sequences can inflate diversity estimates. Removing them ensures that the analysis focuses only on true biological sequences.
# This code removes chimeric sequences from the sequence table, calculates the dimensions of the updated sequence table and determines the proportion of non-chimeric sequences retained.

seqtab.nochim_ID <- removeBimeraDenovo(seqtab_ID, method="consensus", multithread=TRUE, verbose=TRUE)

dim(seqtab.nochim_ID)

## calculate the  frequency of chimeric sequences 
sum(seqtab.nochim_ID)/sum(seqtab_ID)

# Track reads through the pipeline 
# This block of code tracks the number of reads at each stage of the DADA2 pipeline for each sample. This tracking helps monitor data loss, identify problematic samples, and ensure the pipeline is functioning as expected.
getN <- function(x) sum(getUniques(x))

# custom function that calculates the total number of unique reads (amplicon sequences)
track_ID <- cbind(out, sapply(dadaFs, getN), sapply(dadaRs, getN), sapply(mergers_ID, getN), rowSums(seqtab.nochim_ID))
# combines data from multiple steps into a single table
# If processing a single sample, remove the sapply calls: e.g. replace sapply(dadaFs, getN) with getN(dadaFs)
colnames(track_ID) <- c("input", "filtered", "denoisedF", "denoisedR", "merged", "nonchim")
rownames(track_ID) <- sample.names
head(track_ID)

# download table
# before downloading remove the tail of the name
track_ID_adjusted <- track_ID

rownames(track_ID_adjusted) <- rownames(track_ID_adjusted) |>
  sub("_S.*", "", x = _) |>
  sub("^GAD_", "", x = _)

write.xlsx(track_ID_adjusted, file = "202404_ID_PLT_stats_mothur_overview_table.xlsx", rowNames = TRUE)
```

### Assign Taxonomy

```{r}
# Assign taxonomy
# This block of code assigns taxonomic classifications to the non-chimeric sequences (ASVs) in your dataset, enabling the identification of the taxa represented by the sequences

taxa_ID <- assignTaxonomy(seqtab.nochim_ID,"/yourpathname/silva_nr_v138_train_set.fa.gz" , multithread=TRUE)
# uses a reference database to classify each unique sequence (ASV) in the seqtab.nochim table to a taxonomic level
# rows: correspond to unique sequences (ASV)

##optional: make species level assignments based on exact matching
taxa_ID <- addSpecies(taxa_ID, '/yourpathname/silva_species_assignment_v138.fa.gz')
                   
#inspect the taxonomic assignments:
taxa.print_ID <- taxa_ID # Removing sequence rownames for display only
rownames(taxa.print_ID) <- NULL
head(taxa.print_ID)
```

### Prepare Data for plotting

plots the 20 most abundant bacterial families

```{r}

rownames(samdf_ID_download) <- rownames(samdf_ID_download) |>
  sub("_S.*", "", x = _) |>
  sub("^GAD_", "", x = _)


write.xlsx(samdf_ID_download, file = "202404_ID_GAD_metadata.xlsx", rowNames = TRUE)


# Prepare data - FILTERING - seqtab.nochim
# turn matrix
seqtab.nochim_t_ID <- t(seqtab.nochim_ID)

# Filter: Keep only rows with sum > 10
row_sums_ID <- rowSums(seqtab.nochim_t_ID)

seqtab.nochim_t_ID_filtered <- 
  seqtab.nochim_t_ID[row_sums_ID > 10, ]

# before filtering
nrow(seqtab.nochim_t_ID)
# 678

# after filtering
nrow(seqtab.nochim_t_ID_filtered)
#  535

# turn matrix back so it was as before
seqtab.nochim_ID_filtered <- t(seqtab.nochim_t_ID_filtered)
# filtered seqtab table

# FILTER: taxa table
# filter taxa table based on the rows from the seqtab.nochim table
filtered_asvs_ID <- rownames(seqtab.nochim_t_ID_filtered)

# subset the taxonomy table to keep only these ASVs
taxa_ID_filtered <- taxa_ID[filtered_asvs_ID, ]

# filter contaminations every kingdom that is not eukaryotes
# first check how many rows are not bacteria
taxa_df_ID <- as.data.frame(taxa_ID_filtered)

nrow(taxa_df_ID)

# now filter: oly keep rows where Kingdom is "Bacteria"
taxa_ID_filtered_onlybac <- taxa_df_ID[!is.na(taxa_df_ID$Kingdom) & taxa_df_ID$Kingdom == "Bacteria", ]

# Check
dim(taxa_ID_filtered_onlybac)

# make the data frame back into a matrix
# Convert to matrix
taxa_ID_filtered_onlybac <- as.matrix(taxa_ID_filtered_onlybac)

# now remove the rows that were removed from the taxa table also from the asv table before constructing physeq dataframe
# what rows should be kept
keep_ASVs_ID <- rownames(taxa_ID_filtered_onlybac)

# Assuming your ASV table is named asv_table_MY_filtered
seqtab.nochim_t_ID_filtered_onlybac <- seqtab.nochim_t_ID_filtered[keep_ASVs_ID, , drop = FALSE]

# how many rows before filtering
nrow(seqtab.nochim_t_ID_filtered)

# how many rows after filtering
nrow(seqtab.nochim_t_ID_filtered_onlybac)

# turn matrix
seqtab.nochim_ID_filtered_onlybac <- t(seqtab.nochim_t_ID_filtered_onlybac)

# CONSTRUCT PHYLSEQ
# construct a phyloseq object. with filtered seqtab.nochim and taxa data and ASV names
ps_ID_filtered <- phyloseq(otu_table(seqtab.nochim_ID_filtered_onlybac, taxa_are_rows=FALSE), 
               sample_data(samdf_ID), 
               tax_table(taxa_ID_filtered_onlybac))

# its easier to work with the short ASV names than with the full DNA sequence
# store DNA sequences of our ASV in the refseq slot of the phyloseq object
dna_ID <- Biostrings::DNAStringSet(taxa_names(ps_ID_filtered))
names(dna_ID) <- taxa_names(ps_ID_filtered)
ps_ID_filtered <- merge_phyloseq(ps_ID_filtered, dna_ID)
taxa_names(ps_ID_filtered) <- paste0("ASV", seq(ntaxa(ps_ID_filtered)))
ps_ID_filtered


# FOR DOWNLOADING
# asv table
asv_df_ID <- as.data.frame(otu_table(ps_ID_filtered))

# turn matrix
asv_df_ID <- t(asv_df_ID)

# remove not needed sample name tail
asv_df_ID_download <- asv_df_ID

# remove tail from sample name
colnames(asv_df_ID_download) <- colnames(asv_df_ID_download) |>
  sub("_S.*", "", x = _) |>
  sub("^GAD_", "", x = _)

nrow(asv_df_ID_download)

# download taxa table with DNA sequences and with ASV names
write.xlsx(asv_df_ID_download, file = "202404_ID_GAD_asv_table_filtered_onlybac.xlsx", rowNames = TRUE)

# extract taxa table with ASV
taxa_df_ID <- as.data.frame(tax_table(ps_ID_filtered))
nrow(taxa_df_ID)

# download taxa table with DNA sequences and with ASV names
write.xlsx(taxa_df_ID, file = "202404_ID_GAD_taxa_table_asv_filtered_onlybac.xlsx", rowNames = TRUE)


# DNA seq table
# use: seqtab.nochim_t_filtered
# rename sample name: only the important info
seqtab.nochim_t_ID_filtered_onlybac_download <- seqtab.nochim_t_ID_filtered_onlybac

nrow(seqtab.nochim_t_ID_filtered_onlybac_download)

# remove tail from sample name
colnames(seqtab.nochim_t_ID_filtered_onlybac_download) <- colnames(seqtab.nochim_t_ID_filtered_onlybac_download) |>
  sub("_S.*", "", x = _) |>
  sub("^GAD_", "", x = _)

# download ASV table with DNA sequences (filtered: >10 count removed and the sampling locations in colony 2)
write.xlsx(seqtab.nochim_t_ID_filtered_onlybac_download, file = "202404_ID_GAD_DNA_seq_filtered_onlybac_table.xlsx", rowNames = TRUE)

# create fasta file
# extract DNA sequences from phyloseq object
asv_seqs_ID <- refseq(ps_ID_filtered)

# check number of sequences
length(asv_seqs_ID)   # should be 532

# write to FASTA file
Biostrings::writeXStringSet(asv_seqs_ID, "202404_ID_GAD_ASV_DNAseq_filtered_onlybac.fasta")
```

### Barplot: Bacterial Family top 20

```{r}
# prepare data for plotting
# Groups ASVs by Family level, Counts are summed up for each Family
ps_ID_family <- tax_glom(ps_ID_filtered, taxrank = "Family")

# what are the most abundant families in %
ps_ID_family_rel <- transform_sample_counts(ps_ID_family, function(x) x / sum(x) * 100)

# calculate mean abundance per family across all samples
family_abundance_stats_ID <- psmelt(ps_ID_family_rel) %>%
  group_by(Family) %>%
  summarise(
    Mean_Abundance = round(mean(Abundance), 3),
    Variance = round(var(Abundance), 3),
    SD = round(sd(Abundance), 3),
    N = n(),
    .groups = "drop"
  ) %>%
  arrange(desc(Mean_Abundance))

# view most abundant families
head(family_abundance_stats_ID, 10)

# download table
write.xlsx(family_abundance_stats_ID, 
           file = "202404_ID_GAD_family_mean_abundance.xlsx", 
           rowNames = FALSE)

# make dataframe
df_ps_ID_family <- psmelt(ps_ID_family)

# clean sample column name
df_ps_ID_family$Sample <- df_ps_ID_family$Sample |>
  sub("_S.*", "", x = _) |>
  sub("^GAD_", "", x = _)


# rename families that are NA as unknown
df_ps_ID_family <- df_ps_ID_family %>%
  mutate(Family = ifelse(is.na(Family), "Unknown", Family))

# 3. Normalize abundance to relative abundance per sample
df_ps_ID_family <- df_ps_ID_family %>%
  group_by(Sample) %>%
  mutate(RelAbundance = Abundance / sum(Abundance)) %>%
  ungroup()

# download taxa table with DNA sequences and with ASV names
write.xlsx(df_ps_ID_family, file = "202404_ID_GAD_df_ps_ID_family.xlsx", rowNames = TRUE)


# total number of reads
sum(df_ps_ID_family$Abundance)

# how many unique Families do we have?
num_unique_families <- length(unique(df_ps_ID_family$Family))
num_unique_families
# 29 unique families

# top 20
top20_families <- df_ps_ID_family %>%
  group_by(Family) %>%
  summarise(TotalRelAbundance = sum(RelAbundance), .groups = "drop") %>%
  arrange(desc(TotalRelAbundance)) %>%
  slice_head(n = 20) %>%
  pull(Family)

# Group all non-top 20 families as "Other"
df_ps_ID_top20_family <- df_ps_ID_family %>%
  mutate(Family = ifelse(Family %in% top20_families, as.character(Family), "Other"))

# check number of top 20 families
df_ps_ID_top20_family %>% distinct(Family) %>% count()
# 21 (including others)

unique(df_ps_ID_top20_family$Family)

# PLOT
# organize Genotype with increasing depth
df_ps_ID_top20_family <- df_ps_ID_top20_family %>%
  mutate(
    Genotype = factor(Genotype, levels = c("01", "02", "03", "07", "04", "06")))

# Location
df_ps_ID_top20_family <- df_ps_ID_top20_family %>%
  mutate(
    Location = factor(Location, levels = c("Bottom", "Top")))

# when plotting ID and MY together use these colors
# for ID and MY plotted together
# combine ID and MYS
families_combined <- c(
  # Overlapping families
  "Halomonadaceae",
  "Marinobacteraceae",
  "Chlorobiaceae",
  "Idiomarinaceae",
  "Alcanivoracaceae1",
  "Alteromonadaceae",
  "Rhodobacteraceae",
  "Vibrionaceae",
  "Flavobacteriaceae",
  
  # Families only in MY
  "Endozoicomonadaceae",
  "Moraxellaceae",
  "Spirochaetaceae",
  "Marinilabiliaceae",
  "Hungateiclostridiaceae",
  "Propionibacteriaceae",
  "Desulfosarcinaceae",
  "Desulfobacteraceae",
  "Cyclobacteriaceae",
  "Pseudomonadaceae",
  "Clade_I",
  
  # Families only in ID
  "Colwelliaceae",
  "Marinomonadaceae",
  "Peptostreptococcales-Tissierellales_fa",
  "Marinifilaceae",
  "Pseudoalteromonadaceae",
  "Stappiaceae",
  "Nitrincolaceae",
  "Cryomorphaceae",
  "Arcobacteraceae",
  "Kordiimonadaceae",
  "Sphingomonadaceae"
)


colors_combined <- c(
  # Overlapping families — use MY colors
  "Halomonadaceae" = "#00538A",
  "Marinobacteraceae" = "#990000",
  "Chlorobiaceae" = "#FF6800",
  "Idiomarinaceae" = "#FFB300",
  "Alcanivoracaceae1" = "#007D35",
  "Alteromonadaceae" = "#803E75",
  "Rhodobacteraceae" = "#A6BDD7",
  "Vibrionaceae" = "#CC6699",
  "Flavobacteriaceae" = "#93AA00",
  
  # Only in MY
  "Endozoicomonadaceae" = "#53377A",
  "Moraxellaceae" = "#FF0000",
  "Spirochaetaceae" = "#FFEA00",
  "Marinilabiliaceae" = "#CC9999",
  "Hungateiclostridiaceae" = "#99FFFF",
  "Propionibacteriaceae" = "#3399CC",
  "Desulfosarcinaceae" = "#003300",
  "Desulfobacteraceae" = "#593318",
  "Cyclobacteriaceae" = "#66CC66",
  "Pseudomonadaceae" = "#CC3333",
  "Clade_I" = "#CC99FF",
  
  # Only in ID — pick from ID colors if available, else reuse MY
  "Colwelliaceae" = "#006666",   # example ID color
  "Marinomonadaceae" = "#00CC99",
  "Peptostreptococcales-Tissierellales_fa" = "#CCCC66",
  "Marinifilaceae" = "#996633",
  "Pseudoalteromonadaceae" = "#FF6699",
  "Stappiaceae" = "#99CC99",
  "Nitrincolaceae" = "#9999FF",
  "Cryomorphaceae" = "#FF99CC",
  "Arcobacteraceae" = "#0099FF",
  "Kordiimonadaceae" = "#CC00FF",
  "Sphingomonadaceae" = "#0033FF")



# map colors to families
family_colors_combined <- setNames(colors_combined, families_combined)

# add color for "Other"
family_colors_combined["Other"] <- "grey70"

# sort families alphabetically but keep Other first
family_colors_sorted_combined <- c("Other", sort(setdiff(names(family_colors_combined), "Other")))

# reorder color vector
family_colors_combined <- family_colors_combined[family_colors_sorted_combined]

# set Family as factor with Other first
df_ps_ID_top20_family$Family <- factor(
  df_ps_ID_top20_family$Family,
  levels = names(family_colors_combined)  # Other first, top families after
)
# plot
bacteria_family_plot_ID <- ggplot(df_ps_ID_top20_family, aes(x = RelAbundance, y = Location, fill = Family)) +
  geom_bar(stat = "identity", position = "stack") +
  scale_x_continuous(
    limits = c(0, 1),                    # ensures 0–100% scale
    labels = scales::percent_format(),   # show as 0%, 10%, … 100%
    expand = c(0, 0)                     # remove extra padding
  ) +
  facet_wrap(~Genotype, ncol = 1) +
  scale_fill_manual(values = family_colors_combined) +
  theme_minimal(base_size = 13) +
  theme(
    strip.text = element_text(face = "bold", size = 13),
    legend.position = "bottom",
    legend.text = element_text(size = 13),
    legend.title = element_text(size = 13),
    axis.text = element_text(size = 13),
    axis.title = element_text(size = 13),
    panel.grid = element_blank(),
    panel.border = element_blank()
  ) +
  guides(fill = guide_legend(ncol = 3)) +
  labs(
    x = "Relative Abundance [%]",
    y = "Sampling Location",
    fill = "Bacterial Family"
  )

bacteria_family_plot_ID
```

## Malaysia: DADA2

Data processing with [DADA2](https://github.com/benjjneb/dada2) to infer ASVs

```{r}
# read file
cat("Reading files") # prints a message to indicate the script is starting to read files
path <- "." # current working directory
list.files(path) # lists all files in the current directory 

# the files Por2_E5 and E6 and W5 and W6 were removed before the analysis to ensure a standaridzed sampling acorss colony surface and across colonies

# Forward and reverse fastq filenames have format: 
fnFs <- sort(list.files(path, pattern="_R1_001.fastq", full.names = TRUE))
fnRs <- sort(list.files(path, pattern="_R2_001.fastq", full.names = TRUE))
# searches for files matching specific patterns _R1_001...and sorts files, get a sorted list of file paths for forward (fnFs) and reverse reads (fnRs)

# Extract sample names, assuming filenames have format
sample.names <- sapply(strsplit(basename(fnFs), "__|_0"), function(x) x[2])

cat("Processing",length(sample.names),"samples:", sample.names)
# outputs a summary of the number of samples detected and their names, legth(): counts the number, cat():prints the message

#Inspect read quality profiles
plotQualityProfile(fnFs[1:2])
# generates a quality profile plot for the first two forward read files in the fnFs list, DADA2 package in R for processing high-throughput sequence data, to visualize the quality score of sequencing reads across all base positions in a FASTQ file
# selected the first two 1:2 forward read files in fnFs 

# gray-scale: heat map of the frequency of each quality score at each base position
# green line: mean quality score at each position
# orange line: quartiles of the quality score distribution
# red line: scaled proportion of reads that extend to at least that position

# forwards reads
# --> Truncate the forward reads at position 280

plotQualityProfile(fnRs[1:2])

# reversed reads
# worse quality, especially at the end which is common in Illumina sequencing
# DADA2 incorporates quality information into its error model which makes the algorithm robust to lower quality sequence

# reversed reads
# --> Truncate the forward reads at position 200

```

### Filtering and Trimming

This pipeline ensures that only high-quality reads are retained for downstream analysis, which is crucial for the accuracy of biological insights derived from sequencing data.

```{r}
# filter and trimm 
cat("Filtering and trimming")

# prints a message to indicate that the filtering and triming steps are about to begin in the sequencing data processing pipeline

# Place filtered files in filtered/ subdirectory
filtFs <- file.path(path, "filtered", paste0(sample.names, "_F_filt.fastq.gz"))
filtRs <- file.path(path, "filtered", paste0(sample.names, "_R_filt.fastq.gz"))
names(filtFs) <- sample.names
names(filtRs) <- sample.names

### set parameters  maxN=0 (DADA2 requires no Ns), truncQ=2, rm.phix=TRUE and maxEE=2. The maxEE parameter sets the maximum number of “expected errors” allowed in a read, which is a better filter than simply averaging quality scores.
out <- filterAndTrim(fnFs, filtFs, fnRs, filtRs, truncLen=c(280,200),maxN=0, maxEE=c(2,2), truncQ=5, rm.phix=TRUE, compress=TRUE, multithread=TRUE) 
head(out)

```

### Learn error rates & sample interference

```{r}
# learn error rates
# The error model is learned using the filtered and trimmed sequences, which helps in determining the probability of errors at each base position across the reads.Forward (errF) and reverse (errR) errors are learned separately because the forward and reverse reads often have different error profiles.

cat("Learning error rates")
errF <- learnErrors(filtFs, multithread=TRUE)
errR <- learnErrors(filtRs, multithread=TRUE)
# This function in DADA2 is used to learn the error rates for sequencing reads.
# multithread=TRUE: Enables multi-threading to speed up the error rate learning process if the system supports it.

# visualize the esstimated error rates as a sanity check
plotErrors(errF, nominalQ = T)
# everything looks resonable (black line follows the points, the model fits) and we proceed with confidence
```

### Sample Inference

```{r}
# This step is essential for transforming the raw sequencing data into biologically meaningful information by correcting errors and identifying the true sequences present in your samples
# apply the core sample inference algorithm to the filtered and trimmed sequence data.
dadaFs <- dada(filtFs, err=errF, multithread=TRUE, pool = T)
dadaRs <- dada(filtRs, err=errR, multithread=TRUE, pool = T)
# It uses the error rates learned from the reverse reads (errR) to apply the error-correction model and infer the true biological sequences.

#Inspecting the returned dada-class object:
dadaFs[[1]]
dadaRs[[1]]
```

### Merging paired reads and checking on read length

```{r}
mergers <- mergePairs(dadaFs, filtFs, dadaRs, filtRs, verbose=TRUE)
# merges forward (F) and reveres (R) sequences and filtered data in one column
# Inspect the merger data.frame from the first sample
head(mergers[[1]])

# Construct sequencing table
seqtab <- makeSequenceTable(mergers)

dim(seqtab)

# Inspect distribution of sequence lengths
table(nchar(getSequences(seqtab)))

# check visually how the districution of read length is
# Get the read length distribution
length_table <- table(nchar(getSequences(seqtab)))

# Convert to data frame for plotting
length_df <- as.data.frame(length_table)
colnames(length_df) <- c("Length", "Count")

# Make sure Length is numeric
length_df$Length <- as.numeric(as.character(length_df$Length))

ggplot(length_df, aes(x = Length, y = Count)) +
  geom_col(fill = "steelblue") +  # vertical bars
  labs(title = "Distribution of Read Lengths",
       x = "Read Length (bp)",
       y = "Number of Reads") +
  xlim(280, 400)+
  theme_minimal()

```

### Remove Chimeras

```{r}
# Remove chimeras  
# Chimeric sequences are artifacts of the PCR amplification process. These sequences do not represent real biological sequences but rather result from errors in the molecular processes used to prepare and sequence DNA.
# Chimeric sequences can inflate diversity estimates. Removing them ensures that the analysis focuses only on true biological sequences.
# This code removes chimeric sequences from the sequence table, calculates the dimensions of the updated sequence table and determines the proportion of non-chimeric sequences retained.

seqtab.nochim_MY <- removeBimeraDenovo(seqtab, method="consensus", multithread=TRUE, verbose=TRUE)

dim(seqtab.nochim_MY)

## calculate the  frequency of chimeric sequences 
sum(seqtab.nochim_MY)/sum(seqtab)


# Track reads through the pipeline 
# This block of code tracks the number of reads at each stage of the DADA2 pipeline for each sample. This tracking helps monitor data loss, identify problematic samples, and ensure the pipeline is functioning as expected.
getN <- function(x) sum(getUniques(x))

# custom function that calculates the total number of unique reads (amplicon sequences)
track <- cbind(out, sapply(dadaFs, getN), sapply(dadaRs, getN), sapply(mergers, getN), rowSums(seqtab.nochim_MY))
# combines data from multiple steps into a single table
# If processing a single sample, remove the sapply calls: e.g. replace sapply(dadaFs, getN) with getN(dadaFs)
colnames(track) <- c("input", "filtered", "denoisedF", "denoisedR", "merged", "nonchim")
rownames(track) <- sample.names
head(track)

# download table
# before downloading remove the tail of the name
track_adjusted <- track

rownames(track_adjusted) <- rownames(track_adjusted) |>
  sub("_S.*", "", x = _)

write.xlsx(track_adjusted, file = "202404_MY_PLT_stats_mothur_overview_table.xlsx", rowNames = TRUE)
```

### Assign Taxonomy

```{r}
# Assign taxonomy
# This block of code assigns taxonomic classifications to the non-chimeric sequences (ASVs) in your dataset, enabling the identification of the taxa represented by the sequences

taxa_MY <- assignTaxonomy(seqtab.nochim_MY,"/yourpath/silva_nr_v138_train_set.fa.gz" , multithread=TRUE)
# uses a reference database to classify each unique sequence (ASV) in the seqtab.nochim table to a taxonomic level
# rows: correspond to unique sequences (ASV)

##optional: make species level assignments based on exact matching
taxa_MY <- addSpecies(taxa_MY, "/yourpath/silva_species_assignment_v138.fa.gz")
                   
#inspect the taxonomic assignments:
taxa.print_MY <- taxa_MY # Removing sequence rownames for display only
rownames(taxa.print_MY) <- NULL
head(taxa.print_MY)
```

### Prepare Data for plotting

```{r}
# add needed info to data frame
samples.out_MY <- rownames(seqtab.nochim_MY)
info_split_MY <- strsplit(samples.out_MY, "_")
Coral_Species <- sapply(info_split_MY, function(x) sub("[0-9]+", "", x[1]))
Genotype <- sapply(info_split_MY, function(x) gsub("[^0-9]", "", x[1]))
Orientation <- sapply(info_split_MY, function(x) sub("[0-9]+", "", x[2]))
Location <- sapply(info_split_MY, function(x) gsub("[^0-9]", "", x[2]))
samdf_MY <- data.frame(Coral_Species, Genotype, Orientation, Location, stringsAsFactors = FALSE)
rownames(samdf_MY) <- samples.out_MY

# check the amount of sampling locations
table(samdf_MY$Location)

# add the name of the Location
samdf_MY <- samdf_MY %>%
  mutate(
    Location_names = case_when(
      Location == "1" ~ "Top",
      Location == "2" ~ "Upper Middle",
      Location == "3" ~ "Lower Middle",
      Location == "4" ~ "Bottom",
      TRUE ~ NA_character_   # just in case other values exist
    )
  )

# add column Condition to meta data
samdf_MY$Condition <- paste(samdf_MY$Location_names, samdf_MY$Orientation, sep = "_")

# save metadata
samdf_MY_download <- samdf_MY

# remove tail from sample name
rownames(samdf_MY_download) <- rownames(samdf_MY_download) |>
  sub("_S.*", "", x = _)

# download
write.xlsx(samdf_MY_download, file = "202404_MY_PLT_metadata.xlsx", rowNames = TRUE)

# Prepare data - FILTERING - seqtab.nochim
# turn matrix
seqtab.nochim_MY_t <- t(seqtab.nochim_MY)

# Filter: Keep only rows with sum > 10
row_sums <- rowSums(seqtab.nochim_MY_t)

seqtab.nochim_MY_filtered_t <- 
  seqtab.nochim_MY_t[row_sums > 10, ]

# before filtering
nrow(seqtab.nochim_MY_t)
# 4335

# after filtering
nrow(seqtab.nochim_MY_filtered_t)
# 3623

# turn matrix back so it was as before
seqtab.nochim_MY_filtered <- t(seqtab.nochim_MY_filtered_t)
# filtered seqtab table

# FILTER: taxa table
# filter taxa table based on the rows from the seqtab.nochim table
filtered_asvs <- rownames(seqtab.nochim_MY_filtered_t)

# subset the taxonomy table to keep only these ASVs
taxa_MY_filtered <- taxa_MY[filtered_asvs, ]

# filter contaminations every kingdom that is not eukaryotes
# first check how many rows are not bacteria
taxa_df <- as.data.frame(taxa_MY_filtered)

# now filter
# Keep only rows where Kingdom is "Bacteria" AND Family is not NA
taxa_MY_filtered_onlybac <- taxa_df[ !is.na(taxa_df$Kingdom) &  taxa_df$Kingdom == "Bacteria", ]

# after filtering
nrow(taxa_MY_filtered_onlybac)
# 3453

# make the data frame back into a matrix
# Convert to matrix
taxa_MY_filtered_onlybac <- as.matrix(taxa_MY_filtered_onlybac)

# now remove the rows that were removed from the taxa table also from the asv table before constructing physeq dataframe
# what rows should be kept
keep_ASVs <- rownames(taxa_MY_filtered_onlybac)

# remove filtered rows
seqtab.nochim_MY_filtered_onlybac_t <- seqtab.nochim_MY_filtered_t[keep_ASVs, , drop = FALSE]

# after filtering
nrow(seqtab.nochim_MY_filtered_onlybac_t)
# 3453

# turn matrix
seqtab.nochim_MY_filtered_onlybac <- t(seqtab.nochim_MY_filtered_onlybac_t)

# CONSTRUCT PHYLSEQ
# construct a phyloseq object. with filtered seqtab.nochim and taxa data and ASV names
ps_MY_filtered <- phyloseq(otu_table(seqtab.nochim_MY_filtered_onlybac, taxa_are_rows=FALSE), 
               sample_data(samdf_MY), 
               tax_table(taxa_MY_filtered_onlybac))

# store DNA sequences of our ASV in the refseq slot of the phyloseq object
dna_MY <- Biostrings::DNAStringSet(taxa_names(ps_MY_filtered))
names(dna_MY) <- taxa_names(ps_MY_filtered)
ps_MY_filtered <- merge_phyloseq(ps_MY_filtered, dna_MY)
taxa_names(ps_MY_filtered) <- paste0("ASV", seq(ntaxa(ps_MY_filtered)))
ps_MY_filtered


# FOR DOWNLOADING
# asv table
asv_df_MY <- as.data.frame(otu_table(ps_MY_filtered))

# turn matrix
asv_df_MY <- t(asv_df_MY)

# remove not needed sample name tail
asv_df_MY_download <- asv_df_MY

colnames(asv_df_MY_download) <- colnames(asv_df_MY_download) |>
  sub("_S.*", "", x = _)

# download taxa table with DNA sequences and with ASV names
write.xlsx(asv_df_MY_download, file = "202404_MY_PLT_asv_table_filtered_onlybac.xlsx", rowNames = TRUE)


# extract taxa table with ASV
taxa_df_MY <- as.data.frame(tax_table(ps_MY_filtered))

# download taxa table with DNA sequences and with ASV names
write.xlsx(taxa_df_MY, file = "202404_MY_PLT_taxa_table_asv_filtered_onlybac.xlsx", rowNames = TRUE)


# DNA seq table
# use: seqtab.nochim_t_filtered
# rename sample name: only the important info
seqtab.nochim_MY_filtered_onlybac_t_download <- seqtab.nochim_MY_filtered_onlybac_t

colnames(seqtab.nochim_MY_filtered_onlybac_t_download) <- colnames(seqtab.nochim_MY_filtered_onlybac_t_download) |> sub("_S.*", "", x = _)

# download ASV table with DNA sequences (filtered: >10 count removed and the sampling locations in colony 2)
write.xlsx(seqtab.nochim_MY_filtered_onlybac_t_download, file = "202404_MY_PLT_DNA_seq_filtered_onlybac_table.xlsx", rowNames = TRUE)

# create fasta file
# extract DNA sequences from phyloseq object
asv_seqs_MY <- refseq(ps_MY_filtered)

# check number of sequences
length(asv_seqs_MY)   # should be 3453

# write to FASTA file
Biostrings::writeXStringSet(asv_seqs_MY, "202404_MY_PLT_ASV_DNAseq_filtered_onlybac.fasta")
```

### Barplot: Bacterial Family top 20

```{r}
# prepare data for plotting
# Groups ASVs by Family level, Counts are summed up for each Family
ps_MY_family <- tax_glom(ps_MY_filtered, taxrank = "Family")

# what are the most abundant families in %
ps_MY_family_rel <- transform_sample_counts(ps_MY_family, function(x) x / sum(x) * 100)

# calculate mean abundance per family across all samples
family_abundance_stats <- psmelt(ps_MY_family_rel) %>%
  group_by(Family) %>%
  summarise(
    Mean_Abundance = round(mean(Abundance), 3),
    SD = round(sd(Abundance), 3),
    N = n(),
    .groups = "drop"
  ) %>%
  arrange(desc(Mean_Abundance))

# view most abundant families
head(family_abundance_stats, 10)

# download table
write.xlsx(family_abundance_stats, 
           file = "202404_MY_PLT_family_mean_abundance.xlsx", 
           rowNames = FALSE)

# make dataframe
df_ps_MY_family <- psmelt(ps_MY_family)

# clean sample column name
df_ps_MY_family$Sample <- sub("_S.*", "", df_ps_MY_family$Sample)

# rename families that are NA as unknown
df_ps_MY_family <- df_ps_MY_family %>%
  mutate(Family = ifelse(is.na(Family), "Unknown", Family))

# 3. Normalize abundance to relative abundance per sample
df_ps_MY_family <- df_ps_MY_family %>%
  group_by(Sample) %>%
  mutate(RelAbundance = Abundance / sum(Abundance)) %>%
  ungroup()

# download taxa table with DNA sequences and with ASV names
write.xlsx(df_ps_MY_family, file = "202404_MY_PLT_df_ps_MY_family.xlsx", rowNames = TRUE)

# total number of reads
sum(df_ps_MY_family$Abundance)

# how many unique Families do we have?
num_unique_families <- length(unique(df_ps_MY_family$Family))
num_unique_families

# top 20
top20_families <- df_ps_MY_family %>%
  group_by(Family) %>%
  summarise(TotalRelAbundance = sum(RelAbundance), .groups = "drop") %>%
  arrange(desc(TotalRelAbundance)) %>%
  slice_head(n = 20) %>%
  pull(Family)

# Group all non-top 20 families as "Other"
df_ps_MY_top20_family <- df_ps_MY_family %>%
  mutate(Family = ifelse(Family %in% top20_families, as.character(Family), "Other"))

# check number of top 20 families
df_ps_MY_top20_family %>% distinct(Family) %>% count()
# 21 (including others)

unique(df_ps_MY_top20_family$Family)

# PLOT
# organize Genotype
df_ps_MY_top20_family <- df_ps_MY_top20_family %>%
  mutate(
    Genotype = factor(Genotype, levels = c("1", "6", "2", "3", "5", "4")))

# Orientation
df_ps_MY_top20_family <- df_ps_MY_top20_family %>%
  mutate(
    Orientation = factor(Orientation, levels = c("W", "E")))

# Location
df_ps_MY_top20_family <- df_ps_MY_top20_family %>%
  mutate(
    Location_names = factor(Location_names, levels = c("Bottom", "Lower Middle", "Upper Middle", "Top")))


# combine ID and MYS asign colors to overlapping and unique families
families_combined <- c(
  # Overlapping families
  "Halomonadaceae",
  "Marinobacteraceae",
  "Chlorobiaceae",
  "Idiomarinaceae",
  "Alcanivoracaceae1",
  "Alteromonadaceae",
  "Rhodobacteraceae",
  "Vibrionaceae",
  "Flavobacteriaceae",
  
  # Families only in MY
  "Endozoicomonadaceae",
  "Moraxellaceae",
  "Spirochaetaceae",
  "Marinilabiliaceae",
  "Hungateiclostridiaceae",
  "Propionibacteriaceae",
  "Desulfosarcinaceae",
  "Desulfobacteraceae",
  "Cyclobacteriaceae",
  "Pseudomonadaceae",
  "Clade_I",
  
  # Families only in ID
  "Colwelliaceae",
  "Marinomonadaceae",
  "Peptostreptococcales-Tissierellales_fa",
  "Marinifilaceae",
  "Pseudoalteromonadaceae",
  "Stappiaceae",
  "Nitrincolaceae",
  "Cryomorphaceae",
  "Arcobacteraceae",
  "Kordiimonadaceae",
  "Sphingomonadaceae"
)

colors_combined <- c(
  # Overlapping families — use MY colors
  "Halomonadaceae" = "#00538A",
  "Marinobacteraceae" = "#990000",
  "Chlorobiaceae" = "#FF6800",
  "Idiomarinaceae" = "#FFB300",
  "Alcanivoracaceae1" = "#007D35",
  "Alteromonadaceae" = "#803E75",
  "Rhodobacteraceae" = "#A6BDD7",
  "Vibrionaceae" = "#CC6699",
  "Flavobacteriaceae" = "#93AA00",
  
  # Only in MY
  "Endozoicomonadaceae" = "#53377A",
  "Moraxellaceae" = "#FF0000",
  "Spirochaetaceae" = "#FFEA00",
  "Marinilabiliaceae" = "#CC9999",
  "Hungateiclostridiaceae" = "#99FFFF",
  "Propionibacteriaceae" = "#3399CC",
  "Desulfosarcinaceae" = "#003300",
  "Desulfobacteraceae" = "#593318",
  "Cyclobacteriaceae" = "#66CC66",
  "Pseudomonadaceae" = "#CC3333",
  "Clade_I" = "#CC99FF",
  
  # Only in ID — pick from ID colors if available, else reuse MY
  "Colwelliaceae" = "#006666",   # example ID color
  "Marinomonadaceae" = "#00CC99",
  "Peptostreptococcales-Tissierellales_fa" = "#CCCC66",
  "Marinifilaceae" = "#996633",
  "Pseudoalteromonadaceae" = "#FF6699",
  "Stappiaceae" = "#99CC99",
  "Nitrincolaceae" = "#9999FF",
  "Cryomorphaceae" = "#FF99CC",
  "Arcobacteraceae" = "#0099FF",
  "Kordiimonadaceae" = "#CC00FF",
  "Sphingomonadaceae" = "#0033FF")

# map colors to families
family_colors_combined <- setNames(colors_combined, families_combined)

# add color for "Other"
family_colors_combined["Other"] <- "grey70"

# sort families alphabetically but keep Other first
family_colors_sorted_combined <- c("Other", sort(setdiff(names(family_colors_combined), "Other")))

# reorder color vector
family_colors_combined <- family_colors_combined[family_colors_sorted_combined]

# set Family as factor with Other first
df_ps_MY_top20_family$Family <- factor(
  df_ps_MY_top20_family$Family,
  levels = names(family_colors_combined)  # Other first, top families after
)


# plot
bacteria_family_plot_MY <- ggplot(df_ps_MY_top20_family, 
                                  aes(x = RelAbundance, y = Location_names, fill = Family)) +
  geom_bar(stat = "identity", position = "stack") +
  scale_x_continuous(
    limits = c(0, 1),                      # ensures 0–100%
    labels = scales::percent_format(),     # show as %
    expand = c(0, 0)
  ) +
  facet_grid(Genotype ~ Orientation, scales = "free_y") +
  scale_fill_manual(values = family_colors_combined) +
  theme_minimal(base_size = 13) +
  theme(
    strip.text = element_text(face = "bold", size = 13),
    legend.position = "bottom",
    legend.text = element_text(size = 13),
    legend.title = element_text(size = 13),
    axis.text = element_text(size = 13),
    axis.title = element_text(size = 13),
    panel.grid = element_blank(),
    panel.border = element_blank()
  ) +
  guides(fill = guide_legend(ncol = 3)) +
  labs(
    x = "Relative Abundance",
    y = "Sampling Location",
    fill = "Bacterial Family"
  )

bacteria_family_plot_MY
```

## **Fig. 4 Bacterial community composition across the surface of massive *P. cf. lobata* colonies from Indonesia and Malaysia**

```{r}
# Combine bacterial community plots of ID and MY 
final_combined_plot <- bacteria_family_plot_ID + bacteria_family_plot_MY +
  plot_layout(ncol = 2, widths = c(1, 1)) +  # Equal widths for both plots
  plot_annotation(
    theme = theme(
      plot.margin = margin(10, 10, 10, 10)
    )
  )

final_combined_plot

# figure was saved and edited in affinity designer for the manuscript
ggsave(
  filename = "16S_MY_ID.pdf",
  plot = final_16S_combined_plot,
  width = 11.7,      # A4 width in inches
  height = 10,     # A4 height in inches
  units = "in"
)
```

# Statistical tests

### Paired t-test and repeated-measures ANOVA

```{r}
# Indonesia
# checking assumptions
# Is data normally distributed
hist(Final_ED50_ID$ED50) # looks a little right skewed
shapiro.test(Final_ED50_ID$ED50) # p-value: p-value = 0.6716, not signfiicant, data is normally distributed
qqnorm(Final_ED50_ID$ED50) # chekcing QQ Plot: normal distributed, looks good! 
# data is normally distributed

# check for differences in variance between the top and bottom sampling location
leveneTest(ED50 ~ Condition, data = Final_ED50_ID)
# p-value = 0.7608, not significant difference in the variance of the two groups

# Is Ed50 significantly different within colonies?
# Used: paired t.test to take the nested design into account
paired_ED50 <- Final_ED50_ID %>%
  filter(Condition %in% c("Top", "Bottom")) %>%
  select(Genotype, Condition, ED50) %>%
  pivot_wider(names_from = Condition,
              values_from = ED50)

# paired t-test
t.test(paired_ED50$Top,
       paired_ED50$Bottom,
       paired = TRUE)

var.test(ED50 ~ Region, data = cbass_data)


# Malaysia
# check data distribution: normality and homogeneity
hist(Final_ED50_MY$ED50) 
# normallity test
shapiro.test(Final_ED50_MY$ED50) # p-value: not signifiacnt, data normally distributed 
# plot normallity
qqnorm(Final_ED50_MY$ED50) # looks normally distributed

# test homogeneity
leveneTest(ED50 ~ Condition, data = Final_ED50_MY)
# p-value:not significant use ANOVA


# Repeated measures ANOVA: is ed50 significantly different within colonies?
model_rm <- aov(ED50 ~ Condition + Error(Genotype/Condition), 
                data = Final_ED50_MY)

summary(model_rm)


# is ed50 significantly different across depth
# Correlation test to check the effect of depth on ED50

# check normality of ED50 and depth
shapiro.test(Final_ED50_MY$ED50) # normally distributed
shapiro.test(Final_ED50_MY$Depth) # not normally distirbuted
# Spearman correlation safer choice
# Spearman: Relationship is monotonic (consistently increasing or decreasing, not necessarily linear), Data can be skewed, ordinal, or non-normal

# visually inspect
plot(Final_ED50_MY$Depth, Final_ED50_MY$ED50,
     pch = 16,
     xlab = "Depth",
     ylab = "ED50",
     main = "ED50 across Depth")


# Spearman correlation test
cor_test_spearman <- cor.test(
  Final_ED50_MY$ED50,
  Final_ED50_MY$Depth,
  method = "spearman"
)

# View results
cor_test_spearman

```

### ANCOM-BC2

```{r}
# ANCOM-BC2
# Indonesia
# use phyloseq object from DADA2 pipeline: ps_ID24_filtered
# set 
set.seed(124)

# ANCOMBC2
# set reference level, to make sure it is Top relative to Bottom (Bottom: reference)
sample_data(ps_ID24_filtered)$Location <- factor(
  sample_data(ps_ID24_filtered)$Location,
  levels = c("Bottom", "Top")
)

# ANCOM-BC2
ancombc2_ID <- ancombc2(
  data = ps_ID24_filtered,
  fix_formula = "Location", # the fixed factor we are interested in
  rand_formula = "(1 | Genotype)", # accounting for the nested design and repeated measure
  p_adj_method = "BH", # method to adjust p-value for multiple testing, by controlling FDR, Benjamini-Hochberg: common method for ASV data not too conservative and balances false positives and false ne.
  pseudo_sens = TRUE, # checks if significant taxa are sensitive to pseudo counts (diff_robust:..)
  prv_cut = 0.10, # removed very rare ASV presen
  lib_cut = 100, # removes samples with low read number
  s0_perc = 0.05,  # stabilizes the test statistics by adding a small constant, default
  alpha = 0.05, # standard significance threshold
  group = NULL, # NULL because only 2 groups
  struc_zero = FALSE, # small sample size so start with False
  neg_lb = FALSE, # controls that a conservative lower bound-based approach, only used when struc_zero = T, rather for larger sample size
  pairwise = FALSE, # pairwise comparison, not used because only two groups
  global = FALSE, # only two groups
  tax_level = NULL, # ASV level wanted
  em_control = list(tol = 1e-05, max_iter = 100), # default: algorithm calcultation rounds and threshold
  verbose = TRUE # shows what is happening while the alogrith is running
)


# ancom results
res_primary_ID = ancombc2_ID$res

# how many ASVs were tested in total
nrow(res_primary_ID) 

# 528

# count how many are NA
sum(is.na(res_primary_ID$"lfc_(Intercept)"))
# 35

# find significant ASVS across Location
# counts ASVs that are significant in at least one comparison
sig_asvs_ID <- res_primary_ID[res_primary_ID$diff_LocationTop == TRUE, ]

# how many remain
nrow(sig_asvs_ID)
# 16

# extract taxonomy from phyloseq
tax_ID <- tax_table(ps_ID_filtered)
tax_ID_df <- as.data.frame(tax_ID)
tax_ID_df$taxon <- rownames(tax_ID_df)

# join taxonomy
sig_asvs_tax_ID <- sig_asvs_ID %>%
  left_join(
    tax_ID_df %>% select(taxon, Phylum, Order, Family, Genus, Species),
    by = "taxon"
  )

# replace missing Family names
sig_asvs_tax_ID <- sig_asvs_tax_ID %>%
  mutate(Family = ifelse(is.na(Family), "Unknown", Family))

# create combined ASV-family label
sig_asvs_tax_ID <- sig_asvs_tax_ID %>%
  mutate(ASV_Family = paste(taxon, Family, sep = "_"))


# keep only the log fold change column for Top vs Bottom
heatmap_df_ID <- sig_asvs_tax_ID %>%
  dplyr::select(ASV_Family, lfc_LocationTop) %>%
  dplyr::rename(LFC = lfc_LocationTop) %>%
  dplyr::mutate(Comparison = "Top_vs_Bottom")

# order rows by log fold change
asv_order_ID <- heatmap_df_ID %>%
  arrange(LFC) %>%
  pull(ASV_Family)

heatmap_df_ID$ASV_Family <- factor(heatmap_df_ID$ASV_Family, levels = asv_order_ID)

# heatmap
ANCOMBC2_ID_heatmap <- ggplot(heatmap_df_ID, aes(x = Comparison, y = ASV_Family, fill = LFC)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(LFC, 2)), size = 3, color = "black") +
  scale_fill_gradient2(
    low = "blue",
    mid = "white",
    high = "red",
    midpoint = 0,
    na.value = "grey90"
  ) +
  
  labs(
    x = "Comparison",
    y = "ASV_Family",
  fill = "Log fold change"
  ) +
  
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 0, hjust = 0.5), 
    panel.grid = element_blank()
  )

ANCOMBC2_ID_heatmap


# Malayisa
# use phyloseq object from DADA2 pipeline: ps_MY24_filtered

# ancom
ancombc2_MY <- ancombc2(
  data = ps_MY24_filtered,
  fix_formula = "Condition", #fixed factor we are interested in
  rand_formula = "(1 | Genotype)", # accounting for the nested design and repeated measure
  group="Condition", 	
  p_adj_method = "BH", # common method for ASV data
  pseudo_sens = TRUE,
  prv_cut = 0.1, # removed very rare ASV presen:
  lib_cut = 100, # removes samples with low read number, very mild can be up to 1000
  s0_perc = 0.05,  # stabilizes the test statistics by adding a small constant, default
  alpha = 0.05, # standard significance threshold
  struc_zero = FALSE, # small sample size so start with False, check out later with T
  neg_lb = FALSE, # only T when struc_zero = T, rather for larger sample size
  pairwise = TRUE, 
  global = TRUE, 
  tax_level = NULL, # ASV level wanted
  em_control = list(tol = 1e-05, max_iter = 100), # default: algorithm calcultation rounds and threshold
  verbose = TRUE # shows what is happening while the alogrith is running
)

# check ancom results
res_primary <- ancombc2_MY$res

# global
res_global <- ancombc2_MY$res_global

sig_global <- res_global[ancombc2_MY$diff_abn == TRUE, ]
sig_global
```

### Wilcox-test: Within-colony ED50 variability across study regions

```{r}
# Use prior code to claculate ED50 data or upload ED50 data directly from the Supplementary Data: Intracolony CBASS SuppData.xlsx
input_data_path <- selectFile(
  caption = "Select XLSX Input File"
)

# select ED50 ID and MY data
ED50_ID_within  <- read_excel(input_data_path, sheet = "Data S1. ID")
ED50_MY_within <- read_excel(input_data_path, sheet = "Data S2. MY")

# simplify the column names that are needed for the analysis
ED50_ID_within <- ED50_ID_within %>%
  dplyr::rename(
    ED50 = "ED50 [°C]",
    Genotype = "Colony no.",
    Sampling_location = "Sampling Location"
  )

# simplify the column names that are needed for the analysis
ED50_MY_within <- ED50_MY_within %>%
  dplyr::rename(
    ED50 = "ED50 [°C]",
    Genotype = "Colony no.",
    Sampling_location = "Sampling Location",
    Orientation = "Cardinal Direction"
  )

# Indonesia
# calculate top-bottom difference per colony
ID_variability <- ED50_ID_within %>%
  select(Genotype, Sampling_location, ED50) %>%
  pivot_wider(            
    names_from = Sampling_location,
    values_from = ED50
  ) %>%
  mutate(
    ED50_difference = round(abs(Top - Bottom),3) 
  )

ID_variability

# Malaysia
# first average east and west
MY_avg <- ED50_MY_within %>%
  group_by(Genotype, Sampling_location) %>%
  summarise(
    ED50_mean = round(mean(ED50, na.rm = TRUE),2),
    .groups = "drop"
  )

MY_avg

# calculate top-bottom differences
MY_variability <- MY_avg %>%
  pivot_wider(
    names_from = Sampling_location,
    values_from = ED50_mean
  ) %>%
  mutate(
    ED50_difference = abs(Top - Bottom)
  )

MY_variability

# combine data
ID_variability$Country <- "Indonesia"
MY_variability$Country <- "Malaysia"

ID_MY_comparison <- bind_rows(
  ID_variability %>%
    select(Country, Genotype, ED50_difference),
  MY_variability %>%
    select(Country, Genotype, ED50_difference)
)

# statistically testing for it
wilcox.test(
  ED50_difference ~ Country,
  data = ID_MY_comparison)

# p-value: 0.4704
```

### PERMANOVA: Bacteria and ED50 association

```{r}
# use precreated phyloseq object ps_ID_filtered or upload bacteria data and create phyloseq object

# Indonesia
# Load metadata
samdf_ID_within <- read.xlsx('/Users/MarlenSchlotheuber/Nextcloud/10. projects/11. PhD/10. projects/Intraspecific bleaching project/Project planning/MYS_2024/Fieldwork/Data_26042024/16S_Bacteria_data/16S_files/ID_2024_within/202404_ID_GAD_metadata.xlsx',rowNames = TRUE)

# Load ASV count table
seqtab.nochim_ID_filtered_onlybac_within <- read.xlsx('/Users/MarlenSchlotheuber/Nextcloud/10. projects/11. PhD/10. projects/Intraspecific bleaching project/Project planning/MYS_2024/Fieldwork/Data_26042024/16S_Bacteria_data/16S_files/ID_2024_within/202404_ID_GAD_asv_table_filtered_onlybac.xlsx',rowNames = TRUE)

seqtab.nochim_ID_filtered_onlybac_within <- as.matrix(seqtab.nochim_ID_filtered_onlybac_within)

seqtab.nochim_ID_filtered_onlybac_within_t <- t(seqtab.nochim_ID_filtered_onlybac_within)

# Load taxonomy table
taxa_ID_filtered_onlybac_within <- read.xlsx('/Users/MarlenSchlotheuber/Nextcloud/10. projects/11. PhD/10. projects/Intraspecific bleaching project/Project planning/MYS_2024/Fieldwork/Data_26042024/16S_Bacteria_data/16S_files/ID_2024_within/202404_ID_GAD_taxa_table_asv_filtered_onlybac.xlsx', rowNames = TRUE)

taxa_ID_filtered_onlybac_within <- as.matrix(taxa_ID_filtered_onlybac_within)

# Construct phyloseq object
ps_ID_filtered_within <- phyloseq(
  otu_table(seqtab.nochim_ID_filtered_onlybac_within_t, taxa_are_rows = FALSE),
  sample_data(samdf_ID_within),
  tax_table(taxa_ID_filtered_onlybac_within)
)

# Malaysia
# Load metadata
samdf_MY_within <- read.xlsx('/Users/MarlenSchlotheuber/Nextcloud/10. projects/11. PhD/10. projects/Intraspecific bleaching project/Project planning/MYS_2024/Fieldwork/Data_26042024/16S_Bacteria_data/16S_files/202404_MY_PLT_metadata.xlsx',rowNames = TRUE)

# Load ASV count table
seqtab.nochim_MY_filtered_onlybac_within <- read.xlsx('/Users/MarlenSchlotheuber/Nextcloud/10. projects/11. PhD/10. projects/Intraspecific bleaching project/Project planning/MYS_2024/Fieldwork/Data_26042024/16S_Bacteria_data/16S_files/202404_MY_PLT_asv_table_filtered_onlybac.xlsx',rowNames = TRUE)

seqtab.nochim_MY_filtered_onlybac_within <- as.matrix(seqtab.nochim_MY_filtered_onlybac_within)

seqtab.nochim_MY_filtered_onlybac_within_t <- t(seqtab.nochim_MY_filtered_onlybac_within)

# Load taxonomy table
taxa_MY_filtered_onlybac_within <- read.xlsx('/Users/MarlenSchlotheuber/Nextcloud/10. projects/11. PhD/10. projects/Intraspecific bleaching project/Project planning/MYS_2024/Fieldwork/Data_26042024/16S_Bacteria_data/16S_files/202404_MY_PLT_taxa_table_asv_filtered_onlybac.xlsx', rowNames = TRUE)

taxa_MY_filtered_onlybac_within <- as.matrix(taxa_MY_filtered_onlybac_within)

# Construct phyloseq object
ps_MY_filtered_within <- phyloseq(
  otu_table(seqtab.nochim_MY_filtered_onlybac_within_t, taxa_are_rows = FALSE),
  sample_data(samdf_MY_within),
  tax_table(taxa_MY_filtered_onlybac_within)
)


# prepare data for permanova
# check orientation: correct when false
taxa_are_rows(ps_ID_filtered_within)

# Extract ASV table
asv_ID <- as.data.frame(otu_table(ps_ID_filtered_within))

# Check dimensions
dim(asv_ID)

# calculate relative abundance from count data
asv_ID_rel <- asv_ID / rowSums(asv_ID)

# Extract metadata
meta_ID <- data.frame(sample_data(ps_ID_filtered_within))

# add sample_ID to the metadata as column
meta_ID$Sample_ID <- rownames(meta_ID)

# add ed50 to meta_ID based on sample_ID
meta_ID <- meta_ID %>%
  left_join(
    ED50_ID_within %>% select(Sample_ID, ED50),
    by = "Sample_ID"
  )

# make sample id back to row names
rownames(meta_ID) <- meta_ID$Sample_ID
meta_ID$Sample_ID <- NULL

# check col names
colnames(meta_ID)

# check that metadata matches ASV table
identical(
  rownames(asv_ID_rel),
  rownames(meta_ID)
)

# Calculate Bray-Curtis distance between all samples
bray_ID <- vegdist(asv_ID_rel, method = "bray")

# set the starting point for random number generationwaj 
set.seed(123)

# run permanova with strata function
permanova_ID <- adonis2(
  bray_ID ~ ED50,
  data = meta_ID,
  permutations = 999,
  strata = meta_ID$Genotype) # this: restricts permutations within each genotype, use when multiple samples within one genotype

permanova_ID
# not significant, 0.4219


# Malaysia
# check orientation: correct when false
taxa_are_rows(ps_MY_filtered_within)

# Extract ASV table
asv_MY <- as.data.frame(otu_table(ps_MY_filtered_within))

# Check dimensions
dim(asv_MY)

# calculate relative abundance
asv_MY_rel <- asv_MY / rowSums(asv_MY)

# Extract metadata
meta_MY <- data.frame(sample_data(ps_MY_filtered_within))

# add sample_ID to the metadata as column
meta_MY$Sample_ID <- rownames(meta_MY)

# add ed50 to emta_ID based on sample_ID
meta_MY <- meta_MY %>%
  left_join(
    ED50_MY_within %>% select(Sample_ID, ED50),
    by = "Sample_ID"
  )

# make sample id back to row names
rownames(meta_MY) <- meta_MY$Sample_ID
meta_MY$Sample_ID <- NULL

# check col names
colnames(meta_MY)

# check that metadata matches ASV table
identical(
  rownames(asv_MY_rel),
  rownames(meta_MY)
)

# Calculate Bray-Curtis distance between all samples
bray_MY <- vegdist(asv_MY_rel, method = "bray")

# control starting point
set.seed(123)

permanova_MY <- adonis2(
  bray_MY ~ ED50,
  data = meta_MY,
  permutations = 999,
  strata = meta_MY$Genotype
)

permanova_MY
# not significant, 0.422
```

### PERMANOVA: Symbiodiniaceae and ED50 association

```{r}
# code to upload the ITS2 data can be found in the ITS2 type profile chapter

# Indonesia
# Extract ITS2 profile columns
# (all columns after metadata columns)
its2_profiles_ID <- its2_data_ID[, 13:ncol(its2_data_ID)]

# Check numeric
str(its2_profiles_ID)

# change columns to numeric
its2_profiles_ID <- as.data.frame(
  lapply(its2_profiles_ID, as.numeric)
)

# Calculate Bray-Curtis dissimilarity
bray_ITS2_ID <- vegdist(its2_profiles_ID, method = "bray")

# Prepare metadata
meta_ITS2_ID <- its2_data_ID[, 1:12]

# check
identical(
  attr(bray_ITS2_ID, "Labels"),
  rownames(meta_ITS2_ID)
)


# PERMANOVA
set.seed(123)

permanova_ITS2_ID <- adonis2(
  bray_ITS2_ID ~ ED50,
  data = meta_ITS2_ID,
  permutations = 999,
  strata = meta_ITS2_ID$collection_colony
)

permanova_ITS2_ID
# pvalue: 0.9375


# Malayisa
# Extract ITS2 profile columns
# (all columns after metadata columns)
its2_profiles_MY <- its2_data_MY[, 15:ncol(its2_data_MY)]

# Check that rows sum approximately to 1
str(its2_profiles_MY)

# change columns to numeric
its2_profiles_MY <- as.data.frame(
  lapply(its2_profiles_MY, as.numeric)
)

# Calculate Bray-Curtis dissimilarity
bray_ITS2_MY <- vegdist(its2_profiles_MY, method = "bray")

# Prepare metadata
meta_ITS2_MY <- its2_data_MY[, 1:14]

# check
identical(
  attr(bray_ITS2_MY, "Labels"),
  rownames(meta_ITS2_MY)
)


# PERMANOVA
set.seed(123)

permanova_ITS2_MY <- adonis2(
  bray_ITS2_MY ~ ED50,
  data = meta_ITS2_MY,
  permutations = 999,
  strata = meta_ITS2_MY$collection_colony
)

permanova_ITS2_MY
# pvalue: 0.899
```

# Supplementary Figure

## Fig S1: **Thermal profiles during CBASS acute heat stress assays in Indonesia and Malaysia**

```{r}
# Indonesia
# upload data
ID_Hobo <- read.csv("/yourpathname/ID_GAD_Plob1-7_i-vi_CBASS_HOBO_data.csv")

# set time
ID_Hobo <- ID_Hobo %>%
  mutate(
    Temp = as.numeric(Temp),
    DateTime = as.POSIXct(
      paste(Date, Time),
      format = "%m/%d/%y %I:%M:%S%p"
    )
  )

# set colors
cbass_colors <- c(
  "28" = "blue",
  "34" = "yellow",
  "37" = "orange",
  "40" = "red"
)

# plot
plot_ID_hobo <- ggplot(ID_Hobo, aes(x = DateTime, y = Temp, color = as.factor(CBASS), group = CBASS)) +
  geom_line(linewidth = 1) +
  scale_color_manual(values = cbass_colors) +
  
  scale_x_datetime(
    limits = as.POSIXct(c("2024-08-27 12:30:00", "2024-08-27 21:30:00")),
    date_labels = "%I:%M %p"
  ) +
  
  scale_y_continuous(breaks = seq(25, 45, 5)) +
  
  theme_minimal() +
  labs(
    x = "Time",
    y = "Temperature (°C)",
    color = "CBASS"
  )


plot_ID_hobo
```

```{r}
# Malaysia
# upload hobo data
Plob1_data <- read.csv('/yourpathname/MY_PLT_Plob1_i_CBASS_HOBO_data.csv')

# set colors
cbass_colors <- c(
  "30" = "blue",
  "34" = "yellow",
  "38" = "orange",
  "42" = "red"
)

# tranform date time
Plob1_data <- Plob1_data %>%
  mutate(DateTime = as.POSIXct(DateTime, tz = "UTC")) %>%
  filter(!is.na(DateTime))

#plot
plot_Plob1 <- ggplot(Plob1_data, aes(x = DateTime, y = Temp, color = as.factor(CBASS), group = CBASS)) +
  geom_line(linewidth = 1) +
  
  scale_color_manual(
    name = "Temperature profiles",
    values = cbass_colors
  ) +
  
  scale_x_datetime(date_labels = "%I:%M %p") +
scale_y_continuous(
  limits = c(25, 45),
  breaks = seq(25, 45, 5)
) +
  
  labs(
    title = "CBASS Colony 1",
    x = "Time",
    y = "Temperature (°C)"
  ) +
  
  theme_minimal()

plot_Plob1

#calculate mean holding phase
Plob1_data <- Plob1_data %>%
  mutate(Time = format(DateTime, "%H:%M:%S"))

colony1_mean_holdingtemp <- Plob1_data %>%
  filter(Time >= "16:00:00" & Time <= "19:00:00") %>%
  group_by(CBASS) %>%
  summarise(
    mean_Temp = round(mean(Temp, na.rm = TRUE), 2),
    n = n()
  )

colony1_mean_holdingtemp

# COLONY 2
# upload hobo data
Plob2_data <- read.csv('/yourpathname/MY_PLT_Plob2_iii_CBASS_HOBO_data.csv')


# set colors
cbass_colors <- c(
  "30" = "blue",
  "34" = "yellow",
  "38" = "orange",
  "42" = "red"
)

# tranform date time
Plob2_data <- Plob2_data %>%
  mutate(DateTime = as.POSIXct(DateTime, tz = "UTC")) %>%
  filter(!is.na(DateTime))

# plot
plot_Plob2 <- ggplot(Plob2_data, aes(x = DateTime, y = Temp, color = as.factor(CBASS), group = CBASS)) +
  geom_line(linewidth = 1) +
  
  scale_color_manual(
    name = "Temperature profiles",
    values = cbass_colors
  ) +
  
  scale_x_datetime(date_labels = "%I:%M %p") +
scale_y_continuous(
  limits = c(25, 45),
  breaks = seq(25, 45, 5)
) +
  
  labs(
    title = "CBASS Colony 2",
    x = "Time",
    y = "Temperature (°C)"
  ) +
  
  theme_minimal()

plot_Plob2


#calculate mean holding phase
Plob2_data <- Plob2_data %>%
  mutate(Time = format(DateTime, "%H:%M:%S"))

colony2_mean_holdingtemp <- Plob2_data %>%
  filter(Time >= "16:00:00" & Time <= "19:00:00") %>%
  group_by(CBASS) %>%
  summarise(
    mean_Temp = round(mean(Temp, na.rm = TRUE), 2),
    n = n()
  )

colony2_mean_holdingtemp


# COLONY 3
# upload hobo data
Plob3_data <- read.csv('/yourpathname/MY_PLT_Plob3_iv_CBASS_HOBO_data.csv')


# set colors
# Temperature 42°C is in grey because the temperature was not logged so here we added the 42°C from a different CBASS run and colored it grey to highlight where the temperature should be
cbass_colors <- c(
  "30" = "blue",
  "34" = "yellow",
  "38" = "orange",
  "42" = "grey"
)

# tranform date time
Plob3_data$DateTime <- as.POSIXct(
  Plob1_data$DateTime,
  format = "%m/%d/%y %H:%M",
  tz = "UTC"
)

# plot
plot_Plob3 <- ggplot(
  Plob3_data,
  aes(x = DateTime, y = Temp,
      color = as.factor(CBASS),
      group = CBASS)
) +
  geom_line(linewidth = 1) +
  
  scale_x_datetime(
    date_labels = "%I:%M %p",
    date_breaks = "3 hours"
  ) +
  
  scale_y_continuous(
    limits = c(25, 45),
    breaks = seq(25, 45, 5)
  ) +
  
  scale_color_manual(values = cbass_colors) +
  
  labs(
    title = "CBASS Colony 3",
    x = "Time",
    y = "Temperature (°C)"
  ) +
  
  theme_minimal()

plot_Plob3

#calculate mean holding phase
Plob3_data <- Plob3_data %>%
  mutate(Time = format(DateTime, "%H:%M:%S"))

colony3_mean_holdingtemp <- Plob3_data %>%
  filter(Time >= "16:00:00" & Time <= "19:00:00") %>%
  group_by(CBASS) %>%
  summarise(
    mean_Temp = round(mean(Temp, na.rm = TRUE), 2),
    n = n()
  )

colony3_mean_holdingtemp

# COLONY 4
# upload data
Plob4_data <- read.csv('/yourpathname/MY_PLT_Plob4_vi_CBASS_HOBO_data.csv')

# set colors
cbass_colors <- c(
  "30" = "blue",
  "34" = "yellow",
  "38" = "orange",
  "42" = "red"
)

# tranform date time
Plob4_data <- Plob4_data %>%
  mutate(DateTime = as.POSIXct(DateTime, tz = "UTC")) %>%
  filter(!is.na(DateTime))

#plot
plot_Plob4 <-ggplot(Plob4_data, aes(x = DateTime, y = Temp, color = as.factor(CBASS), group = CBASS)) +
  geom_line(linewidth = 1) +
  
  scale_color_manual(
    name = "Temperature profiles",
    values = cbass_colors
  ) +
  
  scale_x_datetime(date_labels = "%I:%M %p") +
scale_y_continuous(
  limits = c(25, 45),
  breaks = seq(25, 45, 5)
) +
  
  labs(
    title = "CBASS Colony 4",
    x = "Time",
    y = "Temperature (°C)"
  ) +
  
  theme_minimal()

plot_Plob4

#calculate mean holding phase
Plob4_data <- Plob4_data %>%
  mutate(Time = format(DateTime, "%H:%M:%S"))

colony4_mean_holdingtemp <- Plob4_data %>%
  filter(Time >= "16:00:00" & Time <= "19:00:00") %>%
  group_by(CBASS) %>%
  summarise(
    mean_Temp = round(mean(Temp, na.rm = TRUE), 2),
    n = n()
  )

colony4_mean_holdingtemp


# COLONY 5
# upload hobo data
Plob5_data <- read.csv('/yourpathname/MY_PLT_Plob5_v_CBASS_HOBO_data.csv')

# set colors
cbass_colors <- c(
  "30" = "blue",
  "34" = "yellow",
  "38" = "orange",
  "42" = "red"
)

# tranform date time
Plob5_data <- Plob5_data %>%
  mutate(DateTime = as.POSIXct(DateTime, tz = "UTC")) %>%
  filter(!is.na(DateTime))

#plot
plot_Plob5 <- ggplot(
  Plob5_data %>%
    filter(between(hour(DateTime), 12, 20)),  # 12:00–20:00
  aes(x = DateTime, y = Temp, color = as.factor(CBASS), group = CBASS)
) +
  geom_line(linewidth = 1) +
  
  scale_color_manual(
    name = "Temperature profiles",
    values = cbass_colors
  ) +
  
  scale_x_datetime(date_labels = "%I:%M %p") +
scale_y_continuous(
  limits = c(25, 45),
  breaks = seq(25, 45, 5)
) +
  
  labs(
    title = "CBASS Colony 5",
    x = "Time",
    y = "Temperature (°C)"
  ) +
  
  theme_minimal()


plot_Plob5

#calculate mean holding phase
colony5_mean_holdingtemp <- Plob5_data %>%
  filter(
    DateTime >= as.POSIXct(min(DateTime, na.rm = TRUE), tz = "UTC") %>%
      floor_date(unit = "day") + hours(16),
    DateTime <= as.POSIXct(min(DateTime, na.rm = TRUE), tz = "UTC") %>%
      floor_date(unit = "day") + hours(19)
  ) %>%
  group_by(CBASS) %>%
  summarise(
    mean_Temp = round(mean(Temp, na.rm = TRUE), 2),
    n = n()
  )

colony5_mean_holdingtemp


# COLONY 6
# upload hobo data
Plob6_data <- read.csv('/yourpathname/MY_PLT_Plob6_ii_CBASS_HOBO_data.csv')


# set colors
cbass_colors <- c(
  "30" = "blue",
  "34" = "yellow",
  "38" = "orange",
  "42" = "red"
)

# tranform date time
Plob6_data <- Plob6_data %>%
  mutate(DateTime = as.POSIXct(DateTime, tz = "UTC")) %>%
  filter(!is.na(DateTime))

Plob6_data <- Plob6_data %>%
  mutate(
    DateTime = as.POSIXct(DateTime, format = "%m/%d/%y %H:%M", tz = "UTC"),
    TimeOnly = as.numeric(format(DateTime, "%H")) +
               as.numeric(format(DateTime, "%M"))/60
  )

#plot
plot_Plob6 <- ggplot(Plob6_data, aes(x = TimeOnly, y = Temp, color = as.factor(CBASS), group = CBASS)) +
  geom_line(linewidth = 1) +
  
  scale_color_manual(
    name = "Temperature profiles",
    values = cbass_colors
  ) +
  
scale_x_continuous(
  breaks = seq(12, 21, 3),
  labels = c("12 pm", "3 pm", "6 pm", "9 pm")
)+
  
scale_y_continuous(
  limits = c(25, 45),
  breaks = seq(25, 45, 5)
) +
  labs(
    title = "CBASS Colony 6",
    x = "Time",
    y = "Temperature (°C)"
  ) +
  
  theme_minimal()

plot_Plob6

#calculate mean holding phase
Plob6_data <- Plob6_data %>%
  mutate(Time = format(as.POSIXct(DateTime), "%H:%M:%S"))

colony6_mean_holdingtemp <- Plob6_data %>%
  filter(Time >= "16:00:00" & Time <= "19:00:00") %>%
  group_by(CBASS) %>%
  summarise(
    mean_Temp = round(mean(Temp, na.rm = TRUE), 2),
    n = n()
  )

colony6_mean_holdingtemp


# PLot all Colonies in one plot
# plot hobo
MY_plot_hobo <- (plot_Plob1 | plot_Plob2 | plot_Plob3) /
(plot_Plob4 | plot_Plob5 | plot_Plob6) +
  plot_layout(guides = "collect") &
  theme(legend.position = "right")

# plot
MY_plot_hobo

```

## Fig. S2: HOBO light and temperature of coral colonies

```{r}
# colony 1
# upload data
Col_1_west <- read.csv('/Users/MarlenSchlotheuber/Nextcloud/10. projects/11. PhD/10. projects/Intraspecific bleaching project/Project planning/MYS_2024/Fieldwork/Data_26042024/HOBO Colony/18042024_MYS_SBR_colony_1/20240418_MY_SBR_Plob_1_i_west_HOBO_TempLight_clean.csv')

# upload data
Col_1_top <- read.csv('/Users/MarlenSchlotheuber/Nextcloud/10. projects/11. PhD/10. projects/Intraspecific bleaching project/Project planning/MYS_2024/Fieldwork/Data_26042024/HOBO Colony/18042024_MYS_SBR_colony_1/20240418_MY_SBR_Plob_1_i_top_HOBO_TempLight_clean.csv')

# East
# upload data
Col_1_east <- read.csv('/Users/MarlenSchlotheuber/Nextcloud/10. projects/11. PhD/10. projects/Intraspecific bleaching project/Project planning/MYS_2024/Fieldwork/Data_26042024/HOBO Colony/18042024_MYS_SBR_colony_1/20240418_MY_SBR_Plob_1_i_west_HOBO_TempLight_clean.csv')

# set position
Col_1_west$Position <- "West"
Col_1_top$Position  <- "Top"
Col_1_east$Position <- "East"

# Set order of panels
all_data$Position <- factor(
  all_data$Position,
  levels = c("West", "Top", "East")
)

#plot
# Plot Colony 1
Plot_Plob_1_hobo <- ggplot(
  all_data,
  aes(x = Date_Time)
) +
  
  # Temperature
  geom_line(
    aes(y = Temp),
    color = temp_color,
    linewidth = 0.8
  ) +
  
  # Light
  geom_line(
    aes(y = Light_scaled),
    color = light_color,
    linewidth = 0.8
  ) +
  
  # Y axes
  scale_y_continuous(
    name = "Temperature [°C]",
    limits = c(30, 35),
    breaks = seq(30, 35, 2),
    sec.axis = sec_axis(
      ~ (. - 30) * 50000,
      name = "Light [Lux]",
      breaks = seq(0, 250000, 25000)
    )
  ) +
  
  # X axis: time only
  scale_x_datetime(
    date_labels = "%H:%M"
  ) +
  
  # Three plots in one row
  facet_wrap(
    ~ Position,
    nrow = 1
  ) +
  
  labs(
    title = "Temperature and Light data of Colony 1 (i)",
    x = "Time [h]"
  ) +
  
  
  theme_classic() +
  
  theme(
    strip.background = element_blank(),
    strip.text = element_text(
      face = "bold",
      size = 12
    ),
    
    axis.title.y.left = element_text(color = temp_color),
    axis.text.y.left  = element_text(color = temp_color),
    
    axis.title.y.right = element_text(color = light_color),
    axis.text.y.right  = element_text(color = light_color),
    
    panel.spacing = unit(1, "lines")
  )

Plot_Plob_1_hobo

# save figure
ggsave(
  filename = "/Users/MarlenSchlotheuber/Nextcloud/10. projects/11. PhD/10. projects/Intraspecific bleaching project/Project planning/MYS_2024/Fieldwork/Data_26042024/HOBO Colony/Manuscript_Figures/HOBO_TempLight_Plob_1_i.pdf",
  plot = Plot_Plob_1_hobo,
  width = 11.7,      # A4 width in inches
  height = 5,     # A4 height in inches
  units = "in"
)


# colony 2
# upload data
# Col_2_west: HOBO loggers didn't log data
Col_2_west <- read.csv('/Users/MarlenSchlotheuber/Nextcloud/10. projects/11. PhD/10. projects/Intraspecific bleaching project/Project planning/MYS_2024/Fieldwork/Data_26042024/HOBO Colony/19042024_MYS_BK_colony_2/20240419_MY_BK_Plob_2_iii_west_HOBO_TempLight_clean.csv')

# upload data
Col_2_top <- read.csv('/Users/MarlenSchlotheuber/Nextcloud/10. projects/11. PhD/10. projects/Intraspecific bleaching project/Project planning/MYS_2024/Fieldwork/Data_26042024/HOBO Colony/19042024_MYS_BK_colony_2/20240419_MY_BK_Plob_2_iii_top_HOBO_TempLight_clean.csv')

# upload data
Col_2_east <- read.csv('/Users/MarlenSchlotheuber/Nextcloud/10. projects/11. PhD/10. projects/Intraspecific bleaching project/Project planning/MYS_2024/Fieldwork/Data_26042024/HOBO Colony/19042024_MYS_BK_colony_2/20240419_MY_BK_Plob_2_iii_east_HOBO_TempLight_clean.csv')

# Prepare data
Col_2_west$Position <- "West"
Col_2_top$Position  <- "Top"
Col_2_east$Position <- "East"

# Scale Light (0–250,000) to Temperature scale (30–40)
Col_2_west$Light_scaled <- 30 + (Col_2_west$Light / 90000) * 10
Col_2_top$Light_scaled  <- 30 + (Col_2_top$Light / 90000) * 10
Col_2_east$Light_scaled <- 30 + (Col_2_east$Light / 90000) * 10

# Convert Date_Time
Col_2_west$Date_Time <- as.POSIXct(
  Col_2_west$Date_Time,
  format = "%m/%d/%y %H:%M"
)

Col_2_top$Date_Time <- as.POSIXct(
  Col_2_top$Date_Time,
  format = "%m/%d/%y %H:%M"
)

Col_2_east$Date_Time <- as.POSIXct(
  Col_2_east$Date_Time,
  format = "%m/%d/%y %H:%M"
)

# Combine Colony 2 data
all_data_col2 <- bind_rows(
  Col_2_west,
  Col_2_top,
  Col_2_east
)

# Remove the 0/0 placeholder from West
all_data_col2 <- all_data_col2 %>%
  filter(!(Position == "West" & Temp == 0 & Light == 0))

# Make sure West remains as an empty panel
all_data_col2$Position <- factor(
  all_data_col2$Position,
  levels = c("West", "Top", "East")
)

#remove data before 6pm before plottinge
all_data_col2 <- all_data_col2 %>%
  filter(
    Date_Time >= as.POSIXct("2024-04-19 18:00:00"),
    Date_Time <= as.POSIXct("2024-04-20 15:00:00")
  )

#plot
Plot_Plob_2_hobo <- ggplot(
  all_data_col2,
  aes(x = Date_Time)
) +
  
  # Temperature
  geom_line(
    aes(y = Temp),
    color = temp_color,
    linewidth = 0.8
  ) +
  
  # Light
  geom_line(
    aes(y = Light_scaled),
    color = light_color,
    linewidth = 0.8
  ) +
  
  # Y axes
  scale_y_continuous(
    name = "Temperature [°C]",
    limits = c(30, 35),
    breaks = seq(30, 35, 2),
    sec.axis = sec_axis(
      ~ (. - 30) * 18000,
      name = "Light [Lux]",
      breaks = seq(0, 90000, 10000)
    )
  ) +
  
  # X axis: time only
  scale_x_datetime(
    date_labels = "%H:%M"
  ) +
  
  # Three plots in one row
  facet_wrap(
    ~ Position,
    nrow = 1,
    drop = FALSE
  ) +
  
  labs(
    title = "Temperature and Light data of Colony 2 (iii)",
    x = "Time [h]"
  ) +
  
  theme_classic() +
  
  theme(
    strip.background = element_blank(),
    strip.text = element_text(
      face = "bold",
      size = 12
    ),
    
    axis.title.y.left = element_text(color = temp_color),
    axis.text.y.left  = element_text(color = temp_color),
    
    axis.title.y.right = element_text(color = light_color),
    axis.text.y.right  = element_text(color = light_color),
    
    panel.spacing = unit(1, "lines")
  )

Plot_Plob_2_hobo

# save figure
ggsave(
  filename = "/Users/MarlenSchlotheuber/Nextcloud/10. projects/11. PhD/10. projects/Intraspecific bleaching project/Project planning/MYS_2024/Fieldwork/Data_26042024/HOBO Colony/Manuscript_Figures/HOBO_TempLight_Plob_2_iii.pdf",
  plot = Plot_Plob_2_hobo,
  width = 11.7,      # A4 width in inches
  height = 5,     # A4 height in inches
  units = "in"
)

# colony 3
# upload data
# Col_3_west: HOBO loggers didn't log data
Col_3_west <- read.csv('/Users/MarlenSchlotheuber/Nextcloud/10. projects/11. PhD/10. projects/Intraspecific bleaching project/Project planning/MYS_2024/Fieldwork/Data_26042024/HOBO Colony/20042024_MYS_BB_colony_3/20240420_MY_BB_Plob_3_iv_west_HOBO_TempLight_clean.csv')

# upload data
Col_3_top <- read.csv('/Users/MarlenSchlotheuber/Nextcloud/10. projects/11. PhD/10. projects/Intraspecific bleaching project/Project planning/MYS_2024/Fieldwork/Data_26042024/HOBO Colony/20042024_MYS_BB_colony_3/20240420_MY_BB_Plob_3_iv_top_HOBO_TempLight_clean.csv')

# upload data
Col_3_east <- read.csv('/Users/MarlenSchlotheuber/Nextcloud/10. projects/11. PhD/10. projects/Intraspecific bleaching project/Project planning/MYS_2024/Fieldwork/Data_26042024/HOBO Colony/20042024_MYS_BB_colony_3/20240420_MY_BB_Plob_3_iv_east_HOBO_TempLight_clean.csv')

# Prepare data
Col_3_west$Position <- "West"
Col_3_top$Position  <- "Top"
Col_3_east$Position <- "East"

# Scale Light (0– 00) to Temperature scale (30–40)
Col_3_west$Light_scaled <- 30 + (Col_3_west$Light / 70000) * 5
Col_3_top$Light_scaled  <- 30 + (Col_3_top$Light / 70000) * 5
Col_3_east$Light_scaled <- 30 + (Col_3_east$Light / 70000) * 5

# Convert Date_Time
Col_3_west$Date_Time <- as.POSIXct(
  Col_3_west$Date_Time,
  format = "%m/%d/%y %H:%M"
)

Col_3_top$Date_Time <- as.POSIXct(
  Col_3_top$Date_Time,
  format = "%m/%d/%y %H:%M"
)

Col_3_east$Date_Time <- as.POSIXct(
  Col_3_east$Date_Time,
  format = "%m/%d/%y %H:%M"
)

# Combine Colony 3 data
all_data_col3 <- bind_rows(
  Col_3_west,
  Col_3_top,
  Col_3_east
)

# Remove the 0/0 placeholder from West
all_data_col3 <- all_data_col3 %>%
  filter(!(Position == "West" & Temp == 0 & Light == 0))

# Make sure West remains as an empty panel
all_data_col3$Position <- factor(
  all_data_col3$Position,
  levels = c("West", "Top", "East")
)

#remove data before 6pm before plottinge
all_data_col3 <- all_data_col3 %>%
  filter(
    Date_Time >= as.POSIXct("2024-04-20 18:00:00")
  )


#plot
Plot_Plob_3_hobo <- ggplot(
  all_data_col3,
  aes(x = Date_Time)
) +
  
  # Temperature
  geom_line(
    aes(y = Temp),
    color = temp_color,
    linewidth = 0.8
  ) +
  
  # Light
  geom_line(
    aes(y = Light_scaled),
    color = light_color,
    linewidth = 0.8
  ) +
  
  # Y axes
  scale_y_continuous(
    name = "Temperature [°C]",
    limits = c(30, 35),
    breaks = seq(30, 35, 2),
    sec.axis = sec_axis(
      ~ (. - 30) * 14000,
      name = "Light [Lux]",
      breaks = seq(0, 70000, 10000)
    )
  ) +
  
  # X axis: time only
  scale_x_datetime(
    date_labels = "%H:%M"
  ) +
  
  # Three plots in one row
  facet_wrap(
    ~ Position,
    nrow = 1,
    drop = FALSE
  ) +
  
  labs(
    title = "Temperature and Light data of Colony 3 (iv)",
    x = "Time [h]"
  ) +
  
  theme_classic() +
  
  theme(
    strip.background = element_blank(),
    strip.text = element_text(
      face = "bold",
      size = 12
    ),
    
    axis.title.y.left = element_text(color = temp_color),
    axis.text.y.left  = element_text(color = temp_color),
    
    axis.title.y.right = element_text(color = light_color),
    axis.text.y.right  = element_text(color = light_color),
    
    panel.spacing = unit(1, "lines")
  )

Plot_Plob_3_hobo

# save figure
ggsave(
  filename = "/Users/MarlenSchlotheuber/Nextcloud/10. projects/11. PhD/10. projects/Intraspecific bleaching project/Project planning/MYS_2024/Fieldwork/Data_26042024/HOBO Colony/Manuscript_Figures/HOBO_TempLight_Plob_3_iv.pdf",
  plot = Plot_Plob_3_hobo,
  width = 11.7,      # A4 width in inches
  height = 5,     # A4 height in inches
  units = "in"
)

# colony 5
# upload data
# Col_5_west: HOBO loggers didn't log data
Col_5_west <- read.csv('/Users/MarlenSchlotheuber/Nextcloud/10. projects/11. PhD/10. projects/Intraspecific bleaching project/Project planning/MYS_2024/Fieldwork/Data_26042024/HOBO Colony/22042024_MYS_BBr_colony_5/20240422_ID_BBr_Plob_5_v_west_HOBO_TempLight_clean.csv')

# upload data
Col_5_top <- read.csv('/Users/MarlenSchlotheuber/Nextcloud/10. projects/11. PhD/10. projects/Intraspecific bleaching project/Project planning/MYS_2024/Fieldwork/Data_26042024/HOBO Colony/22042024_MYS_BBr_colony_5/20240422_ID_BBr_Plob_5_v_top_HOBO_TempLight_clean.csv')

# upload data
Col_5_east <- read.csv('/Users/MarlenSchlotheuber/Nextcloud/10. projects/11. PhD/10. projects/Intraspecific bleaching project/Project planning/MYS_2024/Fieldwork/Data_26042024/HOBO Colony/22042024_MYS_BBr_colony_5/20240422_ID_BBr_Plob_5_v_east_HOBO_TempLight_clean.csv')

# Prepare data
Col_5_west$Position <- "West"
Col_5_top$Position  <- "Top"
Col_5_east$Position <- "East"

# Scale Light (0– 00) to Temperature scale (30–40)
Col_5_west$Light_scaled <- 30 + (Col_5_west$Light / 50000) * 10
Col_5_top$Light_scaled  <- 30 + (Col_5_top$Light / 50000) * 10
Col_5_east$Light_scaled <- 30 + (Col_5_east$Light / 50000) * 10

# Convert Date_Time
Col_5_west$Date_Time <- as.POSIXct(
  Col_5_west$Date_Time,
  format = "%m/%d/%y %H:%M"
)

Col_5_top$Date_Time <- as.POSIXct(
  Col_5_top$Date_Time,
  format = "%m/%d/%y %H:%M"
)

Col_5_east$Date_Time <- as.POSIXct(
  Col_5_east$Date_Time,
  format = "%m/%d/%y %H:%M"
)

# Combine Colony 5 data
all_data_col5 <- bind_rows(
  Col_5_west,
  Col_5_top,
  Col_5_east
)

# Make sure West remains as an empty panel
all_data_col5$Position <- factor(
  all_data_col5$Position,
  levels = c("West", "Top", "East")
)

#remove data before 6pm before plottinge
all_data_col5_filtered <- all_data_col5 %>%
  filter(
    Date_Time >= as.POSIXct("2024-04-22 18:00:00"),
    Date_Time <= as.POSIXct("2024-04-23 18:00:00")
  )

#plot
Plot_Plob_5_hobo <- ggplot(
  all_data_col5_filtered,
  aes(x = Date_Time)
) +
  
  # Temperature
  geom_line(
    aes(y = Temp),
    color = temp_color,
    linewidth = 0.8
  ) +
  
  # Light
  geom_line(
    aes(y = Light_scaled),
    color = light_color,
    linewidth = 0.8
  ) +
  
  # Y axes
  scale_y_continuous(
    name = "Temperature [°C]",
    limits = c(30, 35),
    breaks = seq(30, 35, 2),
    sec.axis = sec_axis(
      ~ (. - 30) * 10000,
      name = "Light [Lux]",
      breaks = seq(0, 50000, 10000)
    )
  ) +
  
  # X axis: time only
  scale_x_datetime(
    date_labels = "%H:%M"
  ) +
  
  # Three plots in one row
  facet_wrap(
    ~ Position,
    nrow = 1,
    drop = FALSE
  ) +
  
  labs(
    title = "Temperature and Light data of Colony 5 (v)",
    x = "Time [h]"
  ) +
  
  theme_classic() +
  
  theme(
    strip.background = element_blank(),
    strip.text = element_text(
      face = "bold",
      size = 12
    ),
    
    axis.title.y.left = element_text(color = temp_color),
    axis.text.y.left  = element_text(color = temp_color),
    
    axis.title.y.right = element_text(color = light_color),
    axis.text.y.right  = element_text(color = light_color),
    
    panel.spacing = unit(1, "lines")
  )

Plot_Plob_5_hobo

# save figure
ggsave(
  filename = "/Users/MarlenSchlotheuber/Nextcloud/10. projects/11. PhD/10. projects/Intraspecific bleaching project/Project planning/MYS_2024/Fieldwork/Data_26042024/HOBO Colony/Manuscript_Figures/HOBO_TempLight_Plob_5_v.pdf",
  plot = Plot_Plob_5_hobo,
  width = 11.7,      # A4 width in inches
  height = 5,     # A4 height in inches
  units = "in"
)

# colony 6
# upload data
# Col_6_west: HOBO loggers didn't log data
Col_6_west <- read.csv('/Users/MarlenSchlotheuber/Nextcloud/10. projects/11. PhD/10. projects/Intraspecific bleaching project/Project planning/MYS_2024/Fieldwork/Data_26042024/HOBO Colony/22042024_MYS_BBr_colony_5/20240424_ID_BBr_Plob_6_ii_west_HOBO_TempLight_clean.csv')

# upload data
Col_6_top <- read.csv('/Users/MarlenSchlotheuber/Nextcloud/10. projects/11. PhD/10. projects/Intraspecific bleaching project/Project planning/MYS_2024/Fieldwork/Data_26042024/HOBO Colony/22042024_MYS_BBr_colony_5/20240424_ID_BBr_Plob_6_ii_top_HOBO_TempLight_clean.csv')

# upload data
Col_6_east <- read.csv('/Users/MarlenSchlotheuber/Nextcloud/10. projects/11. PhD/10. projects/Intraspecific bleaching project/Project planning/MYS_2024/Fieldwork/Data_26042024/HOBO Colony/22042024_MYS_BBr_colony_5/20240424_ID_BBr_Plob_6_ii_east_HOBO_TempLight_clean.csv')

# Prepare data
Col_6_west$Position <- "West"
Col_6_top$Position  <- "Top"
Col_6_east$Position <- "East"

# Scale Light (0– 00) to Temperature scale (30–40)
Col_6_west$Light_scaled <- 30 + (Col_6_west$Light / 90000) * 5
Col_6_top$Light_scaled  <- 30 + (Col_6_top$Light / 90000) * 5
Col_6_east$Light_scaled <- 30 + (Col_6_east$Light / 90000) * 5

# Convert Date_Time
Col_6_west$Date_Time <- as.POSIXct(
  Col_6_west$Date_Time,
  format = "%m/%d/%y %H:%M"
)

Col_6_top$Date_Time <- as.POSIXct(
  Col_6_top$Date_Time,
  format = "%m/%d/%y %H:%M"
)

Col_6_east$Date_Time <- as.POSIXct(
  Col_6_east$Date_Time,
  format = "%m/%d/%y %H:%M"
)

# Combine Colony 5 data
all_data_col6 <- bind_rows(
  Col_6_west,
  Col_6_top,
  Col_6_east
)

# Make order
all_data_col6$Position <- factor(
  all_data_col6$Position,
  levels = c("West", "Top", "East")
)

#plot
Plot_Plob_6_hobo <- ggplot(
  all_data_col6,
  aes(x = Date_Time)
) +
  
  # Temperature
  geom_line(
    aes(y = Temp),
    color = temp_color,
    linewidth = 0.8
  ) +
  
  # Light
  geom_line(
    aes(y = Light_scaled),
    color = light_color,
    linewidth = 0.8
  ) +
  
  # Y axes
  scale_y_continuous(
    name = "Temperature [°C]",
    limits = c(30, 35),
    breaks = seq(30, 35, 2),
    sec.axis = sec_axis(
      ~ (. - 30) * 18000,
      name = "Light [Lux]",
      breaks = seq(0, 90000, 10000)
    )
  ) +
  
  # X axis: time only
  scale_x_datetime(
    date_labels = "%H:%M"
  ) +
  
  # Three plots in one row
  facet_wrap(
    ~ Position,
    nrow = 1,
    drop = FALSE
  ) +
  
  labs(
    title = "Temperature and Light data of Colony 6 (ii)",
    x = "Time [h]"
  ) +
  
  theme_classic() +
  
  theme(
    strip.background = element_blank(),
    strip.text = element_text(
      face = "bold",
      size = 12
    ),
    
    axis.title.y.left = element_text(color = temp_color),
    axis.text.y.left  = element_text(color = temp_color),
    
    axis.title.y.right = element_text(color = light_color),
    axis.text.y.right  = element_text(color = light_color),
    
    panel.spacing = unit(1, "lines")
  )

Plot_Plob_6_hobo

# save figure
ggsave(
  filename = "/Users/MarlenSchlotheuber/Nextcloud/10. projects/11. PhD/10. projects/Intraspecific bleaching project/Project planning/MYS_2024/Fieldwork/Data_26042024/HOBO Colony/Manuscript_Figures/HOBO_TempLight_Plob_6_ii.pdf",
  plot = Plot_Plob_6_hobo,
  width = 11.7,      # A4 width in inches
  height = 5,     # A4 height in inches
  units = "in"
)
```
