---
title: "Exploring Within-Colony Thermal Tolerance Variability in Reef Building Corals"
author: "Marlen Schlotheuber"
format: html
editor: visual
---

# R code for Manuscript: Exploring Within-Colony Thermal Tolerance Variability in Ree-Building Corals

### Install needed packages: CBASS package

```{r}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
if(!require(devtools)){
   install.packages("devtools")
}

devtools::install_github("reefgenomics/CBASSED50@0.1.5.1", force=TRUE)
BiocManager::install("dada2")
install.packages("ggmap")
install.packages("osmdata")
install.packages("ggspatial")
BiocManager::install("dada2", force = TRUE)
install.packages(c("cubature", "ape", "reshape2", "edgeR", "plyr"))
install.packages("Rcpp")
install.packages("seqinr")
BiocManager::install("Biostrings")
BiocManager::install("phyloseq")
install.packages("seqinr")

library(tidyverse)
library(seqinr)
library(phyloseq)
library(Biostrings)
library(reshape2)
library(scales)
library(gridExtra)
library(dada2)
library(Rcpp)
library(sf)
library(rnaturalearth)
library(ggmap)
library(readxl)
library(rstudioapi)
library(RColorBrewer)
library(CBASSED50)
library(patchwork)
library(car)
library(mgcv)
library(stringr)
library(lme4)
library(lmerTest) 
library(emmeans)
library(patchwork)
```

## **Fig.1 Study regions and the sampling locations within massive Porites colonies**

```{r}
# Study 
# Get country boundaries
world <- ne_countries(scale = "medium", returnclass = "sf")

# Subset for Indonesia and Malaysia
countries <- world[world$name %in% c("Indonesia", "Malaysia"), ]

# plot
Fig1A_map <-  ggplot() +
  geom_sf(data = world, fill = "white", color = "black") +
  geom_sf(data = subset(countries, name == "Indonesia"), fill = "white", color = "black") +
  geom_sf(data = subset(countries, name == "Malaysia"), fill = "white", color = "black") +
  ggtitle("Sampling sites: Indonesia and Malaysia") +
  coord_sf(xlim = c(90, 155), ylim = c(-35, 35), expand = FALSE) +
  scale_x_continuous(name = "Longitude (°E)", breaks = seq(90, 155, 10)) +
  scale_y_continuous(name = "Latitude (°N)", breaks = seq(-35, 35, 10)) +
  theme_minimal() +
  theme(
    panel.background = element_rect(fill = "grey90", color = NA),  # ocean
    plot.background  = element_rect(fill = "white", color = NA),   # outer plot bg
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank()
  )

Fig1A_map
```

# Thermal tolerance data from Indonesia and Malaysia

To examine the photosynthetic efficiency data from Malaysia and Indonesia, the R-package from Iakovleva, Colin & Voolstra, 2025 was used ([https://doi.org/10.5281/zenodo.8370644](https://doi.org/10.5281/zenodo.8370644.)).

The photosynthetic efficiency data from the four temperature profiles of each sampling location from Indonesia and Malysia was used to fit log-logistic dose-response curves, using the ‘drc’ R package. These curves were used to determine the effective dose 50 (ED50) for each sampling location (Evensen et al. 2021; Voolstra et al. 2021). The ED50 value defines the temperature at which photosynthetic efficiency has declined by 50% relative to the baseline measurement, and is therefore used as an empirically derived proxy for heat stress tolerance (Evensen et al. 2022). 

### Load Data ED50 data

```{r}
# Indonesia
# Upload Indonesia data: 202408_ID_GA_Por_CBASS_420_ED50.xlsx
input_data_path <- selectFile(
  caption = "Select XLSX or CSV Input File")

# ID
cbass_dataset_ID <- read_data(input_data_path)

# To specify the prefix for output files
output_prefix <- tools::file_path_sans_ext(input_data_path)

rlog::log_info(paste("Your current directory is", getwd()))
rlog::log_info(paste("Your input filename is", basename(input_data_path)))
rlog::log_info(paste("The output files will be written into", output_prefix))

# Malaysia data
# upload Malaysia data: 202404_MY_PLT_Por_CBASS_420_ED50.xlsx
input_data_path <- selectFile(
  caption = "Select XLSX or CSV Input File")

# MYS
cbass_dataset_MY <- read_data(input_data_path)

# To specify the prefix for output files
output_prefix <- tools::file_path_sans_ext(input_data_path)

rlog::log_info(paste("Your current directory is", getwd()))
rlog::log_info(paste("Your input filename is", basename(input_data_path)))
rlog::log_info(paste("The output files will be written into", output_prefix))

```

### Analysis and Preprocessing

```{r}
# ID
cbass_dataset_ID <- preprocess_dataset(cbass_dataset_ID)
validate_cbass_dataset(cbass_dataset_ID)

# MY
cbass_dataset_MY <- preprocess_dataset(cbass_dataset_MY)
validate_cbass_dataset(cbass_dataset_MY)
```

### Explore ED50

```{r}
# ID
# grouping properties Indonesia
grouping_properties <- c("Condition")

# PAM data Indonesia
drm_formula <- "Pam_value ~ Temperature"
models <- fit_drms(cbass_dataset_ID, grouping_properties, drm_formula, is_curveid = TRUE)
str(cbass_dataset_ID)

# PAM data 
ed50_ID <- get_ed50_by_grouping_property(models)
cbass_dataset_ID <- define_grouping_property(cbass_dataset_ID, grouping_properties) %>%
  mutate(GroupingProperty = paste(GroupingProperty, Genotype, sep = "_"))

ed50_df_ID <- 
  left_join(ed50_ID, cbass_dataset_ID, by = "GroupingProperty") %>%
  dplyr::select(names(ed50_ID), Depth, Genotype, all_of(grouping_properties)) %>%
  distinct()


# MY
# grouping properties
grouping_properties <- c("Orientation", "Condition")

# PAM data Malaysia
drm_formula <- "Pam_value ~ Temperature"
models <- fit_drms(cbass_dataset_MY, grouping_properties, drm_formula, is_curveid = TRUE)

# PAM data
ed50_MY <- get_ed50_by_grouping_property(models)
cbass_dataset_MY <- define_grouping_property(cbass_dataset_MY, grouping_properties) %>%
  mutate(GroupingProperty = paste(GroupingProperty, Genotype, sep = "_"))

ed50_df_MY <- 
  left_join(ed50_MY, cbass_dataset_MY, by = "GroupingProperty") %>%
  dplyr::select(names(ed50_MY), Depth, Genotype, all_of(grouping_properties)) %>%
  distinct()

```

### Adjust dataset for plotting

```{r}
# ID
cbass_dataset_ID$Genotype <- as.numeric(cbass_dataset_ID$Genotype)
cbass_dataset_ID$Condition <- as.factor(cbass_dataset_ID$Condition)
cbass_dataset_ID$Temperature <- as.factor(cbass_dataset_ID$Temperature)

# change genotype numbers to numeric so its the same as for the MYS dataset
str(cbass_dataset_MY)
cbass_dataset_MY$Genotype <- as.numeric(cbass_dataset_MY$Genotype)
cbass_dataset_MY$Condition <- as.factor(cbass_dataset_MY$Condition)
```

### Temperature response curve

```{r}
# ID
# set order of the Locations
cbass_dataset_ID$Condition <- factor(cbass_dataset_ID$Condition, 
                                   levels = c("Top", "Bottom"))

# visualize temperature response curve 
exploratory_curve_ID <-
  ggplot(data = cbass_dataset_ID,
       aes(
         x = Temperature,
         y = Pam_value,
         # You can play around with the group value (e.g., Species, Site, Condition)
         group = GroupingProperty,
         color = Condition)) +
  geom_smooth(
    method = drc::drm,
    method.args = list(
      fct = drc::LL.3()),
    se = FALSE,
    size = 0.7
  ) +
  geom_point(size = 1.5) +
  facet_wrap( ~ Genotype, nrow = 2) +
  scale_color_brewer(palette = "Paired") # Colorblind-friendly palette

#plot
exploratory_curve_ID


# MYS
# set order of the Locations
cbass_dataset_MY$Condition <- factor(cbass_dataset_MY$Condition, 
                                   levels = c("Top", "Upper Middle", "Lower Middle", "Bottom"))

# order genotypes
cbass_dataset_MY$Genotype <- factor(cbass_dataset_MY$Genotype, 
                                   levels = c("1", "6", "2", "3", "5", "4"))


# visualize temperature response curve
exploratory_curve_MY <-
  ggplot(data = cbass_dataset_MY,
       aes(
         x = Temperature,
         y = Pam_value,
         # You can play around with the group value (e.g., Species, Site, Condition)
         group = GroupingProperty,
         color = Condition)) +
  geom_smooth(
    method = drc::drm,
    method.args = list(
      fct = drc::LL.3()),
    se = FALSE,
    size = 0.7
  ) +
  geom_point(size = 1.5) +
  facet_wrap( ~ Genotype, nrow = 2) +
  scale_color_brewer(palette = "Paired") # Colorblind-friendly palette

# plot
exploratory_curve_MY
```

## **Fig 2. Thermal tolerance across the surface of massive Porites sp. colonies from Indonesia and Malaysia.**

```{r}
# Indonesia
# only select needed columns from the indonesia dataset
ed50_df_ID_filtered <- ed50_df_ID %>%
  dplyr::select(c("ED50", "Depth", "Condition", "Genotype"))

# rename Genotypes from 1-6 because colony 5 was removed due to missing ist2 and 16s data, currently the genotypes are 1,2,3,4,6,7 so rename: 6 and 7 to 5 and 6

# convert Genotype to numeric
ed50_df_ID_filtered$Genotype <- as.numeric(as.character(ed50_df_ID_filtered$Genotype))

ed50_df_ID_filtered <- ed50_df_ID_filtered %>%
  mutate(Genotype = case_when(
    Genotype == 6 ~ 5,
    Genotype == 7 ~ 6,
    TRUE ~ Genotype  # keep all other values unchanged
  ))

# calculate mean ED50 per colony/genotype and calculate differences
ED50_ID_means <- ed50_df_ID_filtered %>%
  group_by(Genotype) %>%
  summarise(mean_ED50 = mean(ED50))

# create final dataframe with mean ED50
Final_ED50_ID <- ed50_df_ID_filtered %>%
  left_join(ED50_ID_means, by = c("Genotype")) %>%
  mutate(ED50_diff = ED50 - mean_ED50) %>%
  select(-contains(".x")) %>% # remove duplicates
  distinct()  # Re

# prepare for plotting
# set the order of the locations
Final_ED50_ID$Condition <- factor(Final_ED50_ID$Condition, 
                                   levels = c("Bottom", "Top"))

# set the order of the coral colonies
Final_ED50_ID <- Final_ED50_ID %>%
  mutate(
    Genotype = factor(Genotype, levels = c("1", "2", "3", "4", "5", "6"))  # Order genotypes
  )

 # set the colors               
genotype_colors <- c("1" = "skyblue3", 
                     "2" = "lightblue", 
                     "3" = "dodgerblue4", 
                     "4" = "cornflowerblue", 
                     "5" = "lightskyblue", 
                     "6" = "dodgerblue3")


# ED50 Difference Bar Plot
ED50_diff_barplot_ID <- ggplot(Final_ED50_ID, aes(x = ED50_diff, y = Condition, fill = Genotype)) +
  geom_vline(aes(xintercept = 0), color = "black", linetype = "dashed", size = 1) +
  geom_vline(xintercept = -0.5, color = "grey40", linetype = "dashed") +
  geom_vline(xintercept = 0.5, color = "grey40", linetype = "dashed") +
  geom_bar(stat = "identity", position = "dodge") +
  facet_wrap(~Genotype, ncol = 1) +
  xlim(-1, 1) +
  scale_y_discrete(labels = c("Bottom", "Top")) +
  scale_fill_manual(values = genotype_colors) +
  labs(
    x = "Deviation from Mean ED50 [°C]", 
    y = "Location within Colony",
    fill = "Genotype"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    text = element_text(size = 13),
    strip.text = element_blank(),
    axis.title = element_text(size = 13, face = "bold"),
    axis.text = element_text(size = 13),
    legend.position = "none",
    panel.spacing = unit(0.3, "lines"),
    strip.background = element_blank() )+
      # Add mean ED50 above each facet
  geom_text(
    data = ED50_ID_means,
    aes(
      x = 0,              
      y = 2.5,              
      label = paste0("Mean ED50: ", round(mean_ED50, 2))
    ),
    inherit.aes = FALSE,
    size = 3,
    fontface = "bold" )


ED50_diff_barplot_ID


# Malayisa
# only select needed columns from the indonesia dataset
ed50_df_MY_filtered <- ed50_df_MY %>%
  dplyr::select(c("ED50", "Depth", "Orientation", "Condition", "Genotype"))

# calculate mean ED50 per colony/genotype and calculate differences
ED50_MY_means <- ed50_df_MY_filtered %>%
  group_by(Genotype) %>%
  summarise(mean_ED50 = mean(ED50))

# create final dataframe with mean ED50
Final_ED50_MY <- ed50_df_MY_filtered %>%
  left_join(ED50_MY_means, by = c("Genotype")) %>%
  mutate(ED50_diff = ED50 - mean_ED50) %>%
  select(-contains(".x")) %>% # remove duplicates
  distinct()

# set order of the Locations
Final_ED50_MY$Condition <- factor(Final_ED50_MY$Condition, 
                                   levels = c("Bottom", "Lower Middle", "Upper Middle", "Top"))
 
# set order of the colonies with increasing depth
Final_ED50_MY <- Final_ED50_MY %>%
  mutate(
    Genotype = factor(Genotype, levels = c("1", "6", "2", "3", "5", "4")))

# adjust Orientation column
Final_ED50_MY <- Final_ED50_MY %>%
  mutate(
    Orientation = factor(Orientation, levels = c("West", "East")))


# set the colors    
genotype_colors <- c("1" = "pink4", 
                     "2" = "orchid4", 
                     "3" = "pink2", 
                     "4" = "mistyrose2", 
                     "5" = "lightcoral",
                     "6" = "hotpink4")


# Bar plot
ED50_diff_barplot_MY <- ggplot(Final_ED50_MY, aes(x = ED50_diff, y = Condition, fill = Genotype)) +
  geom_vline(aes(xintercept = 0), color = "black", linetype = "dashed", size = 1) +
  geom_vline(xintercept = -0.5, color = "grey40", linetype = "dashed") +
  geom_vline(xintercept = 0.5, color = "grey40", linetype = "dashed") +
  geom_bar(stat = "identity", position = "dodge") +
  facet_grid(Genotype ~ Orientation, scales = "free_y") +
  xlim(-1, 1) +
  scale_y_discrete(labels = c("Bottom", "Lower Middle", "Upper Middle", "Top")) +
  scale_fill_manual(values = genotype_colors) +
  labs(
    x = "Deviation from Mean ED50 [°C]", 
    y = "Location within Colony",
    fill = "Genotype"
  ) +
  theme_minimal(base_size = 13) +  # Global text size
  theme(
    text = element_text(size = 13),
    strip.text = element_blank(),
    axis.title = element_text(size = 13, face = "bold"),  
    axis.text = element_text(size = 13),  
    legend.position = "none",
    panel.spacing = unit(0.4, "lines"),
    strip.background = element_blank() )+
      
      # Add mean ED50 above each facet
  geom_text(
    data = ED50_MY_means,
    aes(
      x = 0,              
      y = 4.2,              
      label = paste0("Mean ED50: ", round(mean_ED50, 2))
    ),
    inherit.aes = FALSE,
    size = 3,
    fontface = "bold" )

# plot Malaysia
ED50_diff_barplot_MY

# Final Figure. 2
# Plot both plots
ED50_diff_barplot_ID + ED50_diff_barplot_MY

# Figure was further adjusted in Affinity designer
```

###  Testing for significant differences in thermal tolerance

```{r}
# Indonesia
# check for significant differences

# data distribution
hist(Final_ED50_ID$ED50) 
shapiro.test(Final_ED50_ID$ED50) # p-value: not signfiicant, so data is normally distributed
qqnorm(Final_ED50_ID$ED50) # normal distributed, looks good! 

# within colonies
# check for differences in variance between the top and bottom sampling location
leveneTest(ED50 ~ Condition, data = Final_ED50_ID)
# p-value = not significant difference in the variance of groups

# statistically testing for significant differences
# paired t.test since 2 measurements were takin within the same colony, so colony is our pairing unit, controlls for between colony variability and tests within colony

# Supplementary Table 5
paired_ED50 <- Final_ED50_ID %>%
  filter(Condition %in% c("Top", "Bottom")) %>%
  select(Genotype, Condition, ED50) %>%
  pivot_wider(names_from = Condition,
              values_from = ED50)

t.test(paired_ED50$Top,
       paired_ED50$Bottom,
       paired = TRUE)
# p-value = 0.8047

# Supplementary Table 6
# testing for significant differences across colonies
anova <- aov(ED50 ~ Genotype, data = Final_ED50_ID)
summary(anova)

# Supplementary Table 7
# testing for significant differences across depth
anova <- aov(ED50 ~ Depth, data = Final_ED50_ID)
summary(anova)


# Malaysia
# check data distribution: normality and homogeneity
hist(Final_ED50_MY$ED50) 
# normallity test
shapiro.test(Final_ED50_MY$ED50) # p-value: not signifiacnt, data normally distributed 
# plot normallity
qqnorm(Final_ED50_MY$ED50) # looks normally distributed

# Supplementary Table 8
# test homogeneity
leveneTest(ED50 ~ Condition, data = Final_ED50_MY)
# p-value:not significant use ANOVA

# test for significance using ANOVA (4 locations within colonies)
# using a linear mixed model to account for nested design
model <- lmer(ED50 ~ Condition + (1 | Genotype), data = Final_ED50_MY)
anova(model)

# Supplementary Table 9
# check for East
Final_ED50_East <- subset(Final_ED50_MY, Orientation == "East")

# using a linear mixed model to account for nested design
model <- lmer(ED50 ~ Condition + (1 | Genotype), data = Final_ED50_East)
anova(model)
# p-value= not signfificant

# Supplementary Table 10
# Check for West
Final_ED50_West <- subset(Final_ED50_MY, Orientation == "West")

# using a linear mixed model to account for nested design
model <- lmer(ED50 ~ Condition + (1 | Genotype), data = Final_ED50_West)
anova(model)
# p-value= not signfificant

# Supplementary Tabel 11
# is ED50 significantly different across colonies?
anova <- aov(ED50 ~ Genotype, data = Final_ED50_MY)
summary(anova)

# Supplementary Table 12
# Pearson to check for correlation between ed50 and depth
Final_ED50_MY$Depth <- as.numeric(as.character(Final_ED50_MY$Depth))

cor.test(Final_ED50_MY$ED50, Final_ED50_MY$Depth, method = "pearson")
# is ED50 significantly different across colonies?
anova <- aov(ED50 ~ Genotype, data = Final_ED50_MY)
summary(anova) # p-value = significant

# Supplemenatry Table  13
# is depth significantly affecting ed50
anova <- aov(ED50 ~ Depth, data = Final_ED50_MY)
summary(anova)
# p-value: significant


```

# ITS2 type profiles - Symbiodiniaceae community composition

ITS2 sequencing data from colonies collected in Indonesia and Malaysia were seperatly analysed using SymPortal (Hume et al., 2019).\
ITS2 type profiles being representative of putative Symbiodiniaceae genotypes were characterized by specific sets of defining intragenomic ITS2 sequence variants (DIVs) (Hume et al. 2019) and visualized using R studio.

### Indonesia: Load Data and Preprocessing

```{r}
# ITS2 type profile metadata
input_data_path <- selectFile(
  caption = "Select XLSX or CSV Input File")

# ITS2 profile data
its2_metadata_ID <- read_data(input_data_path)


#upload ITS2 profile absolute abundance
its2_profile_ID <- read.table('/yourpathname/its2_type_profiles/582_20250509T135802_DBV_20250510T135051.profiles.relative.abund_and_meta.txt', header = TRUE, sep = "\t", stringsAsFactors = FALSE)

colnames(its2_profile)[1] <- "ITS2_profile"
colnames(its2_profile_ID)[2] <- "sample_name"

# merge dataframes
its2_data_ID <- merge(its2_metadata_ID, its2_profile_ID, by = "sample_name", all.x = TRUE)

# prepare for plotting
# Convert to long format for ggplot
its2_profile_ID_long <- its2_profile_ID %>%
  gather(key = "ITS2_profile", value = "value", C15.C15aae:C15.C15ad.C15ai.C116.C116aa)

# split colunm into seperate columns to extract the inforamtion
# Ensure sample_name is a character vector
its2_profile_ID_long$sample_name <- as.character(its2_profile_ID_long$sample_name)

# Split the sample_naem column and extract the components
split_info <- do.call(rbind, strsplit(its2_profile_ID_long$sample_name, "_"))

# Assign the components to new columns
its2_profile_ID_long$Coral_species <- split_info[, 1]
its2_profile_ID_long$Genotype <- split_info[, 2]
its2_profile_ID_long$Location <- split_info[, 3]

its2_data_ID_final <- its2_profile_ID_long

# Ensure order
its2_profile_ID_long <- its2_profile_ID_long %>%
  mutate(
    collection_location = factor(Location, levels = c("Bottom", "Top")))
```

### Indonesia: Plot ITS2 Type profiles

```{r}
ITS2_plot_ID <- ggplot(its2_profile_ID_long, aes(x = value, y = Location, fill = ITS2_profile)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_x_continuous(labels = scales::percent_format(), expand = c(0, 0)) +
  facet_grid(Genotype ~ ., scales = "free_y") +
  theme_minimal(base_size = 13) +
    guides(fill = guide_legend(ncol = 1)) +
  labs(
    x = "Relative Abundance",
    y = "Location within Colony",
    fill = "ITS2 Type Profiles"
  ) +
  scale_fill_manual(
    values = c(
      "C15.C15aae" = '#006633',
      "C15.C15bn.C15by" = "#99CC33",
      "C15.C116aa.C15ad" = "#339900",
      "C57d.C57a" = "#000099",
      "C15.C15kl.C15he" = "#009966",
      "C15.C15ad.C15ai.C116.C116aa" = "#6699CC"
    ),
    labels = c(
      "C15.C15aae" = "C15-C15aae",
      "C15.C15bn.C15by" = "C15-C15bn-C15by",
      "C15.C116aa.C15ad" = "C15/C116aa-C15ad",
      "C57d.C57a" = "C57d/C57a",
      "C15.C15kl.C15he" = "C15-C15kl-C15he",
      "C15.C15ad.C15ai.C116.C116aa" = "C15-C15ad-C15ai-C116-C116aa"
    )
  ) +
  theme(
    legend.position = "bottom",
    legend.box = "horizontal",
    legend.text = element_text(size = 13),
    legend.title = element_text(size = 13),
    axis.text = element_text(size = 13),
    axis.title = element_text(size = 13),
    strip.text = element_text(size = 13),
    panel.grid = element_blank(),
    panel.border = element_blank()
  )

ITS2_plot_ID

```

### Malaysia: Load Data and Preprocessing

```{r}
# upload ITS2 type profile meta data
input_data_path <- selectFile(
  caption = "Select XLSX or CSV Input File")

its2_metadata <- read_data(input_data_path)

#upload ITS2 profile absolute abundance
its2_profile <- read.table('/Users/MarlenSchlotheuber/Nextcloud/10. projects/11. PhD/10. projects/Intraspecific bleaching project/Project planning/MYS_2024/Fieldwork/Data_26042024/SymPortal/Sequencing_data_ITS2_SymPortal/ITS2_SymPortal_data/202404_MYS_ITS2_data/its2_type_profiles/560_20250219T125105_DBV_20250221T224139.profiles.relative.abund_and_meta.txt', header = TRUE, sep = "\t", stringsAsFactors = FALSE)

colnames(its2_profile)[1] <- "ITS2_profile"
colnames(its2_profile)[2] <- "sample_name"

# merge dataframes
its2_data<- merge(its2_metadata, its2_profile, by = "sample_name", all.x = TRUE)

# prepare for plotting
# Convert to long format for ggplot
its2_data_long <- its2_data %>%
  gather(key = "ITS2_profile", value = "value", C15.C15ev.C15dt.C15kl.C15pm:C15.C15by.C15ai)

# just select data: top, upper_middle, upper_middle_1, lower_middle, lower_middle_1, bottom
# Colony 2 was so large that we were able to sample at 6 sampling locations within the colonies (top, upper middle 1, upper middle 2, lower middle 1, lower middle 2, bottom) but since no differences within the colony were detected we removed the two additional middle sampling locations upper and lower middle 2. Here we select the top, upper_middle, lower_middle and bottom als well as upper_middle_1 and lower_middle_1 to select all needed data points.

its2_data_long_filtered <- its2_data_long %>%
  filter(collection_depthrange %in% c("top", "upper_middle", "upper_middle_1", "lower_middle","lower_middle_1", "bottom"))

# rename upper and lower middle 1 of colony 2 to just lower and upper middle
its2_data_long_filtered <- its2_data_long_filtered %>%
  mutate(collection_depthrange = dplyr::recode(
    collection_depthrange,
    lower_middle_1 = "lower_middle",
    upper_middle_1 = "upper_middle"
  ))

# First, ensure your collection_depthrange is a factor with appropriate labels.
its2_data_long_filtered$collection_depthrange <- factor(its2_data_long_filtered$collection_depthrange,
                                              levels = c(4, 3, 2, 1),
                                              labels = c("bottom", "lower_middle", "upper_middle", "top"))

# change the naming in collection_depthrange
# rename upper and lower middle 1 to just lower and upper midlee for colony 2
its2_data_long_filtered <- its2_data_long_filtered %>%
  mutate(collection_depthrange = dplyr::recode(collection_depthrange,
                                      lower_middle = "Lower Middle",
                                      upper_middle = "Upper Middle",
                                      top = "Top",
                                      bottom = "Bottom"))

# Ensure 'collection_depthrange' is ordered and 'collection_orientation' is a factor with "West" and "East"
its2_data_long_filtered <- its2_data_long_filtered %>%
  mutate(
    collection_depthrange = factor(collection_depthrange, levels = c("Bottom", "Lower Middle","Upper Middle", "Top")),
    collection_orientation = factor(collection_orientation, levels = c("West", "East"))
  )

```

### Plot ITS2 Type Profile

```{r}
# organize data
# set order of the colonies with decreasing depth
its2_profile_MY_long_filtered <- its2_data_long_filtered %>%
  mutate(
    collection_colony = factor(collection_colony, levels = c("1", "6", "2", "3", "5", "4")))

str(its2_profile_MY_long_filtered)

#plot
ITS2_plot_MYS <- ggplot(its2_profile_MY_long_filtered, aes(x = value, y = collection_depthrange, fill = ITS2_profile)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_x_continuous(labels = scales::percent_format(), expand = c(0, 0)) +
  facet_grid(collection_colony ~ collection_orientation, scales = "free_y") +
  theme_minimal(base_size = 13) +
  guides(fill = guide_legend(ncol = 1)) +
  labs(
    x = "Relative Abundance",
    y = "Location within Colony",
    fill = "ITS2 Type Profiles"
  ) +
  scale_fill_manual(
    values = c(
      "C15.C15ev.C15dt.C15kl.C15pm" = "#66CC66",
         "C15.C15ad.C15ai.C116.C116aa" = "#6699CC",
      "C15.C15ev.C15by" = "#66CC99",
      "C15.C15ev.C15dt" = "#009999",
      "C15.C15ev" = "#99CCCC",
        "C15.C15by.C15ai" = "#006666"
    ),
    labels = c(
      "C15.C15ev.C15dt.C15kl.C15pm" = "C15-C15ev-C15dt-C15kl-C15pm",
      "C15.C15ad.C15ai.C116.C116aa" = "C15-C15ad-C15ai-C116-C116aa",
      "C15.C15ev.C15by" = "C15-C15ev-C15by",
      "C15.C15ev.C15dt" = "C15-C15ev-C15dt",
      "C15.C15ev" = "C15-C15ev",
      "C15.C15by.C15ai" = "C15-C15by-C15ai"
    )
  ) +
  theme(
    legend.position = "bottom",
    legend.box = "horizontal",
    legend.text = element_text(size = 13),
    legend.title = element_text(size = 13),
    axis.text = element_text(size = 13),
    axis.title = element_text(size = 13),
    strip.text = element_text(size = 13),
    panel.grid = element_blank(),
    panel.border = element_blank()
  )

ITS2_plot_MYS

```

## **Fig. 3 ITS2 type profiles across the surface of massive Porites sp. colonies from Indonesia and Malaysia.**

```{r}
# Combine plots side-by-side with equal width
final_combined_plot <- ITS2_plot_ID + ITS2_plot_MYS +
  plot_layout(ncol = 2, widths = c(1, 1)) +  # Equal widths for both plots
  plot_annotation(
    theme = theme(
      plot.margin = margin(10, 10, 10, 10)
    )
  )
final_combined_plot

# plot as saved and edited with Affinity designer.
# Save to A4 size (portrait orientation) as PDF
ggsave(
  filename = "ITS2_MYS_ID_new.pdf",
  plot = final_combined_plot,
  width = 11.7,      # A4 width in inches
  height = 10,     # A4 height in inches
  units = "in"
)
```

# 16S - bacterial community composition

16S sequencing data was processed in R studio using the DADA2 pipeline. DADA2 (Callahan et al., 2016) filtered and trimmed reads, calculated error rates, inferred sequence variants, merged paired reads, and removed chimeras. Taxonomy was assigned by using the Silva v138 dataset from the Silva reference database (Benjamin, 2024). The top 20 bacterial families across the colony surface were visualised using R studio. <https://benjjneb.github.io/dada2/tutorial.html>

### DADA2

```{r}
# Save as FASTA file
write.fasta(sequences = as.list(raw_fasta),
            names = names(raw_fasta),
            file.out = "top_ASVs.fasta")

# set working directory
# read file
cat("Reading files") # prints a message to indicate the script is starting to read files
path <- "." # current working directory
list.files(path) # lists all files in the current directory  --> make sure you are in the correct directory so the 16S files can be uploaded!


# Forward and reverse fastq filenames have format: 
fnFs <- sort(list.files(path, pattern="_R1_001.fastq", full.names = TRUE))
fnRs <- sort(list.files(path, pattern="_R2_001.fastq", full.names = TRUE))
# searches for files matching specific patterns _R1_001...and sorts files, get a sorted list of file paths for forward (fnFs) and reverse reads (fnRs)

# Extract sample names, assuming filenames have format
sample.names <- sapply(strsplit(basename(fnFs), "__|_0"), function(x) x[2])

# basename(fnFs): Removes the directory path and keeps only the file names. Example: "/path/to/SAMPLE1_R1_001.fastq" → "SAMPLE1_R1_001.fastq"
# strsplit(..., "_"): Splits each filename into parts using the underscore (_) as the delimiter. Example: "SAMPLE1_R1_001.fastq" → c("SAMPLE1", "R1", "001.fastq")
# sapply(..., [, 1): Extracts the first element from each split result (the part before the first underscore). Example: c("SAMPLE1", "R1", "001.fastq") → "SAMPLE1"

cat("Processing",length(sample.names),"samples:", sample.names)
# outputs a summary of the number of samples detected and their names, legth(): counts the number, cat():prints the message

#Inspect read quality profiles
plotQualityProfile(fnFs[1:2])
# generates a quality profile plot for the first two forward read files in the fnFs list, DADA2 package in R for processing high-throughput sequence data, to visualize the quality score of sequencing reads across all base positions in a FASTQ file
# selected the first two 1:2 forward read files in fnFs 

# gray-scale: heat map of the frequency of each quality score at each base position
# green line: mean quality score at each position
# orange line: quartiles of the quality score distribution
# red line: scaled proportion of reads that extend to at least that position

# forwards reads
# --> Truncate the forward reads at position 280

plotQualityProfile(fnRs[1:2])

# reversed reads
# worse quality, especially at the end which is common in Illumina sequencing
# DADA2 incorporates quality information into its error model which makes the algorithm robust to lower quality sequence

# reversed reads
# --> Truncate the forward reads at position 200


```

### Filtering and Trimming

This pipeline ensures that only high-quality reads are retained for downstream analysis, which is crucial for the accuracy of biological insights derived from sequencing data.

```{r}
# filter and trimm 
cat("Filtering and trimming")

# prints a message to indicate that the filtering and triming steps are about to begin in the sequencing data processing pipeline

# Place filtered files in filtered/ subdirectory
filtFs <- file.path(path, "filtered", paste0(sample.names, "_F_filt.fastq.gz"))
filtRs <- file.path(path, "filtered", paste0(sample.names, "_R_filt.fastq.gz"))
names(filtFs) <- sample.names
names(filtRs) <- sample.names

### set parameters  maxN=0 (DADA2 requires no Ns), truncQ=2, rm.phix=TRUE and maxEE=2. The maxEE parameter sets the maximum number of “expected errors” allowed in a read, which is a better filter than simply averaging quality scores.
out <- filterAndTrim(fnFs, filtFs, fnRs, filtRs, truncLen=c(280,200),maxN=0, maxEE=c(2,2), truncQ=5, rm.phix=TRUE, compress=TRUE, multithread=TRUE) 
head(out)

```

### Learn error rates & sample interference

```{r}
# learn error rates
# The error model is learned using the filtered and trimmed sequences, which helps in determining the probability of errors at each base position across the reads.Forward (errF) and reverse (errR) errors are learned separately because the forward and reverse reads often have different error profiles.

cat("Learning error rates")
errF <- learnErrors(filtFs, multithread=TRUE)
errR <- learnErrors(filtRs, multithread=TRUE)
# This function in DADA2 is used to learn the error rates for sequencing reads.
# multithread=TRUE: Enables multi-threading to speed up the error rate learning process if the system supports it.

# visualize the esstimated error rates as a sanity check
plotErrors(errF, nominalQ = T)
# everything looks resonable (black line follows the points, the model fits) and we proceed with confidence

```

### Sample Inference

```{r}
# This step is essential for transforming the raw sequencing data into biologically meaningful information by correcting errors and identifying the true sequences present in your samples
# apply the core sample inference algorithm to the filtered and trimmed sequence data.
dadaFs <- dada(filtFs, err=errF, multithread=TRUE, pool = T)
dadaRs <- dada(filtRs, err=errR, multithread=TRUE, pool = T)
# It uses the error rates learned from the reverse reads (errR) to apply the error-correction model and infer the true biological sequences.

#Inspecting the returned dada-class object:
dadaFs[[1]]
dadaRs[[1]]
```

### Merging paired reads and checking on read length

```{r}
mergers <- mergePairs(dadaFs, filtFs, dadaRs, filtRs, verbose=TRUE)
# merges forward (F) and reveres (R) sequences and filtered data in one column
# Inspect the merger data.frame from the first sample
head(mergers[[1]])

# Construct sequencing table
seqtab <- makeSequenceTable(mergers)
# creates a sequence table from the merged paired-end reads. A sequence table is a matrix where: rows represent individual samples, columns represent unique DNA sequences, entries contain the counts of each sequence in each sample
dim(seqtab)
# returns the dimensions of the sequence table (1st value number of rows (sample), 2nd value is the number of columns (unique sequences))

# Inspect distribution of sequence lengths
table(nchar(getSequences(seqtab)))
# looks good
# check visually how the districution of read length is
# Get the read length distribution
length_table <- table(nchar(getSequences(seqtab)))

# Convert to data frame for plotting
length_df <- as.data.frame(length_table)
colnames(length_df) <- c("Length", "Count")

# Make sure Length is numeric
length_df$Length <- as.numeric(as.character(length_df$Length))

ggplot(length_df, aes(x = Length, y = Count)) +
  geom_col(fill = "steelblue") +  # vertical bars
  labs(title = "Distribution of Read Lengths",
       x = "Read Length (bp)",
       y = "Number of Reads") +
  xlim(280, 400)+
  theme_minimal()

```

### Remove Chimeras

```{r}
# Remove chimeras  
# Chimeric sequences are artifacts of the PCR amplification process. These sequences do not represent real biological sequences but rather result from errors in the molecular processes used to prepare and sequence DNA.
# Chimeric sequences can inflate diversity estimates. Removing them ensures that the analysis focuses only on true biological sequences.
# This code removes chimeric sequences from the sequence table, calculates the dimensions of the updated sequence table and determines the proportion of non-chimeric sequences retained.

seqtab.nochim <- removeBimeraDenovo(seqtab, method="consensus", multithread=TRUE, verbose=TRUE)


# removeBim..: identifies and removes chimeric sequences from the sequence table
# method: chimera removal method that identifies chimeras based on agreement across samples
# multithread=T: enables multi-threading for faster computation
# verbose=T: provides detailed output about the chimera removal process including the number of chimeric sequences identified.
dim(seqtab.nochim)
# tells me there are 678 ASVs in the table

## calculate the  frequency of chimeric sequences 
sum(seqtab.nochim)/sum(seqtab)
# 0.828 = 83% of your reads were retained, only 17% of reads were discarded as chimeras


# Track reads through the pipeline 
# This block of code tracks the number of reads at each stage of the DADA2 pipeline for each sample. This tracking helps monitor data loss, identify problematic samples, and ensure the pipeline is functioning as expected.
getN <- function(x) sum(getUniques(x))

# custom function that calculates the total number of unique reads (amplicon sequences)
track <- cbind(out, sapply(dadaFs, getN), sapply(dadaRs, getN), sapply(mergers, getN), rowSums(seqtab.nochim))
# combines data from multiple steps into a single table
# If processing a single sample, remove the sapply calls: e.g. replace sapply(dadaFs, getN) with getN(dadaFs)
colnames(track) <- c("input", "filtered", "denoisedF", "denoisedR", "merged", "nonchim")
rownames(track) <- sample.names
head(track)

# Purpose
# Quality Control: This table allows you to monitor read retention at each step of the pipeline, ensuring that no unexpected data loss occurs.
# Troubleshooting: If a sample shows a significant drop in reads at a specific step, it may i indicate a problem (e.g., poor quality reads or insufficient overlap for merging).
# Summary: Provides an overview of pipeline performance, enabling comparison across samples.
```

### Assign Taxonomy

```{r}
# Assign taxonomy
# This block of code assigns taxonomic classifications to the non-chimeric sequences (ASVs) in your dataset, enabling the identification of the taxa represented by the sequences

taxa <- assignTaxonomy(seqtab.nochim,"/yourpathname/silva_nr_v138_train_set.fa.gz" , multithread=TRUE)
# uses a reference database to classify each unique sequence (ASV) in the seqtab.nochim table to a taxonomic level
# rows: correspond to unique sequences (ASV)

##optional: make species level assignments based on exact matching
taxa <- addSpecies(taxa, "/yourpathname/silva_species_assignment_v138.fa.gz")
                   
#inspect the taxonomic assignments:
taxa.print <- taxa # Removing sequence rownames for display only
rownames(taxa.print) <- NULL
head(taxa.print)
```

### Prepare Data for plotting

plots the 20 most abundant bacterial families

```{r}
# add needed info to data frame
samples.out_MY <- rownames(seqtab.nochim)
info_split_MY <- strsplit(samples.out_MY, "_")
Coral_Species <- sapply(info_split_MY, function(x) sub("[0-9]+", "", x[1]))
Genotype <- sapply(info_split_MY, function(x) gsub("[^0-9]", "", x[1]))
Orientation <- sapply(info_split_MY, function(x) sub("[0-9]+", "", x[2]))
Location <- sapply(info_split_MY, function(x) gsub("[^0-9]", "", x[2]))
samdf_MY <- data.frame(Coral_Species, Genotype, Orientation, Location, stringsAsFactors = FALSE)
rownames(samdf_MY) <- samples.out_MY

# also add the column Location
# Convert Location to numeric for mapping
samdf_MY$Location <- as.numeric(samdf_MY$Location)

# Map numeric sampling location to descriptive Location
location_map_MY <- c("top", "upper middle", "lower middle", "bottom")

samdf_MY$Location <- location_map_MY[samdf_MY$Location]

rownames(samdf_MY) <- samples.out_MY

# colony 2 has two more sampling regions but for comparison I will remove it for now
# Remove rows with NA in any column
samdf_MY <- na.omit(samdf_MY)

# construct a phyloseq object directly from the dada2 outputs
ps_MY <- phyloseq(otu_table(seqtab.nochim, taxa_are_rows=FALSE), 
               sample_data(samdf_MY), 
               tax_table(taxa))

# its easier to work with the short ASV names than with the full DNA sequence
# store DNA sequences of our ASC in the refseq slot of the phyloseq object
dna_MY <- Biostrings::DNAStringSet(taxa_names(ps_MY))
names(dna_MY) <- taxa_names(ps_MY)
ps_MY <- merge_phyloseq(ps_MY, dna_MY)
taxa_names(ps_MY) <- paste0("ASV", seq(ntaxa(ps_MY)))
ps_MY
```

### Indonesia: DADA2

Data processing with [DADA2](https://github.com/benjjneb/dada2) to infer ASVs

```{r}
# Save as FASTA file
write.fasta(sequences = as.list(raw_fasta),
            names = names(raw_fasta),
            file.out = "top_ASVs.fasta")

# set working directory
# read file
cat("Reading files") # prints a message to indicate the script is starting to read files
path <- "." # current working directory
list.files(path) # lists all files in the current directory 

# Forward and reverse fastq filenames have format: 
fnFs <- sort(list.files(path, pattern="_R1_001.fastq", full.names = TRUE))
fnRs <- sort(list.files(path, pattern="_R2_001.fastq", full.names = TRUE))
# searches for files matching specific patterns _R1_001...and sorts files, get a sorted list of file paths for forward (fnFs) and reverse reads (fnRs)

# Extract sample names, assuming filenames have format
sample.names <- sapply(strsplit(basename(fnFs), "__|_0"), function(x) x[2])

# basename(fnFs): Removes the directory path and keeps only the file names. Example: "/path/to/SAMPLE1_R1_001.fastq" → "SAMPLE1_R1_001.fastq"
# strsplit(..., "_"): Splits each filename into parts using the underscore (_) as the delimiter. Example: "SAMPLE1_R1_001.fastq" → c("SAMPLE1", "R1", "001.fastq")
# sapply(..., [, 1): Extracts the first element from each split result (the part before the first underscore). Example: c("SAMPLE1", "R1", "001.fastq") → "SAMPLE1"

cat("Processing",length(sample.names),"samples:", sample.names)
# outputs a summary of the number of samples detected and their names, legth(): counts the number, cat():prints the message

#Inspect read quality profiles
plotQualityProfile(fnFs[1:2])
# generates a quality profile plot for the first two forward read files in the fnFs list, DADA2 package in R for processing high-throughput sequence data, to visualize the quality score of sequencing reads across all base positions in a FASTQ file
# selected the first two 1:2 forward read files in fnFs 

# gray-scale: heat map of the frequency of each quality score at each base position
# green line: mean quality score at each position
# orange line: quartiles of the quality score distribution
# red line: scaled proportion of reads that extend to at least that position

# forwards reads
# --> Truncate the forward reads at position 280

plotQualityProfile(fnRs[1:2])

# reversed reads
# worse quality, especially at the end which is common in Illumina sequencing
# DADA2 incorporates quality information into its error model which makes the algorithm robust to lower quality sequence

# reversed reads
# --> Truncate the forward reads at position 200


```

### Filtering and Trimming

This pipeline ensures that only high-quality reads are retained for downstream analysis, which is crucial for the accuracy of biological insights derived from sequencing data.

```{r}
# filter and trimm 
cat("Filtering and trimming")

# prints a message to indicate that the filtering and triming steps are about to begin in the sequencing data processing pipeline

# Place filtered files in filtered/ subdirectory
filtFs <- file.path(path, "filtered", paste0(sample.names, "_F_filt.fastq.gz"))
filtRs <- file.path(path, "filtered", paste0(sample.names, "_R_filt.fastq.gz"))
names(filtFs) <- sample.names
names(filtRs) <- sample.names

### set parameters  maxN=0 (DADA2 requires no Ns), truncQ=2, rm.phix=TRUE and maxEE=2. The maxEE parameter sets the maximum number of “expected errors” allowed in a read, which is a better filter than simply averaging quality scores.
out <- filterAndTrim(fnFs, filtFs, fnRs, filtRs, truncLen=c(280,200),maxN=0, maxEE=c(2,2), truncQ=5, rm.phix=TRUE, compress=TRUE, multithread=TRUE) 

head(out)
```

### Learn error rates & sample interference

```{r}
# learn error rates
# The error model is learned using the filtered and trimmed sequences, which helps in determining the probability of errors at each base position across the reads.Forward (errF) and reverse (errR) errors are learned separately because the forward and reverse reads often have different error profiles.

cat("Learning error rates")
errF <- learnErrors(filtFs, multithread=TRUE)
errR <- learnErrors(filtRs, multithread=TRUE)
# This function in DADA2 is used to learn the error rates for sequencing reads.
# multithread=TRUE: Enables multi-threading to speed up the error rate learning process if the system supports it.

# visualize the esstimated error rates as a sanity check
plotErrors(errF, nominalQ = T)
# everything looks resonable (black line follows the points, the model fits) and we proceed with confidence
```

### Sample Inference

```{r}
# This step is essential for transforming the raw sequencing data into biologically meaningful information by correcting errors and identifying the true sequences present in your samples
# apply the core sample inference algorithm to the filtered and trimmed sequence data.
dadaFs <- dada(filtFs, err=errF, multithread=TRUE, pool = T)
dadaRs <- dada(filtRs, err=errR, multithread=TRUE, pool = T)
# It uses the error rates learned from the reverse reads (errR) to apply the error-correction model and infer the true biological sequences.

#Inspecting the returned dada-class object:
dadaFs[[1]]
dadaRs[[1]]
```

### Merging paired reads and checking on read length

```{r}
mergers <- mergePairs(dadaFs, filtFs, dadaRs, filtRs, verbose=TRUE)
# merges forward (F) and reveres (R) sequences and filtered data in one column
# Inspect the merger data.frame from the first sample
head(mergers[[1]])

# Construct sequencing table
seqtab <- makeSequenceTable(mergers)
# creates a sequence table from the merged paired-end reads. A sequence table is a matrix where: rows represent individual samples, columns represent unique DNA sequences, entries contain the counts of each sequence in each sample
dim(seqtab)
# returns the dimensions of the sequence table (1st value number of rows (sample), 2nd value is the number of columns (unique sequences))

# Inspect distribution of sequence lengths
table(nchar(getSequences(seqtab)))
# looks good
# check visually how the districution of read length is
# Get the read length distribution
length_table <- table(nchar(getSequences(seqtab)))

# Convert to data frame for plotting
length_df <- as.data.frame(length_table)
colnames(length_df) <- c("Length", "Count")

# Make sure Length is numeric
length_df$Length <- as.numeric(as.character(length_df$Length))

ggplot(length_df, aes(x = Length, y = Count)) +
  geom_col(fill = "steelblue") +  # vertical bars
  labs(title = "Distribution of Read Lengths",
       x = "Read Length (bp)",
       y = "Number of Reads") +
  xlim(280, 400)+
  theme_minimal()
```

### Remove Chimeras

```{r}
# Remove chimeras  
# Chimeric sequences are artifacts of the PCR amplification process. These sequences do not represent real biological sequences but rather result from errors in the molecular processes used to prepare and sequence DNA.
# Chimeric sequences can inflate diversity estimates. Removing them ensures that the analysis focuses only on true biological sequences.
# This code removes chimeric sequences from the sequence table, calculates the dimensions of the updated sequence table and determines the proportion of non-chimeric sequences retained.

seqtab.nochim <- removeBimeraDenovo(seqtab, method="consensus", multithread=TRUE, verbose=TRUE)
# removeBim..: identifies and removes chimeric sequences from the sequence table
# method: chimera removal method that identifies chimeras based on agreement across samples
# multithread=T: enables multi-threading for faster computation
# verbose=T: provides detailed output about the chimera removal process including the number of chimeric sequences identified.
dim(seqtab.nochim)

## calculate the  frequency of chimeric sequences 
sum(seqtab.nochim)/sum(seqtab)
# 0.654 = 65.4% of your reads were retained, only the rest% of reads were discarded as chimeras


# Track reads through the pipeline 
# This block of code tracks the number of reads at each stage of the DADA2 pipeline for each sample. This tracking helps monitor data loss, identify problematic samples, and ensure the pipeline is functioning as expected.
getN <- function(x) sum(getUniques(x))

# custom function that calculates the total number of unique reads (amplicon sequences)
track <- cbind(out, sapply(dadaFs, getN), sapply(dadaRs, getN), sapply(mergers, getN), rowSums(seqtab.nochim))
# combines data from multiple steps into a single table
# If processing a single sample, remove the sapply calls: e.g. replace sapply(dadaFs, getN) with getN(dadaFs)
colnames(track) <- c("input", "filtered", "denoisedF", "denoisedR", "merged", "nonchim")
rownames(track) <- sample.names
head(track)

# Purpose
# Quality Control: This table allows you to monitor read retention at each step of the pipeline, ensuring that no unexpected data loss occurs.
# Troubleshooting: If a sample shows a significant drop in reads at a specific step, it may i indicate a problem (e.g., poor quality reads or insufficient overlap for merging).
# Summary: Provides an overview of pipeline performance, enabling comparison across samples.
```

### Assign Taxonomy

```{r}
# Assign taxonomy
# This block of code assigns taxonomic classifications to the non-chimeric sequences (ASVs) in your dataset, enabling the identification of the taxa represented by the sequences

taxa <- assignTaxonomy(seqtab.nochim,"/yourpathname/16S_files/silva_nr_v138_train_set.fa.gz" , multithread=TRUE)
# uses a reference database to classify each unique sequence (ASV) in the seqtab.nochim table to a taxonomic level
# rows: correspond to unique sequences (ASV)

#inspect the taxonomic assignments:
taxa.print <- taxa # Removing sequence rownames for display only
rownames(taxa.print) <- NULL
head(taxa.print)
```

### Prepare Data for plotting

```{r}
# add needed info to data frame
samples.out <- rownames(seqtab.nochim)
info_split <- strsplit(samples.out, "_")
Species <- sapply(info_split, function(x) sub("[0-9]+", "", x[2]))
Genotype <- sapply(info_split, function(x) gsub("[^0-9]", "", x[2]))
SamplingLocation <- sapply(info_split, function(x) x[3])
samdf <- data.frame(Species, Genotype, SamplingLocation, stringsAsFactors = FALSE)
rownames(samdf) <- samples.out

# construct a phyloseq object directly from the dada2 outputs
ps <- phyloseq(otu_table(seqtab.nochim, taxa_are_rows=FALSE), 
               sample_data(samdf), 
               tax_table(taxa))

# its easier to work with the short ASV names than with the full DNA sequence
# store DNA sequences of our ASC in the refseq slot of the phyloseq object
dna <- Biostrings::DNAStringSet(taxa_names(ps))
names(dna) <- taxa_names(ps)
ps_ID <- merge_phyloseq(ps, dna)
taxa_names(ps_ID) <- paste0("ASV", seq(ntaxa(ps_ID)))
ps_ID
```

### Barplot: Bacterial Family

```{r}
# Malaysia
# plot it again but with all other bacteria families in grey
# Groups ASVs by Family level, Counts are summed up for each Family
ps_MY_family <- tax_glom(ps_MY, taxrank = "Family")

# check what happens if I work with the df
df_ps_MY_family <- psmelt(ps_MY_family)

# clean sample column name
df_ps_MY_family$Sample <- sub("_S.*", "", df_ps_MY_family$Sample)

# 3. Normalize abundance to relative abundance per sample
df_ps_MY_family <- df_ps_MY_family %>%
  group_by(Sample) %>%
  mutate(RelAbundance = Abundance / sum(Abundance)) %>%
  ungroup()

# total number of reads
total_reads_MY <- df_ps_MY_family %>%
  summarise(TotalReads = sum(Abundance, na.rm = TRUE))

print(total_reads_MY)
# total reads: 198.725.1	

# how many ASV do we have?
ntaxa(ps_MY)
# 4417 MY
length(unique(df_ps_MY_family$OTU))
# 274

# how many families do we have?
length(unique(tax_table(ps)[, "Family"]))
#  43

# top 20
top20_families <- df_ps_MY_family %>%
  group_by(Family) %>%
  summarise(TotalRelAbundance = sum(RelAbundance)) %>%
  arrange(desc(TotalRelAbundance)) %>%
  slice_head(n = 20) %>%
  pull(Family)

# Group all non-top 20 families as "Other"
df_ps_MY_top20_family <- df_ps_MY_family %>%
  mutate(Family = ifelse(Family %in% top20_families, as.character(Family), "Other"))

# check number of top 20 families
df_ps_MY_top20_family %>% distinct(Family) %>% count()
# 21 (including others)
unique(df_ps_MY_top20_family$Family)

# check on % of each family of the top 20
# Step 1: Sum abundance per family
family_sums <- aggregate(RelAbundance ~ Family, data = df_ps_MY_top20_family, sum)

# Step 2: Compute total abundance
total_abundance <- sum(family_sums$RelAbundance)

# Step 3: Calculate percent abundance
family_sums$Percent_Abundance <- (family_sums$RelAbundance / total_abundance) * 100

# Set factor levels so "Other" appears last in stack
df_ps_MY_top20_family <- df_ps_MY_top20_family %>%
  mutate(Family = factor(Family, levels = c("Other", top20_families)))

# PLOT
# organize Genotype
df_ps_MY_top20_family <- df_ps_MY_top20_family %>%
  mutate(
    Genotype = factor(Genotype, levels = c("1", "6", "2", "3", "5", "4")))

# Orientation
df_ps_MY_top20_family <- df_ps_MY_top20_family %>%
  mutate(
    Orientation = factor(Orientation, levels = c("W", "E")))

# write the y-axis labesl in capital letters
df_ps_MY_top20_family <- df_ps_MY_top20_family %>%
  mutate(Location = str_to_title(Location))

# Location
df_ps_MY_top20_family <- df_ps_MY_top20_family %>%
  mutate(
    Location = factor(Location, levels = c("Bottom", "Lower Middle", "Upper Middle", "Top")))

# wehn combining MY and ID make sure the colorcode aligns than use this code if not continue as below
# for plotting MY and ID together asign each fmaily to a colors for plotting later
# 9 overlapping bacterial family names in total
# 31 unique family names, so in total we need 31 unique colors

# first list bacterial family names
# 1. Define all unique bacterial families from both MY and ID (31 unique)
all_families <- c(
  # 9 overlapping families first (same order for consistency)
  "Halomonadaceae",
  "Marinobacteraceae",
  "Chlorobiaceae",
  "Alcanivoracaceae1",
  "Idiomarinaceae",
  "Alteromonadaceae",
  "Rhodobacteraceae",
  "Vibrionaceae",
  "Flavobacteriaceae",
  
  # Families unique to MY
  "Endozoicomonadaceae",
  "Moraxellaceae",
  "Spirochaetaceae",
  "Marinilabiliaceae",
  "Hungateiclostridiaceae",
  "Cyclobacteriaceae",
  "Propionibacteriaceae",
  "Desulfosarcinaceae",
  "Desulfobacteraceae",
  "Clade_I",
  "Pseudomonadaceae",
  
  # Families unique to ID
  "Pseudoalteromonadaceae",
  "Colwelliaceae",
  "Marinomonadaceae",
  "Peptostreptococcales-Tissierellales_fa",
  "Arcobacteraceae",
  "Nitrincolaceae",
  "Kordiimonadaceae",
  "Stappiaceae",
  "Cryomorphaceae",
  "Marinifilaceae",
  "Saprospiraceae"
)

# assign colors to each family
kelly_colors <- c(
  "#00538A", # Strong Blue
  "#990000", # Vivid Red
  "#FF6800", # Vivid Orange
  "#007D35", # Vivid Green
  "#FFB300", # Vivid Yellow
  "#803E75", # Strong Purple
  "#A6BDD7", # Very Light Blue
  "#CC6699", # Medium Gray
  "#F6768E", # Strong Purplish Pink
  "#53377A", # Strong Violet
  "#FF0000", # Strong Purplish Red
  "#FFEA00", # Vivid Greenish Yellow
  "#99CCFF", # Light Blue
  "#93AA00", # Yellowish Green
  "#593318", # Deep Yellowish Brown
  "#3399CC", # Vivid Reddish Orange
  "#003300", # Dark Olive Green
  "#33CC99", # Aquamarine
  "#CC99FF", # Lavender
  "#CC3333", # Soft Red
  "#66CC66", # Soft Green
  "#FFCC99", # Warm Yellow
  "#6699CC", # Muted Blue
  "#996633", # Brown
  "#FF6699", # Light Pink
  "#1B9E77", # Teal
  "#D95F02", # Orange
  "#7570B3", # Indigo
  "#E7298A", # Magenta
  "#66A61E", # Olive Green
  "#1F78B4"  # Medium Blue
)

# 3. Create a named vector mapping families to colors
family_colors <- setNames(kelly_colors, all_families)

# 4. Add 'Other' color for families not listed explicitly
family_colors <- c(family_colors, Other = "grey70")

# colors by family
df_ps_MY_top20_family <- df_ps_MY_top20_family %>%
  mutate(Family = factor(Family, levels = c("Other", all_families)))

# plot
bacteria_family_plot_MY <- ggplot(df_ps_MY_top20_family, aes(x = RelAbundance * 100, y = Location, fill = Family)) +
  geom_bar(stat = "identity", position = "stack") +
  facet_grid(Genotype ~ Orientation, scales = "free_y") +
  scale_x_continuous(labels = scales::percent_format(scale = 1)) +
  scale_fill_manual(values = family_colors) +
  theme_minimal(base_size = 13) +
  theme(
    strip.text = element_text(face = "bold", size = 13),
    legend.position = "bottom",
    legend.text = element_text(size = 13),
    legend.title = element_text(size = 13),
    axis.text = element_text(size = 13),
    axis.title = element_text(size = 13),
    panel.grid = element_blank(),
    panel.border = element_blank()
  ) +
  guides(fill = guide_legend(ncol = 3)) +
  labs(
    x = "Relative Abundance (%)",
    y = "Sampling Location",
    fill = "Bacterial Family"
  )

bacteria_family_plot_MY


# Indonesia
# check overall numbers 
# Groups ASVs by Family level, Counts are summed up for each Family
ps_ID_family <- tax_glom(ps_ID, taxrank = "Family")

# check what happens if I work with the df
df_ps_ID_family <- psmelt(ps_ID_family)

#  Normalize abundance to relative abundance per sample
df_ps_ID_family <- df_ps_ID_family %>%
  group_by(Sample) %>%
  mutate(RelAbundance = Abundance / sum(Abundance)) %>%
  ungroup()

# total number of reads
total_reads <- df_ps_ID_family %>%
  summarise(TotalReads = sum(Abundance, na.rm = TRUE))

print(total_reads)
# total reads:502.311	

# how many ASV do we have?
ntaxa(ps)
# 678

# how many families do we have?
length(unique(tax_table(ps)[, "Family"]))
# 44

# top 20
top20_families <- df_ps_ID_family %>%
  group_by(Family) %>%
  summarise(TotalRelAbundance = sum(RelAbundance)) %>%
  arrange(desc(TotalRelAbundance)) %>%
  slice_head(n = 20) %>%
  pull(Family)

# check top 20 families
top20_families

# Group all non-top 20 families as "Other"
df_ID_top20_family <- df_ps_ID_family %>%
  mutate(Family = ifelse(Family %in% top20_families, Family, "Other"))

# check number of bacteria families
df_ID_top20_family %>%
  distinct(Family) %>%
  count()
# 21 which is correct because I have 20 families and the rest is other

# check on % of each family of the top 20
# Step 1: Sum abundance per family
family_sums <- aggregate(RelAbundance ~ Family, data = df_ID_top20_family, sum)

# Step 2: Compute total abundance
total_abundance <- sum(family_sums$RelAbundance)

# Step 3: Calculate percent abundance
family_sums$Percent_Abundance <- (family_sums$RelAbundance / total_abundance) * 100

# Plot
# organize Genotype
df_ID_top20_family <- df_ID_top20_family %>%
  mutate(
    Genotype = factor(Genotype, levels = sort(unique(Genotype))))

# order of colonies
df_ID_top20_family <- df_ID_top20_family %>%
  mutate(
    Genotype = factor(Genotype, levels = c("01", "02", "03", "04", "05", "06", "07")))

# Location
df_ID_top20_family <- df_ID_top20_family %>%
  mutate(
    SamplingLocation = factor(SamplingLocation, levels = c("bottom", "top")))

df_ID_top20_family <- df_ID_top20_family %>%
  mutate(
    SamplingLocation = factor(SamplingLocation,
                              levels = c("bottom", "top"),
                              labels = c("Bottom", "Top"))
  )

# Set factor levels so "Other" appears last in stack
df_ID_top20_family <- df_ID_top20_family %>%
  mutate(Family = factor(Family, levels = c("Other", all_families)))

#plot
bacteria_family_plot_ID <- ggplot(df_ID_top20_family, aes(x = RelAbundance * 100, y = SamplingLocation, fill = Family)) +
  geom_bar(stat = "identity", position = "stack") +
  facet_wrap(~Genotype, scales = "free_y", ncol = 1) +
  scale_x_continuous(labels = scales::percent_format(scale = 1)) +
  scale_fill_manual(values = family_colors) +
  theme_minimal(base_size = 13) +
  theme(
    strip.text = element_blank(),  # Remove genotype labels
    legend.position = "bottom",    # Move legend underneath the plot
    legend.text = element_text(size = 13),
    legend.title = element_text(size = 13),
    axis.text = element_text(size = 13),
    axis.title = element_text(size = 13),
    panel.grid = element_blank(),
    legend.box = "horizontal"      # Layout legend horizontally
  ) +
  guides(fill = guide_legend(ncol = 3)) +  # Adjust number of rows as needed
  labs(
    x = "Relative Abundance[%]",
    y = "Sampling Location",
    fill = "Bacterial Family"
  )

bacteria_family_plot_ID
```

## **Fig. 4 Bacterial community composition across the surface of massive Porites sp. colonies from Indonesia and Malaysia**

```{r}
# combine ID and MYS
# Combine plots side-by-side with equal width
final_16S_combined_plot <-  bacteria_family_plot_ID + bacteria_family_plot_MY +
  plot_layout(ncol = 2, widths = c(1, 1)) +
  plot_annotation(
    theme = theme(
      plot.margin = margin(10, 10, 10, 10)
    )
  )

final_16S_combined_plot

# save plot and edit in affinity designer
# Save as PDF
ggsave(
  filename = "16S_MY_ID.pdf",
  plot = final_16S_combined_plot,
  width = 11.7,      # A4 width in inches
  height = 10,     # A4 height in inches
  units = "in"
)
```

# Supplementary Material

## Supplementary Figure 1 and 2: TS2 sequences - absolute abundance

```{r}
# plot absolute abundance of ITS2 sequences
# Indonesia
# upload data
its2_sequences_ID <- read.table('/yourpathname/post_med_seqs/582_20250509T135802_DBV_20250510T135051.seqs.absolute.abund_and_meta.txt', header = TRUE, sep = "\t", stringsAsFactors = FALSE)

# remove columns with meta data only keep sample name and ITS2 sequences
its2_sequences_ID_filtered <- its2_sequences_ID[, -c(1, 3:39)]

# remove rows with NA
its2_sequences_ID_filtered <- its2_sequences_ID_filtered %>%
  filter(sample_name != "" & !is.na(sample_name))

# Convert to long format for ggplot
its2_sequences_ID_filtered_long <- its2_sequences_ID_filtered %>%
  pivot_longer(
    cols = -sample_name,          # all columns except SampleID
    names_to = "Sequence",
    values_to = "AbsoluteAbundance"
  )

# make sure the order of sequences is correct (highest abundnace at the bottom of the barplot)
seq_ID_order <- its2_sequences_ID_filtered_long %>%
  group_by(Sequence) %>%
  summarize(total_abundance = sum(AbsoluteAbundance, na.rm = TRUE)) %>%
  arrange(desc(total_abundance)) %>%      # highest abundance first
  pull(Sequence)

# Set factor levels so that highest abundance is at the bottom
its2_sequences_ID_filtered_long$Sequence <- factor(
  its2_sequences_ID_filtered_long$Sequence,
  levels = seq_ID_order
)

# organize data
# Example sample vector
its2_sequences_ID_filtered_long$sample_name <- factor(its2_sequences_ID_filtered_long$sample_name, levels = 
            c("Por_01_Top", "Por_01_Bottom","Por_02_Top", "Por_02_Bottom", 
             "Por_03_Top",  "Por_03_Bottom" , "Por_04_Top" , "Por_04_Bottom", "Por_06_Top","Por_06_Bottom",
             "Por_07_Bottom", "Por_07_Top"))



# set colors
palette_marlen <- c(
             "#E704C4", "#C2FFED", "#A079BF", "#CC0744", "#C0B9B2", "#C2FF99", "#001E09", "#00489C", "#6F0062",
            "#0CBD66",
            "#EEC3FF", "#456D75", "#B77B68", "#7A87A1", "#788D66", "#885578", "#FAD09F", "#FF8A9A", "#D157A0",
            "#BEC459",
            "#456648", "#0086ED", "#886F4C", "#34362D", "#B4A8BD", "#00A6AA", "#452C2C", "#636375", "#A3C8C9",
            "#FF913F",
            "#938A81", "#575329", "#00FECF", "#B05B6F", "#8CD0FF", "#3B9700", "#04F757", "#C8A1A1", "#1E6E00",
            "#7900D7",
            "#A77500", "#6367A9", "#A05837", "#6B002C", "#772600", "#D790FF", "#9B9700", "#549E79", "#FFF69F",
            "#201625",
            "#72418F", "#BC23FF", "#99ADC0", "#3A2465", "#922329", "#5B4534", "#FDE8DC", "#404E55", "#0089A3",
            "#CB7E98",
            "#A4E804", "#324E72", "#6A3A4C", "#83AB58", "#001C1E", "#D1F7CE", "#004B28", "#C8D0F6", "#A3A489",
            "#806C66",
            "#222800", "#BF5650", "#E83000", "#66796D", "#DA007C", "#FF1A59", "#8ADBB4", "#1E0200", "#5B4E51",
            "#C895C5",
            "#320033", "#FF6832", "#66E1D3", "#CFCDAC", "#D0AC94", "#7ED379", "#012C58", "#7A7BFF", "#D68E01",
            "#353339",
            "#78AFA1", "#FEB2C6", "#75797C", "#837393", "#943A4D", "#B5F4FF", "#D2DCD5", "#9556BD", "#6A714A",
            "#001325",
            "#02525F", "#0AA3F7", "#E98176", "#DBD5DD", "#5EBCD1", "#3D4F44", "#7E6405", "#02684E", "#962B75",
            "#8D8546",
            "#9695C5", "#E773CE", "#D86A78", "#3E89BE", "#CA834E", "#518A87", "#5B113C", "#55813B","#FFB500",
            "#00005F",
            "#A97399", "#4B8160", "#59738A", "#FF5DA7", "#F7C9BF", "#643127", "#513A01", "#6B94AA", "#51A058",
            "#A45B02",
            "#1D1702", "#E20027", "#E7AB63", "#4C6001", "#9C6966", "#64547B", "#97979E", "#006A66", "#391406",
            "#F4D749",
            "#0045D2", "#006C31", "#DDB6D0", "#7C6571", "#9FB2A4", "#00D891", "#15A08A", "#BC65E9", "#FFFFFE",
            "#C6DC99",
            "#203B3C", "#671190", "#6B3A64", "#F5E1FF", "#FFA0F2", "#CCAA35", "#374527", "#8BB400", "#797868",
            "#C6005A",
            "#3B000A", "#C86240", "#29607C", "#402334", "#7D5A44", "#CCB87C", "#B88183", "#AA5199", "#B5D6C3",
            "#A38469",
            "#9F94F0", "#A74571", "#B894A6", "#71BB8C", "#00B433", "#789EC9", "#6D80BA", "#953F00", "#5EFF03",
            "#E4FFFC",
            "#1BE177", "#BCB1E5", "#76912F", "#003109", "#0060CD", "#D20096", "#895563", "#29201D", "#5B3213",
            "#A76F42",
            "#89412E", "#1A3A2A", "#494B5A", "#A88C85", "#F4ABAA", "#A3F3AB", "#00C6C8", "#EA8B66", "#958A9F",
            "#BDC9D2",
            "#9FA064", "#BE4700", "#658188", "#83A485", "#453C23", "#47675D", "#3A3F00", "#061203", "#DFFB71",
            "#868E7E",
            "#98D058", "#6C8F7D", "#D7BFC2", "#3C3E6E", "#D83D66", "#2F5D9B", "#6C5E46", "#D25B88", "#5B656C",
            "#00B57F",
            "#545C46", "#866097", "#365D25", "#252F99", "#00CCFF", "#674E60", "#FC009C", "#92896B")

# plot
ID_ITS2seq <- ggplot(its2_sequences_ID_filtered_long, aes(x = sample_name, y = AbsoluteAbundance, fill = Sequence)) +
  geom_bar(stat = "identity", position = position_stack(reverse = TRUE)) +
  scale_fill_manual(values = palette_marlen ) +
  labs(x = "Sample", y = "Absolute Abundance") +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "none"   # hide legend
  )
ID_ITS2seq 


# Malaysia
# upload data
its2_sequences_MY <- read.table('/yourpatthname/post_med_seqs/560_20250219T125105_DBV_20250221T224139.seqs.absolute.abund_and_meta.txt', header = TRUE, sep = "\t", stringsAsFactors = FALSE)

# remove columns with meta data only keep sample name and ITS2 sequences
its2_sequences_MY_filtered <- its2_sequences_MY[, -c(1, 3:39)]

# remove rows with NA
its2_sequences_MY_filtered <- its2_sequences_MY_filtered %>%
  filter(sample_name != "" & !is.na(sample_name))

# Convert to long format for ggplot
its2_sequences_MY_filtered_long <- its2_sequences_MY_filtered %>%
  pivot_longer(
    cols = -sample_name,          # all columns except SampleID
    names_to = "Sequence",
    values_to = "AbsoluteAbundance"
  )


# make sure the order of sequences is correct (highest abundnace at the bottom of the barplot)
seq_MY_order <- its2_sequences_MY_filtered_long %>%
  group_by(Sequence) %>%
  summarize(total_abundance = sum(AbsoluteAbundance, na.rm = TRUE)) %>%
  arrange(desc(total_abundance)) %>%      # highest abundance first
  pull(Sequence)

# Set factor levels so that highest abundance is at the bottom
its2_sequences_MY_filtered_long$Sequence <- factor(
  its2_sequences_MY_filtered_long$Sequence,
  levels = seq_MY_order
)


# plot
MY_ITS2_seq <- ggplot(its2_sequences_MY_filtered_long, aes(x = sample_name, y = AbsoluteAbundance, fill = Sequence)) +
  geom_bar(stat = "identity", position = position_stack(reverse = TRUE)) +
  scale_fill_manual(values = palette_marlen ) +
  labs(x = "Sample", y = "Absolute Abundance") +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    legend.position = "none"   # hide legend
  )

MY_ITS2_seq
```

## Supplementary Figure 3: Visual Bleaching assessment

```{r}
# Indonesia
VBA_data_ID <- read.csv('/yourpathname/VBA_ID_0_420.csv')

# only select needed temperatures
VBA_data_ID <- VBA_data_ID[VBA_data_ID$Temperature %in% 40, ]

# adjust dataframe so it looks like the MYS
VBA_data_ID <- VBA_data_ID %>%
  mutate(Timepoint = ifelse(Timepoint == 420, 1, Timepoint)) %>%
  rename(VBA = VBA_mean)

# calculate relative pigemntation
# so first the baseline is the % of pigmentation at timepoint 0 for each genotype, orientation and location
baseline <- VBA_data_ID %>%
  filter(Timepoint == 0) %>%
  group_by(Genotype, Location) %>%
  summarise(Baseline_Pigmentation = first(VBA), .groups = "drop")

# than we merge the datasets so we have the baseline value in it
VBA_ID_relative <- VBA_data_ID %>%
  left_join(baseline, by = c("Genotype", "Location"))

# than we calculate the relative pigmentation loss: so we take the baseline pigmentation value - the VBA at time point 1 which tells us how much pigmentation we lost between timepoint 0 and timepoint 1 so its the loss in pigemntation relative to the starting point each fragment had
VBA_ID_relative <- VBA_ID_relative %>%
   group_by(Genotype, Location) %>%
  mutate(Relative_Pigmentation_Loss = (Baseline_Pigmentation-VBA))

VBA_ID_relative <- VBA_ID_relative %>%
   group_by(Genotype,Location) %>%
  mutate(Relative_Pigmentation = 100- Relative_Pigmentation_Loss)

# only select timepoint 1
VBA_ID_relative <- VBA_ID_relative %>%
  filter(Timepoint == 1)

# calculate the mean Relative Pigmentation and differences for each colony
Relative_Pigmentation_means <- VBA_ID_relative %>%
  group_by(Genotype) %>%
  summarise(mean_Relative_Pigmentation = mean(Relative_Pigmentation))

# Merge back to main dataset (on Genotype *and* Orientation)
Final_ID_VBA <- VBA_ID_relative %>%
  left_join(Relative_Pigmentation_means, by = c("Genotype")) %>%  # Match correctly
  mutate(Relative_Pigmentation_diff = Relative_Pigmentation - mean_Relative_Pigmentation) %>%  # Compute difference
  select(-contains(".x")) %>%  # Remove duplicate columns (like Orientation.x)
  distinct()

# plot
# set order of the Locations
 Final_ID_VBA$Location <- factor(Final_ID_VBA$Location, 
                                   levels = c("bottom",  "top"))
 
 # remove colony 5
Final_ID_VBA <- Final_ID_VBA[Final_ID_VBA$Genotype != 5, ]
 
# set order of the colonies
Final_ID_VBA <- Final_ID_VBA %>%
  mutate(
    Genotype = factor(Genotype, levels = c("1", "2", "3", "4", "6", "7")))

 # set the colors               
genotype_colors <- c("1" = "pink4", 
                     "2" = "orchid4", 
                     "3" = "pink2", 
                     "4" = "mistyrose2", 
                     "6" = "lightcoral",
                     "7" = "hotpink4")

    
# Plot VBA differences (Bar plot)
ID_VBA_diff_barplot <- ggplot(Final_ID_VBA, aes(x = Relative_Pigmentation_diff, y = Location, fill = Genotype)) +
  
  # Add mean ED50 line for each genotype
  geom_vline(aes(xintercept = 0), color = "black", linetype = "dashed", size = 1) +
  
  # Add bars showing ED50 relative to the mean
  geom_bar(stat = "identity", position = "dodge") +
  
  # Facet by Genotype (7 rows)
  facet_wrap(~Genotype, nrow = 7) +
  
  # Labels and themes
  labs(
    x = "Deviation from Mean Relative Pigmentation [%]", 
    y = "Location within Colony",
    fill = "Genotype"
  ) +
  
  # Custom x-axis: -50 to 50 with ticks every 10 units
  scale_x_continuous(
    limits = c(-50, 50),
    breaks = seq(-50, 50, by = 10)
  ) +
  
  # Custom y-axis labels
  scale_y_discrete(labels = c("Bottom", "Top")) +  
  
  # Custom fill colors
  scale_fill_manual(values = genotype_colors) + 
  
  # Theme
  theme_minimal() +
  theme(
    strip.text = element_blank(),  
    axis.title = element_text(size = 14, face = "bold"),  
    axis.text = element_text(size = 12),  
    legend.position = "none",  
    panel.spacing = unit(1, "lines"),
    strip.background = element_blank(),
    strip.text.x = element_text(size = 14, face = "bold")
  )

ID_VBA_diff_barplot

# Violin plot with the mean of the entire colony (per genotype)
ID_VBA_violin_plot <- ggplot(Final_ID_VBA, aes(x = 1, y = Relative_Pigmentation, fill = Genotype)) +
  
  # Create violin plots showing the distribution of ED50 for each Genotype
  geom_violin() +
  
  # Facet by Genotype (stacked vertically)
  facet_wrap(~Genotype, nrow = 7) +  # 6 rows, 1 column
  
  # Add mean ED50 within each plot (as a black point)
  stat_summary(fun = "mean", geom = "point", color = "black", size = 3, shape = 18, show.legend = FALSE) + 
  
  # Add mean ED50 value as a text label on the plot
  stat_summary(fun = "mean", geom = "text", color = "black", size = 4, 
               aes(label = round(..y.., 2)), vjust = -1) +  # Show mean value as text (rounded to 2 decimal places)
  
  # Labels and themes
  labs(
    x = NULL,  # Remove x-axis label
    y = "Relative Pigmentaiton [%]",
    fill = "Genotype"
  ) +
  
  scale_fill_manual(values = genotype_colors) +  # Apply the custom genotype colors
  #guides(fill = guide_legend(title = "Coral Colony")) +  # Change the legend title
  theme_minimal() +
  theme(
    strip.text = element_text(size = 14, face = "bold"),  # Bold genotype labels in facets
    axis.title = element_text(size = 14, face = "bold"),  
    axis.text = element_text(size = 12),  
    legend.position = "none",  # Position the legend to the right
    panel.spacing = unit(1, "lines"),  # Spacing between plots
    strip.background = element_blank(),  # Remove the facet background
    axis.text.x = element_blank(),  # Remove x-axis text labels
    axis.ticks.x = element_blank()  # Remove x-axis ticks
  )


# Show the plot
ID_VBA_violin_plot

# Combine the two plots side by side
ID_VBA_combined_plot <-  VBA_violin_plot + VBA_diff_barplot + plot_layout(ncol = 2, widths = c(0.8,1.2))+
  plot_annotation(
    title = "Pigmentation Differences and Violin Plot of Coral Colonies from Indonesia",  # Title for the combined plot
    theme = theme(
      plot.title = element_text(size = 16, face = "bold", hjust = 0.5)  # Customize the title appearance
    )
  )

VBA_combined_plot


# Malaysia
# load data
VBA_data_MYS<- read.csv("/yourpathname/VBA_MY_0_420.csv")

# only select needed temperatures
VBA_data_MYS <- VBA_data_MYS[VBA_data_MYS$Temperature %in% 42, ]

# create column Location
VBA_data_MYS<- VBA_data_MYS %>%
  mutate(Location = case_when(
    Condition %in% c("E1", "W1") ~ "top",
    Condition %in% c("E2", "W2") ~ "upper_middle",
    Condition %in% c("E3", "W3") ~ "lower_middle",
    Condition %in% c("E4", "W4") ~ "bottom",
    TRUE ~ NA_character_  # This will assign NA to any conditions not specified above
  ))
# add column orientation
VBA_data_MYS <- VBA_data_MYS %>%
  mutate(Orientation = case_when(
    Condition %in% c("E1", "E2", "E3", "E4") ~ "East",
     Condition %in% c("W1", "W2", "W3", "W4") ~ "West",
    TRUE ~ NA_character_  # This will assign NA to any conditions not specified above
  ))

# calculate relative pigemntation
# so first the baseline is the % of pigmentation at timepoint 0 for each genotype, orientation and location
baseline <- VBA_data_MYS %>%
  filter(Timepoint == 0) %>%
  group_by(Genotype, Orientation, Condition) %>%
  summarise(Baseline_Pigmentation = first(VBA), .groups = "drop")

# than we merge the datasets so we have the baseline value in it
VBA_MYS_relative <- VBA_data_MYS %>%
  left_join(baseline, by = c("Genotype", "Orientation", "Condition"))

# than we calculate the relative pigmentation loss: so we take the baseline pigmentation value - the VBA at time point 1 which tells us how much pigmentation we lost between timepoint 0 and timepoint 1 so its the loss in pigemntation relative to the starting point each fragment had
VBA_MYS_relative <- VBA_MYS_relative %>%
   group_by(Genotype, Orientation, Condition) %>%
  mutate(Relative_Pigmentation_Loss = (Baseline_Pigmentation-VBA))

VBA_MYS_relative <- VBA_MYS_relative %>%
   group_by(Genotype, Orientation, Condition) %>%
  mutate(Relative_Pigmentation = 100- Relative_Pigmentation_Loss)

# only select timepoint 1
VBA_MYS_relative <- VBA_MYS_relative %>%
  filter(Timepoint == 1)

# calculate the mean Relative Pigmentation and differences for each colony
Relative_Pigmentation_means <- VBA_MYS_relative %>%
  group_by(Genotype) %>%
  summarise(mean_Relative_Pigmentation = mean(Relative_Pigmentation))

# Merge back to main dataset (on Genotype *and* Orientation)
Final_VBA <- VBA_MYS_relative %>%
  left_join(Relative_Pigmentation_means, by = c("Genotype")) %>%  # Match correctly
  mutate(Relative_Pigmentation_diff = Relative_Pigmentation - mean_Relative_Pigmentation) %>%  # Compute difference
  select(-contains(".x")) %>%  # Remove duplicate columns (like Orientation.x)
  distinct()

# plot
# set order of the Locations
 Final_VBA$Location <- factor(Final_VBA$Location, 
                                   levels = c("bottom", "lower_middle", "upper_middle", "top"))
 
# set order of the colonies
Final_VBA <- Final_VBA %>%
  mutate(
    Genotype = factor(Genotype, levels = c("1", "6", "2", "3", "5", "4")))



# adjust Orientation column
Final_VBA <- Final_VBA %>%
  mutate(
    Orientation = factor(Orientation, levels = c("West", "East")))

           

# set the colors    
 genotype_colors <- c("1" = "skyblue3", 
                     "2" = "lightblue", 
                     "3" = "dodgerblue4", 
                     "4" = "cornflowerblue", 
                     "5" = "lightskyblue", 
                     "6" = "dodgerblue3")
 

# Plot ED50 differences (Bar plot)
MY_VBA_diff_barplot <- ggplot(Final_VBA, aes(x = Relative_Pigmentation_diff, y = Location, fill = Genotype)) +
  
  # Add mean ED50 line for each genotype
  geom_vline(aes(xintercept = 0), color = "black", linetype = "dashed", size = 1) +
  
  # Add bars showing ED50 relative to the mean
  geom_bar(stat = "identity", position = "dodge") +
  
  # Facet by Orientation (West on left, East on right)
  facet_grid(Genotype ~ Orientation, scales = "free_y") +
  
  # Labels and themes
  labs(
    x = "Deviation from Mean Relative Pigmentation [%]", 
    y = "Location within Colony",
    fill = "Genotype"
  ) +
  
  # Custom x-axis: -50 to 50 in 10-step intervals
  scale_x_continuous(
    limits = c(-55, 55),
    breaks = seq(-50, 50, by = 20)
  ) +
  
  # Custom y-axis labels
  scale_y_discrete(labels = c("Bottom", "Lower Middle", "Upper Middle", "Top")) +  
  
  # Custom fill colors
  scale_fill_manual(values = genotype_colors) + 
  
  theme_minimal() +
  theme(
    strip.text = element_blank(),  
    axis.title = element_text(size = 14, face = "bold"),  
    axis.text = element_text(size = 12),  
    legend.position = "none",  
    panel.spacing = unit(1, "lines"),
    strip.background = element_blank(),  
    strip.text.x = element_text(size = 14, face = "bold")
  )

MY_VBA_diff_barplot

# Violin plot with the mean of the entire colony (per genotype)
MY_VBA_violin_plot <- ggplot(Final_VBA, aes(x = 1, y = Relative_Pigmentation, fill = Genotype)) +
  
  # Create violin plots showing the distribution of ED50 for each Genotype
  geom_violin() +
  
  # Facet by Genotype (stacked vertically)
  facet_wrap(~Genotype, nrow = 6) +  # 6 rows, 1 column
  
  # Add mean ED50 within each plot (as a black point)
  stat_summary(fun = "mean", geom = "point", color = "black", size = 3, shape = 18, show.legend = FALSE) + 
  
  # Add mean ED50 value as a text label on the plot
  stat_summary(fun = "mean", geom = "text", color = "black", size = 4, 
               aes(label = round(..y.., 2)), vjust = -1) +  # Show mean value as text (rounded to 2 decimal places)
  
  # Labels and themes
  labs(
    x = NULL,  # Remove x-axis label
    y = "Relative Pigmentation [%]",
    fill = "Genotype"
  ) +
  
  scale_fill_manual(values = genotype_colors) +  # Apply the custom genotype colors
  guides(fill = guide_legend(title = "Coral Colony")) +  # Change the legend title
  theme_minimal() +
  theme(
    strip.text = element_text(size = 14, face = "bold"),  # Bold genotype labels in facets
    axis.title = element_text(size = 14, face = "bold"),  
    axis.text = element_text(size = 12),  
    legend.position = "none",  # Position the legend to the right
    panel.spacing = unit(1, "lines"),  # Spacing between plots
    strip.background = element_blank(),  # Remove the facet background
    axis.text.x = element_blank(),  # Remove x-axis text labels
    axis.ticks.x = element_blank()  # Remove x-axis ticks
  )

MY_VBA_violin_plot


# Combine the two plots side by side
MY_VBA_combined_plot <-  MY_VBA_violin_plot + MY_VBA_diff_barplot + plot_layout(ncol = 2, widths = c(0.8,1.2))+
  plot_annotation(
    title = "Pigmentation Differences and Violin Plot of Coral Colonies from Malaysia",  # Title for the combined plot
    theme = theme(
      plot.title = element_text(size = 16, face = "bold", hjust = 0.5)  # Customize the title appearance
    )
  )

MY_VBA_combined_plot


# combining the barplots of Indonesia and Malaysia for the Supplementary Table
final_combined_plot <- ID_VBA_diff_barplot + MY_VBA_diff_barplot +
  plot_layout(ncol = 2, widths = c(1, 1)) +  # Equal widths for both plots
  plot_annotation(
    theme = theme(
      plot.margin = margin(10, 10, 10, 10)
    )
  )
final_combined_plot

# The plot was edited in Affinity designer
```

### Supplementary Tables: Statistical analysis Bacterial and Symbiodinaiceae community composition

```{r}
# Supplementary Tbale 14:
# Indonesia
# check that assumptions are given variance across groups must be homogeneous

# Example 1: Genotype
disp_genotype <- betadisper(dist_matrix_ID, metadata_ID$Genotype)
anova(disp_genotype)        # test for significant differences
permutest(disp_genotype)    # permutation test (more robust)
boxplot(disp_genotype)      # visualize group dispersions

# Example 2: Location
disp_location <- betadisper(dist_matrix_ID, metadata_ID$Location)
anova(disp_location)
permutest(disp_location)
boxplot(disp_location)

# Example 3: Depth
disp_depth <- betadisper(dist_matrix_ID, metadata_ID$Depth)
anova(disp_depth)
permutest(disp_depth)
boxplot(disp_depth)


# 2. Separate metadata and abundance
# Adjust the metadata columns if you have more than 3 (Genotype, Orientation, Location)
metadata_ID <- VBA_Bacteria_ID_wide[, 1:7]  # includes SampleID

# Everything after the first 4 columns are bacterial family abundances
bacteria_matrix_ID <- VBA_Bacteria_ID_wide[, 8:ncol(VBA_Bacteria_ID_wide)]

# 3. Convert to data frame for vegan functions
bacteria_matrix_ID <- as.data.frame(bacteria_matrix_ID)

# 4. Assign row names for alignment
rownames(bacteria_matrix_ID) <- metadata_ID$SampleID

# 5. Create distance matrix
dist_matrix_ID <- vegdist(bacteria_matrix_ID, method = "bray")

# 6. PERMANOVA all
permanova_result <- adonis2(
   dist_matrix_ID ~ Genotype + Location + Depth,
   data = metadata_ID,
   permutations = 999,
   method = "bray",
   by = "margin"   
 )

print(permanova_result)
# nothing significant



# Malaysia 
# combine data set: VBA and bacteria
# first make sure the grouping columns are the same
Final_VBA_MY <- Final_VBA_MY %>%
  mutate(
    Orientation = case_when(
      Orientation == "East" ~ "E",
      Orientation == "West" ~ "W",
      TRUE ~ Orientation  # keep others unchanged
    ),
    Location = case_when(
      Location == "top" ~ "Top",
      Location == "upper_middle" ~ "Upper Middle",
      Location == "lower_middle" ~ "Lower Middle",
      Location == "bottom" ~ "Bottom",
      TRUE ~ Location  # keep others unchanged
    )
  )

# only select needed columns
VBA_MY_clean <- Final_VBA_MY %>%
  select(Country, Site, Genotype, Orientation, Location, VBA, Relative_Pigmentation)

# merge data
VBA_Bacteria_MY <- df_ps_MYS_top20_family %>%
  left_join(VBA_MY_clean, by = c("Genotype", "Orientation", "Location"))

# 1. Wide format for PERMANOVA
VBA_Bacteria_MY_wide <- VBA_Bacteria_MY %>%
  pivot_wider(
    id_cols = c(Genotype, Orientation, Location, Relative_Pigmentation),
    names_from = Family,
    values_from = RelAbundance,
    values_fill = 0,
    values_fn = sum
  ) %>%
  mutate(SampleID = paste(Genotype, Orientation, Location, sep = "_"))

# adjust dataframe with Ed50 and depth data
Combined_data_MYS <- Combined_data_MYS %>%
  mutate(
    Orientation = case_when(
      Orientation == "East" ~ "E",
      Orientation == "West" ~ "W",
      TRUE ~ Orientation  # keep others unchanged
    ),
    Location = case_when(
      Location == "top" ~ "Top",
      Location == "upper_middle" ~ "Upper Middle",
      Location == "lower_middle" ~ "Lower Middle",
      Location == "bottom" ~ "Bottom",
      TRUE ~ Location  # keep others unchanged
    )
  )

# add ed50 and depth
VBA_Bacteria_MY_wide <- VBA_Bacteria_MY_wide %>%
  left_join(Combined_data_MYS %>% select(Genotype, Orientation, Location, Depth, ED50),
    by = c("Genotype", "Location", "Orientation"))
  
# move new column to the front
VBA_Bacteria_MY_wide <- VBA_Bacteria_MY_wide %>%
  mutate(SampleID = paste(Genotype, Orientation, Location, sep = "_")) %>%
  relocate(SampleID, Depth, ED50, .before = 1)

# 2. Separate metadata and abundance
# Adjust the metadata columns if you have more than 3 (Genotype, Orientation, Location)
metadata_MY <- VBA_Bacteria_MY_wide[, 1:7]  # includes SampleID

# Everything after the first 4 columns are bacterial family abundances
bacteria_matrix_MY <- VBA_Bacteria_MY_wide[, 8:ncol(VBA_Bacteria_MY_wide)]

# 3. Convert to data frame for vegan functions
bacteria_matrix_MY <- as.data.frame(bacteria_matrix_MY)

# 4. Assign row names for alignment
rownames(bacteria_matrix_MY) <- metadata_MY$SampleID

# 5. Create distance matrix
dist_matrix_MY <- vegdist(bacteria_matrix_MY, method = "bray")

# 6. PERMANOVA checks what impacts bacterial composition: Genotype

# Supplementary Table 15
# run permanova with all predictors in one 
permanova_result <- adonis2(
  dist_matrix_MY ~ Genotype + Orientation + Location + Depth,
  data = metadata_MY,
  permutations = 999,
  method = "bray",
  by = "margin"   
)

print(permanova_result)
# only genotype is significant


# Indonesia
# are its2 type profiles significantly different across colonies?
# dataframe: its2_profile_ID wide format

# Copy dataframe
ITS2_ID <- its2_profile_ID

# Extract Genotype and Location from sample name
# Assuming format: Por_01_Top
ITS2_ID <- its2_profile_ID %>%
  separate(sample_name,
           into = c("Species", "Genotype", "Location"),
           sep = "_") %>%
  relocate(Species, Genotype, Location)

# Metadata: species, genotype, location
metadata_ID <- ITS2_ID[, 1:3]

# Only keep numeric ITS2 columns
ITS2_matrix_ID <- ITS2_ID[, 5:ncol(ITS2_ID)] %>% 
  mutate(across(everything(), as.numeric))

# Supplementary table 16
# compute bray-curtis dissimilarity matrix
dist_ITS2_ID <- vegdist(ITS2_matrix_ID, method = "bray")

# check distance
dist_ITS2_ID

# permanova
ITS2_permanova2 <- adonis2(
  dist_ITS2_ID ~ Genotype,
  data = metadata_ID,
  permutations = 999,
  method = "bray",
  by = "margin"
)

print(ITS2_permanova2)

# SUpplementary table 17
# Malaysia
# ITS2: its2_profile_MY_long_filtered
# Bacteria, ED50, VBA.: VBA_Bacteria_MY_wide
str(its2_profile_MY_long_filtered)

# checking coklumn names
colnames(its2_profile_MY_long_filtered)


its2_MY <- its2_profile_MY_long_filtered %>%
  dplyr::rename(
    Orientation = collection_orientation,
    Location = collection_location,
    Genotype = collection_colony
  ) 

its2_MY <- its2_MY %>%
  mutate(
    Orientation = case_when(
      Orientation == "East" ~ "E",
      Orientation == "West" ~ "W",
      TRUE ~ Orientation  # keeps all other values unchanged
    )
  )

# only select needed columns
its2_MY <- its2_MY %>%
  select(Genotype, collection_depthrange, Orientation, ITS2_profile, value)

# make the data frame wide just with ITS2 type profile data
ITS2_MY_wide <- its2_MY %>%
  pivot_wider(
    id_cols = c(Genotype, Orientation, collection_depthrange),
    names_from = ITS2_profile,
    values_from = value,
    values_fill = 0,
    values_fn = sum
  ) %>%
  mutate(SampleID = paste(Genotype, Orientation, collection_depthrange, sep = "_"))

# move new column to the front
ITS2_MY_wide <- ITS2_MY_wide %>%
  mutate(SampleID = paste(Genotype, Orientation, collection_depthrange, sep = "_")) %>%
  relocate(SampleID, .before = 1)

# put dataframes together
combined_MY <- ITS2_MY_wide %>%
  left_join(
    VBA_Bacteria_MY_wide %>%
      select(SampleID, Genotype, Orientation, collection_depthrange,
             ED50, Depth, Relative_Pigmentation,
             everything()),  # keeps all other bacterial columns too
    by = c("SampleID", "Genotype", "Orientation", "Location")
  )

str(combined_MY)

# move metadata columns to the front
combined_MY <- combined_MY %>%
  relocate(ED50, Depth, Relative_Pigmentation, .after = Location)


# PERMANOVA just ITS2 
ITS2_MY_wide <- combined_MY[, 1:13]


# Extract ITS2 profiles (columns 5 to end)
ITS2_matrix <- ITS2_MY_wide[, 8:ncol(ITS2_MY_wide)]

# Metadata
metadata <- ITS2_MY_wide[, 1:7]

# or just do it with the ITS2 data
# Select only ITS2 profile columns (abundance data)
ITS2_matrix <- ITS2_MY_wide[, 5:ncol(ITS2_MY_wide)]

# Metadata
metadata <- ITS2_MY_wide[, 1:4]

# Convert to a matrix
ITS2_matrix <- as.matrix(ITS2_matrix)

# Compute Bray–Curtis dissimilarity
dist_ITS2 <- vegdist(ITS2_matrix, method = "bray")

# Run PERMANOVA
# run permanova
ITS2_permanova <- adonis2(
  dist_ITS2 ~ Genotype + Orientation + collection_depthrange,
  data = metadata,
  permutations = 999,
  method = "bray",
  by = "margin"   
)

print(ITS2_permanova)
# only genotype significant

# checking whether its2 type profile differ across depth locations (shallow: 1,6,2 and deep: 3,5,4)
# Add DepthGroup based on Genotype
ITS2_MY_wide <- ITS2_MY_wide %>%
  mutate(
    DepthGroup = case_when(
      Genotype %in% c("1", "2", "6") ~ "shallow",
      Genotype %in% c("3", "4", "5") ~ "deep",
      TRUE ~ NA_character_
    )
  )

# move metadata columns to the front
ITS2_MY_wide <- ITS2_MY_wide %>%
  relocate(DepthGroup, .after = collection_depthrange)

# Select only ITS2 profile columns
ITS2_matrix <- ITS2_MY_wide[, 6:ncol(ITS2_MY_wide)]

# Metadata
metadata <- ITS2_MY_wide[, 1:5]

# Convert to a matrix
ITS2_matrix <- as.matrix(ITS2_matrix)

# Compute Bray–Curtis dissimilarity
dist_ITS2 <- vegdist(ITS2_matrix, method = "bray")

# Run PERMANOVA
# run permanova
ITS2_permanova <- adonis2(
  dist_ITS2 ~ DepthGroup,
  data = metadata,
  permutations = 999,
  method = "bray",
  by = "margin"   
)

print(ITS2_permanova)
# significant difference in ITS2 type profiles across depth groups



```
